### 2022-08-10 paper/paper-infer-supp.tex
init paper

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}
\usepackage{placeins}
\usepackage{subcaption}

%\renewcommand{\arraystretch}{1.8}

%\linenumbers
%\doublespacing
\onehalfspacing
\usepackage[numbers,super]{natbib}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
\renewcommand*{\arraystretch}{1.2}

\usepackage{babel}
\makeatother

\begin{document}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\setcounter{equation}{0}

\section*{Supplementary Materials}

\vspace*{7em}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{figure}[!h]
	%\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
	\caption{ \label{fig:hist-info}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\FloatBarrier

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\FloatBarrier
\clearpage

\bibliographystyle{natbib}
\bibliography{refs}

\end{document}

### 2022-08-10 paper/paper-infer.tex
init paper

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{xurl}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}
\usepackage{makecell}
%\usepackage[font=small]{caption}

\renewcommand{\arraystretch}{1.3}

\usepackage{xr}
\externaldocument{paper-infer-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
\usepackage[authoryear]{natbib}
%\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\usepackage{algorithm}
\usepackage{algpseudocode}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{}
\author{Florian Priv\'e,$^{\text{1,}*}$  and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\

\noindent$^{\text{\sf 5}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contact: \url{florian.prive.21@gmail.com}

%\vspace*{6em}
\clearpage

\begin{abstract}

\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Materials and Methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

\begin{figure}[!h]
	%\centerline{\includegraphics[width=0.95\textwidth]{res-FIN}}
	\caption{}
	\label{fig:finngen}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\clearpage
%\vspace*{3em}

\section*{Acknowledgements}

Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.

\section*{Funding}

F.P.\ and B.J.V.\ are supported by a Lundbeck Foundation Fellowship (R335-2019-2339 to B.J.V.).

\section*{Declaration of Interests}

B.J.V.\ is on Allelica's international advisory board.
The other authors have no competing interests to declare.

\section*{Code and data availability}

The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code}.
We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN [TODO: VERIFY].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage
%\vspace*{3em}

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2022-08-10 paper/paper-infer.tex
describe new LDpred2-auto

@@
 
-\section{Materials and Methods}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Results}
@@
 \section{Discussion}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Materials and Methods}
+
+\subsection{New model and inference with LDpred2-auto}
+
+LDpred2 originally assumed the following model for effect sizes,
+\begin{equation}\label{eq:prev_model}
+\beta_j = S_j \gamma_j \sim \left\{
+\begin{array}{ll}
+\mathcal N\left(0,~\dfrac{h^2}{M p}\right) & \mbox{with probability $p$,} \\
+0 & \mbox{otherwise,}
+\end{array}
+\right.
+\end{equation}
+where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\gamma$ the effect sizes on the allele scale, $S$ the standard deviations of the genotypes, and $\beta$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
+In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters.
+This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
+Previously, $p$ was sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
+
+Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
+First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
+\begin{equation}\label{eq:new_model}
+\beta_j = S_j \gamma_j \sim \left\{
+\begin{array}{ll}
+\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
+0 & \mbox{otherwise.}
+\end{array}
+\right.
+\end{equation}
+Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
+This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite{zeng2021widespread}. 
+In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
+In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before finding the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
+This maximum likelihood estimation is implemented using R package roptim, and we bound the estimate of $\alpha$ to be within -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the previous estimate.
+We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling in order to account for the reduced effective number of correlated variants.
+As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants.
+We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
+
+\subsection{Extending the set of HapMap3 variants}
 

### 2022-08-10 paper/paper-infer.tex
start intro

@@
 
-\title{}
-\author{Florian Priv\'e,$^{\text{1,}*}$  and Bjarni J. Vilhj\'almsson$^{\text{1,5}}$}
+\title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
+\author{Florian Priv\'e,$^{\text{1,}*}$  and Bjarni J. Vilhj\'almsson$^{\text{1,2}}$}
 
@@
 \noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-
-\noindent$^{\text{\sf 5}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
+\noindent$^{\text{\sf 2}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 
+Most traits and diseases are heritable, if not all. 
+What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
+Some phenotypes, such as height and schizophrenia are highly heritable [TODO: ADD CITATIONS]. 
+These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them [TODO: ADD CITATIONS].
+Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
+Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
+These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.
+
+These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone [TODO: ADD CITATIONS]. 
+Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases [TODO: ADD CITATIONS].
+LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model.
+In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two hyper-parameters of the model \cite[]{prive2020ldpred2}.
+
+Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
+We show that we can also reliably estimate the predictive ability of PGS derived from LDpred2-auto.
+
+[TO FINISH]
+
+[ALSO TALK ABOUT NEW SET OF VARIANTS]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \section{Discussion}
+
+[KEEP RG FOR DISCU + CITE EXISTING METHODS]
+
+[BETTER SET OF VARIANTS, BUT STILL NOT PERFECT E.G. FOR PROTEIN LEVELS]
 
@@
 In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before finding the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
-This maximum likelihood estimation is implemented using R package roptim, and we bound the estimate of $\alpha$ to be within -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the previous estimate.
+This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be within -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the previous estimate.
 We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling in order to account for the reduced effective number of correlated variants.

### 2022-08-11 paper/paper-infer-supp.tex
save

@@
 \vspace*{7em}
+\clearpage
 
@@
 
-\begin{figure}[!h]
-	%\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
-	\caption{ \label{fig:hist-info}}
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{est_h2_one}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{est_p_one}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{est_alpha_one}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{est_r2_one}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{postp_calib}}
+\caption{ \label{}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+[ ADD THE SAME, BUT AVERAGING OVER 10 SIMUS -> CI OF THE MEAN ]
+
+\FloatBarrier
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{est_binary_h2_one}}
+	\caption{All h2 estimates were transformed to the liability scale with K\_pop=K and K\_GWAS either K or 0.5 (when using Neff) \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{est_binary_p_one}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{est_binary_alpha_one}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{est_binary_r2_one}}
+	\caption{Transformed both r2 to the liability scale, where the inferred one using Neff used K\_GWAS=0.5  \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{binary_postp_calib}}
+	\caption{ \label{}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_p}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_alpha}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_r2}}
+	\caption{Height has estimated h2 of 0.555 and 0.626 -> probably too large, and why estimated r2 is too large as well.
+The other outliers is 695.4 (Lupus), with a very small number of cases -> should probably stop using 16 PCs \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_sets}}
+	\caption{ \label{}}
 \end{figure}

### 2022-08-18 paper/paper-infer.tex
save
describe the extended set

@@
 \title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
-\author{Florian Priv\'e,$^{\text{1,}*}$  and Bjarni J. Vilhj\'almsson$^{\text{1,2}}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6}}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Bioinformatics Research Centre, Aarhus University, Aarhus, 8000, Denmark. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
+\noindent$^{\text{\sf 1}}$National Centre for Register-based Research, Aarhus University, Aarhus, Denmark. \\
+\noindent$^{\text{\sf 2}}$Bioinformatics Interdepartmental Program, University of California, Los Angeles, CA. \\
+\noindent$^{\text{\sf 3}}$Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
+\noindent$^{\text{\sf 4}}$Department of Pathology and Laboratory Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
+\noindent$^{\text{\sf 5}}$Department of Computational Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
+\noindent$^{\text{\sf 6}}$Bioinformatics Research Centre, Aarhus University, Aarhus, Denmark. \\
+\noindent$^\ast$To whom correspondence should be addressed. \\
 
@@
 
+[ALSO TALK ABOUT IMPUTATION]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Extending the set of HapMap3 variants}
+\subsection{Extending the set of HapMap3 variants used}
+
+The HapMap3 variants generally provide a good coverage of the whole genome.
+We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
+Here we propose to extend the set we have used for LDpred2 until then. 
+This extension aims at making sure many genetic variants are well tagged by the extended set, which we call HapMap3+.
+To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the one in the MFI files). 
+We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (Field 22006) to have a diverse set of populations.
+Finally, we design an algorithm which aims at maximizing the tagging of previously read variants.
+We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set.
+We start by including all previously used HapMap3 variants.
+For the sake of simplicity, we then use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
+Note that we allow only non-ambiguous SNPs to be included.
+This results in an extended set of 1,444,196 SNPs.
 
@@
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN [TODO: VERIFY].
+The latest version of R package bigsnpr can be installed from GitHub.%, and a recent enough version can be installed from CRAN [TODO: VERIFY].
 

### 2022-08-18 paper/paper-infer-supp.tex
start writing simulation results

@@
 
-\section*{Supplementary Materials}
-
-\vspace*{7em}
-\clearpage
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \centerline{\includegraphics[width=\textwidth]{est_h2_one}}
-\caption{ \label{}}
+\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
+Horizontal lines represent the true simulated values.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+\label{fig:simu_h2}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{est_p_one}}
-\caption{ \label{}}
+\caption{Inferred polygenicity $p$ in simulations with continuous outcomes. \label{fig:simu_p}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{est_alpha_one}}
-\caption{ \label{}}
+\caption{Inferred $\alpha$ in simulations with continuous outcomes. \label{fig:simu_alpha}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{postp_calib}}
+\caption{Calibration of per-variant posterior probabilities of being causal in simulations with continuous outcomes. \label{fig:simu_postp_calib}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{local_h2_calib}}
+\caption{Calibration of per-block heritability estimates in simulations with continuous outcomes. \label{fig:simu_h2_calib}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{est_r2_one}}
-\caption{ \label{}}
-\end{figure}
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{postp_calib}}
-\caption{ \label{}}
+\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes. \label{fig:simu_r2}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
-	\caption{ \label{}}
-\end{figure}
-
-\begin{figure}[p]
 	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_p}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_sets}}
+\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_altpop}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+	\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_chains}}
+\caption{ \label{}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
 \caption{ \label{}}

### 2022-08-18 paper/paper-infer.tex
start writing simulation results

@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{Results}
+\section*{Results}
+
+\subsection*{Validating the inference with simulations}
+
+For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}; we use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
+We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e.
+the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
+This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix.
+Finally, some Gaussian noise is added so that the full phenotype has a variance of 1.
+Then, a GWAS is run to obtain summary statistics using $N$ (either the 200,000 dedicated to this, or a subset of 20,000) individuals using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
+Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
+LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$.
+
+LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2},\ref{fig:simu_p} and \ref{fig:simu_alpha}).
+Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates that are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+Finally, LDpred2-auto can be used to infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Figure \ref{fig:simu_h2}).
+[TODO: TALK ABOUT WHERE IT FAILED AND LIMITATIONS]
+[TODO: RUN AND REPORT THE MEAN]
+
+We also run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence [TODO: MAYBE REWORD].
+Results are very similar as with the continuous phenotypes above, and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
+The only difference is the transformation to the liability scale that needs to be applied to the $h^2$ and $r^2$ estimates from LDpred2-auto, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ \cite[]{grotzinger2021pervasive}.
+ [TO FINISH]
+
+\subsection*{Genetic architectures of 250 phenotypes from the UK Biobank}
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-\section{Materials and Methods}
+\section*{Materials and Methods}
 
-\subsection{New model and inference with LDpred2-auto}
+\subsection*{Data for simulations}
+
+For simulations, we use the UK Biobank imputed (BGEN) data, read as allele dosages with function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{bycroft2018uk,prive2017efficient}. 
+We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
+Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (Field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
+This results in 356,409 individuals of Northwestern European ancestry.
+We randomly sample 200,000 individuals to form a training set (to run the GWAS), and use the remaining individuals to form a test set (to evaluate the predictive models).
+
+\subsection*{New model and inference with LDpred2-auto}
 
@@
 
-\subsection{Extending the set of HapMap3 variants used}
+\subsection*{Extending the set of HapMap3 variants used}
 
@@
 This results in an extended set of 1,444,196 SNPs.
+
+[TODO: EXPLAIN HOW TO INFER R2]
 

### 2022-08-19 paper/paper-infer.tex
save

@@
 
+\subsection*{Data for the UK Biobank analyses}
+
+We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 individuals of Northwestern European ancestry as in the simulations.
+As covariates, we first recompute PCs for this subset of individuals using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
+We also use sex (Field 22001), age (Field 21022), birth date (combining Fields 34 and 52) and deprivation index (Field 189) as additional covariates.
+We randomly sample 50,000 individuals to form a test set (to evaluate the predictive models), and use the remaining individuals to form a training set (to run the GWAS).
+
+We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
+We design two other LD matrices: one using a subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
+We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
+We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').
+
+\subsection*{Extending the set of HapMap3 variants used}
+
+The HapMap3 variants generally provide a good coverage of the whole genome.
+We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
+Here we propose to extend the set we have used for LDpred2 until then. 
+This extension aims at making sure many genetic variants are well tagged by the extended set, which we call HapMap3+.
+To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the one in the MFI files). 
+We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (Field 22006) to have a diverse set of populations.
+Finally, we design an algorithm which aims at maximizing the tagging of previously read variants.
+We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set.
+We start by including all previously used HapMap3 variants.
+For the sake of simplicity, we then use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
+Note that we only allow non-ambiguous SNPs to be included.
+This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+
 \subsection*{New model and inference with LDpred2-auto}
@@
 We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
-
-\subsection*{Extending the set of HapMap3 variants used}
-
-The HapMap3 variants generally provide a good coverage of the whole genome.
-We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
-Here we propose to extend the set we have used for LDpred2 until then. 
-This extension aims at making sure many genetic variants are well tagged by the extended set, which we call HapMap3+.
-To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the one in the MFI files). 
-We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (Field 22006) to have a diverse set of populations.
-Finally, we design an algorithm which aims at maximizing the tagging of previously read variants.
-We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set.
-We start by including all previously used HapMap3 variants.
-For the sake of simplicity, we then use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
-Note that we allow only non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs.
 

### 2022-08-31 paper/paper-infer-supp.tex
save
write some UKBB results

@@
 
-[ ADD THE SAME, BUT AVERAGING OVER 10 SIMUS -> CI OF THE MEAN ]
+%[ ADD THE SAME, BUT AVERAGING OVER 10 SIMUS -> CI OF THE MEAN ]
 
@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_h2}}
+	\caption{SNP heritability estimates from either LDpred2-auto or LD Score regression for all 248 phenotypes defined from the UK Biobank.
+These are stratified by estimated polygenicity (from LDpred2-auto).
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+\label{fig:ukbb_h2}}
+\end{figure}
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_p}}
+	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or the polygenicity $p$ for all 248 phenotypes defined from the UK Biobank.
+The 95\% confidence intervals for the LDpred2-auto estimates are obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+\label{fig:ukbb_h2_p}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_alpha}}
+	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or $\alpha$ for all 248 phenotypes defined from the UK Biobank.
+The 95\% confidence intervals for the LDpred2-auto estimates are obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+We only show phenotypes for which there are more than 25 chains kept, because simulations have shown that $\alpha$ estimates are unreliable when a small number of chains is kept (Figure \ref{fig:simu_alpha}).
+\label{fig:ukbb_h2_alpha}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_r2}}
+	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
+These are stratified by estimated heritability (from LDpred2-auto).
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:ukbb_r2}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
+\caption{TODO: CHANGE LABELS \label{fig:small_LD}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_altpop}}
+\caption{TODO: CHANGE LABELS \label{fig:alt_LD}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+	\caption{TODO: CHANGE LABELS \label{fig:hm3_plus}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_chains}}
+\caption{ \label{fig:ukbb_chains}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
+\caption{TODO: START AT 0 \label{fig:ukbb_runtimes}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_local_h2}}
 	\caption{ \label{}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_h2_alpha}}
+	\centerline{\includegraphics[width=\textwidth]{protein_local_h2}}
+	\caption{ \label{}}
+\end{figure}
+
+\FloatBarrier
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{mean-local-h2}}
 	\caption{ \label{}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_r2}}
-	\caption{Height has estimated h2 of 0.555 and 0.626 -> probably too large, and why estimated r2 is too large as well.
-The other outliers is 695.4 (Lupus), with a very small number of cases -> should probably stop using 16 PCs \label{}}
-\end{figure}
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
+	\centerline{\includegraphics[width=\textwidth]{mean-local-h2-perc}}
 	\caption{ \label{}}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_altpop}}
+	\centerline{\includegraphics[width=\textwidth]{local-h2-p}}
 	\caption{ \label{}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+\centerline{\includegraphics[width=\textwidth]{ukbb_compare-transfo}}
 \caption{ \label{}}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_chains}}
+	\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
 	\caption{ \label{}}
 \end{figure}
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
-\caption{ \label{}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2022-08-31 paper/paper-infer.tex
write some UKBB results

@@
 
-\subsection*{Genetic architectures of 250 phenotypes from the UK Biobank}
+\subsection*{Genetic architectures of 248 phenotypes from the UK Biobank}
 
-\begin{figure}[!h]
-	%\centerline{\includegraphics[width=0.95\textwidth]{res-FIN}}
-	\caption{}
-	\label{fig:finngen}
-\end{figure}
+We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations.
+50,000 of these will be used as test set, while the other 306,409 are used to run a GWAS using linear regression (function \texttt{big\_univLinReg}) for each of all 248 phenotypes and using 8 covariates (Methods).
+Then the new implementation of LDpred2-auto (Methods) is used.
+
+Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
+Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
+Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
+Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
+As for the inferred predictive performance, they are highly consistent with the ones derived from the test; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
+Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and these are higher than values reported in the literature.
+
+We then investigate previous estimates using two alternative LD references (Methods).
+Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar (Figure \ref{fig:small_LD}).
+When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance (Figure \ref{fig:alt_LD}).
+
+We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods).
+As expected, compared to HapMap3, higher $h^2$ and lower $p$ estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
+A much larger $h^2$ estimate is obtained for lipoprotein(a) concentration [TODO: ADD NUMBERS], which is also reflected in a larger predictive performance.
+[TODO: COMPUTE AVERAGE R2 BOOST FOR OTHER PHENO]
+Interestingly, when using this extended set of variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
+However, fitting with this extended set of variants is more computationally demanding [TODO RATIO]; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).
 
@@
 
+[ALSO TALK ABOUT LDSC?]
+
+[FUTURE WORK + LIMITATION: OVERESTIMATED H2 WITH SMALL LD + NEED TO WORK ON CORRECTION]
+
+[ASSORTATIVE MATING FOR HEIGHT?]
+
+[VERY POLYGENIC, BUT NOT INFINITESIMAL -> THEREFORE LDPRED2-INF NEVER RECOMMENDED (LDPRED2 + MISSPEC)]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (Field 22020).
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
 These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
-To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (Field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
 This results in 356,409 individuals of Northwestern European ancestry.
@@
 We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 individuals of Northwestern European ancestry as in the simulations.
-As covariates, we first recompute PCs for this subset of individuals using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
-We also use sex (Field 22001), age (Field 21022), birth date (combining Fields 34 and 52) and deprivation index (Field 189) as additional covariates.
 We randomly sample 50,000 individuals to form a test set (to evaluate the predictive models), and use the remaining individuals to form a training set (to run the GWAS).
 
+We construct and use the same phenotypes as in \cite{prive2021high}.
+About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
+As covariates, for the subset of individuals previously defined, we first recompute PCs using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
+We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates.
+
 We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
-We design two other LD matrices: one using a subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
+We design two other LD matrices: one using a smaller subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
 We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
@@
 The HapMap3 variants generally provide a good coverage of the whole genome.
-We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
+We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
 Here we propose to extend the set we have used for LDpred2 until then. 
-This extension aims at making sure many genetic variants are well tagged by the extended set, which we call HapMap3+.
-To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the one in the MFI files). 
-We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (Field 22006) to have a diverse set of populations.
-Finally, we design an algorithm which aims at maximizing the tagging of previously read variants.
-We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set.
+This extension aims at making sure many genetic variants are well tagged by the extended set.
+To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
+We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (field 22006) to have a set of individuals from diverse ancestries.
+Finally, we design an algorithm which aims at maximizing the tagging of all these variants read.
+We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
 We start by including all previously used HapMap3 variants.
-For the sake of simplicity, we then use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
+Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
 Note that we only allow non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
 
@@
 
+\subsection*{Inference of predictive performance from the Gibbs sampler}
+
 [TODO: EXPLAIN HOW TO INFER R2]
@@
 
+Authors thank members of the NCRR/QGG StatGen group for helpful discussions.
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.

### 2022-09-02 paper/paper-infer.tex
save

@@
 
+[OVERVIEW TABLE OF WHAT LDPRED2 CAN PROVIDE?]
+
 \section*{Results}
@@
 We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods).
-As expected, compared to HapMap3, higher $h^2$ and lower $p$ estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
-A much larger $h^2$ estimate is obtained for lipoprotein(a) concentration [TODO: ADD NUMBERS], which is also reflected in a larger predictive performance.
-[TODO: COMPUTE AVERAGE R2 BOOST FOR OTHER PHENO]
+As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.6-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
+This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
+A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
 Interestingly, when using this extended set of variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
 However, fitting with this extended set of variants is more computationally demanding [TODO RATIO]; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).
+
+Finally, we investigate different transformations to apply to some continuous phenotypes used here.
+Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
+However, we do investigate alternative transformations here to decide which one should be preferred and to check the impact on the inference from LDpred2-auto.
+Note that we use the HapMap3+ set of variants here.
+We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates for $p$ and $\alpha$ are highly consistent.
+However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
+We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
+Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.
+
+\subsection*{Local heritability and poligenicity}
+
+In this section, we use the new HapMap3+ set of variants constructed here. 
+We recall that we have defined 431 independent LD blocks for this set (Methods).
+We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks.
 
@@
 [VERY POLYGENIC, BUT NOT INFINITESIMAL -> THEREFORE LDPRED2-INF NEVER RECOMMENDED (LDPRED2 + MISSPEC)]
+
+[HM3\_PLUS AND MISSING HERITABILITY]
 
@@
 
+[FURTHER RESTRICT VARIANTS THAT CAN ENTER THIS NEW SET?]
+
 \subsection*{New model and inference with LDpred2-auto}
@@
 In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
-In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before finding the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
-This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be within -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the previous estimate.
-We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling in order to account for the reduced effective number of correlated variants.
-As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants.
+In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters.
+This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration.
+We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of correlated variants.
+As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
 We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
+Note that this $h^2$ estimate can be restricted to e.g.\ variants from a single LD block to get estimates of local heritability.
 
@@
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
-This research has been conducted using the UK Biobank Resource under Application Number 58024.
+This research has been conducted using the UK Biobank Resource under Application Number 58024; authors thank all the participants of the UK Biobank for contributing to such useful data for Research.
 
@@
 
+[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-04 paper/paper-infer-supp.tex
save

@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
-\caption{TODO: CHANGE LABELS \label{fig:small_LD}}
+\caption{ \label{fig:small_LD}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_altpop}}
-\caption{TODO: CHANGE LABELS \label{fig:alt_LD}}
+\caption{ \label{fig:alt_LD}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
-	\caption{TODO: CHANGE LABELS \label{fig:hm3_plus}}
+	\caption{ \label{fig:hm3_plus}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_chains}}
-\caption{ \label{fig:ukbb_chains}}
+\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
-\caption{TODO: START AT 0 \label{fig:ukbb_runtimes}}
+\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_notransfo}}
+	\caption{ \label{fig:notransfo}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_transfo}}
+	\caption{ \label{fig:RINT}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	\centerline{\includegraphics[width=\textwidth]{ukbb_local_h2}}
-	\caption{ \label{}}
+	\caption{ \label{fig:ukbb_local_h2}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{protein_local_h2}}
-	\caption{ \label{}}
-\end{figure}
-
-\FloatBarrier
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{mean-local-h2}}
-	\caption{ \label{}}
+	\caption{ \label{fig:protein_local_h2}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{mean-local-h2-perc}}
-	\caption{ \label{}}
+	\centerline{\includegraphics[width=\textwidth]{median_local_h2}}
+	\caption{Per-block median heritability across 169 UKBB phenotypes.
+The HapMap3+ set of variants is used, with 431 independent LD blocks.
+Only phenotypes with more than 25 chains kept are used here.
+The top block is on chromosome 6 [22.1-41.4 Mb], which contains the HLA region.
+\label{fig:median_local_h2}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{local-h2-p}}
-	\caption{ \label{}}
-\end{figure}
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_compare-transfo}}
-\caption{ \label{}}
+\centerline{\includegraphics[width=\textwidth]{median_postp}}
+\caption{Per-variant median probabilities of being causal across 169 UKBB phenotypes.
+The HapMap3+ set of variants is used.
+Only phenotypes with more than 25 chains kept are used here.
+\label{median_postp}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
-	\caption{ \label{}}
+	\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations. $N$ is the GWAS sample size. \label{fig:enrichment}}
 \end{figure}

### 2022-09-04 paper/paper-infer.tex
save

@@
 We then investigate previous estimates using two alternative LD references (Methods).
-Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar (Figure \ref{fig:small_LD}).
-When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance (Figure \ref{fig:alt_LD}).
+Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
+When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).
 
 We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods).
-As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.6-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
+As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
 This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
 A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
-Interestingly, when using this extended set of variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, fitting with this extended set of variants is more computationally demanding [TODO RATIO]; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).
+Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
+However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).
 
@@
 We recall that we have defined 431 independent LD blocks for this set (Methods).
-We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks.
+We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
+For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
+We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
+
+Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks in Figure \ref{fig:median_local_h2}.
+The median heritability explained by a block of variants is largely proportional to the number of variants in this block.
+Unsurprisingly, the outlier block explaining a much larger heritability contains the HLA region (Figure \ref{fig:median_local_h2}).
+Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
+Some variants in multiple small regions across the genome seem to have a much larger probability of being causal; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
+
+[TODO: HEIGHT ENRICHMENT]
 
@@
 This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
-
-[FURTHER RESTRICT VARIANTS THAT CAN ENTER THIS NEW SET?]
 

### 2022-09-05 paper/paper-infer-supp.tex
save

@@
 
+\newcommand{\NEFF}{Option \texttt{Neff} controls whether a logistic regression is used for the GWAS with the effective sample in LDpred2, or a linear regression and then the total sample size.}
+
 \begin{document}
@@
 \caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
-Horizontal lines represent the true simulated values.
+Horizontal dashed lines represent the true simulated values.
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
 The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
 \label{fig:simu_h2}}
@@
 \centerline{\includegraphics[width=\textwidth]{est_p_one}}
-\caption{Inferred polygenicity $p$ in simulations with continuous outcomes. \label{fig:simu_p}}
+\caption{Inferred polygenicity $p$ in simulations with continuous outcomes.
+Horizontal dashed lines represent the true simulated values.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+\label{fig:simu_p}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{est_alpha_one}}
-\caption{Inferred $\alpha$ in simulations with continuous outcomes. \label{fig:simu_alpha}}
+\caption{Inferred $\alpha$ in simulations with continuous outcomes.
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the estimates.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+\label{fig:simu_alpha}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{postp_calib}}
-\caption{Calibration of per-variant posterior probabilities of being causal in simulations with continuous outcomes. \label{fig:simu_postp_calib}}
+\caption{Calibration of per-variant posterior probabilities of being causal in simulations with continuous outcomes.
+Green dashed lines represent the 1:1 line.
+Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented.
+\label{fig:simu_postp_calib}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{local_h2_calib}}
-\caption{Calibration of per-block heritability estimates in simulations with continuous outcomes. \label{fig:simu_h2_calib}}
+\caption{Calibration of per-block heritability estimates in simulations with continuous outcomes.
+Green dashed lines represent the 1:1 line.
+Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented.
+\label{fig:simu_h2_calib}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{est_r2_one}}
-\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes. \label{fig:simu_r2}}
+\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes.
+Green segments represent $r^2$ in the test set.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+\label{fig:simu_r2}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{est_binary_h2_one}}
-	\caption{All h2 estimates were transformed to the liability scale with K\_pop=K and K\_GWAS either K or 0.5 (when using Neff) \label{}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with binary outcomes.
+Horizontal dashed lines represent the true simulated values.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+{\NEFF} All $h^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}.
+\label{fig:simu_h2_bin}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{est_binary_p_one}}
-	\caption{ \label{}}
+	\caption{Inferred polygenicity $p$ in simulations with binary outcomes. Horizontal dashed lines represent the true simulated values. {\NEFF} \label{fig:simu_p_bin}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{est_binary_alpha_one}}
-	\caption{ \label{}}
+	\caption{Inferred $\alpha$ in simulations with continuous outcomes.
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the estimates.
+{\NEFF} \label{fig:simu_alpha_bin}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{est_binary_r2_one}}
-	\caption{Transformed both r2 to the liability scale, where the inferred one using Neff used K\_GWAS=0.5  \label{}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{binary_postp_calib}}
-	\caption{ \label{}}
+	\caption{Inferred predictive performance $r^2$ in simulations with binary outcomes.
+Green segments represent $r^2$ in the test set.
+{\NEFF} All $r^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}; the ones from the test set are transformed with K and K.
+\label{fig:simu_r2_bin}}
 \end{figure}
@@
 	\caption{SNP heritability estimates from either LDpred2-auto or LD Score regression for all 248 phenotypes defined from the UK Biobank.
-These are stratified by estimated polygenicity (from LDpred2-auto).
+These are stratified by the polygenicity estimated from LDpred2-auto.
+Green dashed lines represent the 1:1 line.
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
@@
 	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
-These are stratified by estimated heritability (from LDpred2-auto).
+These are stratified by the polygenicity estimated from LDpred2-auto.
+Green dashed lines represent the 1:1 line.
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept.
@@
 \begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
+\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations. $N$ is the GWAS sample size. \label{fig:enrichment}}
+\end{figure}
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=\textwidth]{median_local_h2}}
@@
 
-\begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
-	\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations. $N$ is the GWAS sample size. \label{fig:enrichment}}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-05 paper/paper-infer.tex
save

@@
 
-LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2},\ref{fig:simu_p} and \ref{fig:simu_alpha}).
-Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates that are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
-Finally, LDpred2-auto can be used to infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Figure \ref{fig:simu_h2}).
-[TODO: TALK ABOUT WHERE IT FAILED AND LIMITATIONS]
-[TODO: RUN AND REPORT THE MEAN]
+LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
+Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
+When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
+The $\alpha$ estimate becomes unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
+Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+Finally, LDpred2-auto can be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).
 
-We also run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence [TODO: MAYBE REWORD].
-Results are very similar as with the continuous phenotypes above, and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
+We also run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
+Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
 The only difference is the transformation to the liability scale that needs to be applied to the $h^2$ and $r^2$ estimates from LDpred2-auto, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ \cite[]{grotzinger2021pervasive}.
- [TO FINISH]
 
@@
 We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
+Back to the UKBB data, instead of computing per-block heritability estimates, we compute per-annotation ones, to investigate functional enrichment.
+We perform this analysis for height using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
+Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K (Figure \ref{fig:enrichment}).
 
@@
 Some variants in multiple small regions across the genome seem to have a much larger probability of being causal; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
-
-[TODO: HEIGHT ENRICHMENT]
 
@@
 [HM3\_PLUS AND MISSING HERITABILITY]
+
+[MENTION YI'S PAPER]
 

### 2022-09-05 paper/paper-infer-supp.tex
save

@@
 	\centerline{\includegraphics[width=\textwidth]{est_binary_alpha_one}}
-	\caption{Inferred $\alpha$ in simulations with continuous outcomes.
+	\caption{Inferred $\alpha$ in simulations with binary outcomes.
 Horizontal dashed lines represent the true simulated values.
@@
 	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or the polygenicity $p$ for all 248 phenotypes defined from the UK Biobank.
-The 95\% confidence intervals for the LDpred2-auto estimates are obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
 \label{fig:ukbb_h2_p}}
@@
 	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or $\alpha$ for all 248 phenotypes defined from the UK Biobank.
-The 95\% confidence intervals for the LDpred2-auto estimates are obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
 We only show phenotypes for which there are more than 25 chains kept, because simulations have shown that $\alpha$ estimates are unreliable when a small number of chains is kept (Figure \ref{fig:simu_alpha}).
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_r2}}
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_r2}}
 	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
-\caption{ \label{fig:small_LD}}
+\caption{LDpred2-auto estimates for UKBB phenotypes with either a small or a large LD reference.
+Only 154 phenotypes with more than 25 chains kept when using the large LD reference are represented here.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:small_LD}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_altpop}}
-\caption{ \label{fig:alt_LD}}
+\caption{LDpred2-auto estimates for UKBB phenotypes with either a N.W.
+European or a S.
+European (``altpop'') LD reference.
+Only 154 phenotypes with more than 25 chains kept when using the N.W.
+European LD reference are represented here.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:alt_LD}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
-	\caption{ \label{fig:hm3_plus}}
+	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
+Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:hm3_plus}}
 \end{figure}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_chains}}
+\centerline{\includegraphics[width=0.9\textwidth]{ukbb_chains}}
 \caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{ukbb_runtimes}}
+\centerline{\includegraphics[width=0.9\textwidth]{ukbb_runtimes}}
 \caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_notransfo}}
-	\caption{ \label{fig:notransfo}}
+	\centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_notransfo}}
+	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their raw versions.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:notransfo}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_transfo}}
-	\caption{ \label{fig:RINT}}
+	\centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_transfo}}
+	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their rank-based inverse normal (RIN) transformed versions.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:RINT}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{ukbb_local_h2}}
-	\caption{ \label{fig:ukbb_local_h2}}
+	\caption{Genome-wide and per-block heritability estimates from LDpred2-auto for UKBB phenotypes.
+The HapMap3+ variants are used here.
+The maximum local $h^2$ is the maximum heritability estimate across all 431 independent LD blocks defined for this set of variants.
+Only phenotypes for which this represents at least 10\% of the total heritability are represented.
+\label{fig:ukbb_local_h2}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=\textwidth]{protein_local_h2}}
-	\caption{ \label{fig:protein_local_h2}}
+	\caption{Genome-wide and per-block heritability estimates from LDpred2-auto for 90 protein concentrations \cite[]{folkersen2020genomic}.
+The HapMap3+ variants are used here.
+The maximum local $h^2$ is the maximum heritability estimate across all 431 independent LD blocks defined for this set of variants.
+\label{fig:protein_local_h2}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{height_enrichment}}
-\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations. $N$ is the GWAS sample size. \label{fig:enrichment}}
+\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+\label{fig:enrichment}}
 \end{figure}

### 2022-09-05 paper/paper-infer.tex
save

@@
 
+	TODO
+
 \end{abstract}
@@
 What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
-Some phenotypes, such as height and schizophrenia are highly heritable [TODO: ADD CITATIONS]. 
-These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them [TODO: ADD CITATIONS].
+Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
+These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
 Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
@@
 
-These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone [TODO: ADD CITATIONS]. 
-Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases [TODO: ADD CITATIONS].
+These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone \cite[]{daetwyler2008accuracy,dudbridge2013power}. 
+Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
 LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
 LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model.
-In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two hyper-parameters of the model \cite[]{prive2020ldpred2}.
+In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.
 
 Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
-We show that we can also reliably estimate the predictive ability of PGS derived from LDpred2-auto.
-
-[TO FINISH]
-
-[ALSO TALK ABOUT NEW SET OF VARIANTS]
+So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
+Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS (without needing an independent test set).
+Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
+We call this new preferred set of 1,444,196 SNPs HapMap3+.
 
@@
 \section*{Results}
+
+[TODO: CHOOSE FIGURES TO HAVE IN MAIN TEXT]
 
@@
 Finally, some Gaussian noise is added so that the full phenotype has a variance of 1.
-Then, a GWAS is run to obtain summary statistics using $N$ (either the 200,000 dedicated to this, or a subset of 20,000) individuals using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
+Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
 Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
@@
 Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
-Finally, LDpred2-auto can be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).
+Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).
 
-We also run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
+We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
-The only difference is the transformation to the liability scale that needs to be applied to the $h^2$ and $r^2$ estimates from LDpred2-auto, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ \cite[]{grotzinger2021pervasive}.
+The only difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
 
@@
 Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
-As for the inferred predictive performance, they are highly consistent with the ones derived from the test; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
-Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and these are higher than values reported in the literature.
+As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
+Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and because these estimates are higher than values reported in the literature.
 
-We then investigate previous estimates using two alternative LD references (Methods).
+We then investigate using two alternative LD references (Methods).
 Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
@@
 
-We then investigate previous estimates using the extended set of HapMap3 variants proposed here (Methods).
+We then investigate using the extended set of HapMap3 variants proposed here (Methods).
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
@@
 Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
-However, we do investigate alternative transformations here to decide which one should be preferred and to check the impact on the inference from LDpred2-auto.
-Note that we use the HapMap3+ set of variants here.
-We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates for $p$ and $\alpha$ are highly consistent.
+However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
+Note that we now use the HapMap3+ set of variants here.
+We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
 However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
@@
 
-Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks in Figure \ref{fig:median_local_h2}.
-The median heritability explained by a block of variants is largely proportional to the number of variants in this block.
-Unsurprisingly, the outlier block explaining a much larger heritability contains the HLA region (Figure \ref{fig:median_local_h2}).
+Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
+The outlier block explaining a much larger heritability contains the HLA region.
 Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
-Some variants in multiple small regions across the genome seem to have a much larger probability of being causal; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
+Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
 
@@
 
+[MENTION PARTIAL COR?]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-07 paper/paper-infer.tex
save need discu

@@
 
-[OVERVIEW TABLE OF WHAT LDPRED2 CAN PROVIDE?]
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
+	\caption{[OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
+\end{figure}
 
@@
 We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
-Back to the UKBB data, instead of computing per-block heritability estimates, we compute per-annotation ones, to investigate functional enrichment.
-We perform this analysis for height using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
-Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K (Figure \ref{fig:enrichment}).
 
@@
 Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
+
+\subsection*{Application to height}
+
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals as before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
+Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W.
+Europe, 9.5\% from E.
+Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W.
+Europe, and 0.2\% from W.
+Africa.
+We can therefore use the same N.W. European LD with these GWAS summary statistics. 
+Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).
+
+Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
+SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively.
+And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
+The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
+As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
+Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
+We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
+Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%).
+Therefore, we estimate that height has at least 50,000 causal variants.
+We also identify 1744 SNPs with a probability of being causal larger than 95\%.
+As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
+Finally, we compute per-annotation heritability estimates, to investigate functional enrichment.
+We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
+Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
 
@@
 [MENTION YI'S PAPER]
+
+[COMPARE FA ENRICHMENT TO YENGO2022]
 

### 2022-09-07 paper/paper-infer.tex
save

@@
 These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
-Knowing how much a phenotype is heritable and polygenic can teach a lot about it.
+Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
 Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
-These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.
+These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.
 
@@
 So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
-Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS (without needing an independent test set).
+Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
@@
 	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
-	\caption{[OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
+	\caption{[TODO: ADD AN OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
 \end{figure}
@@
 As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
-Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$, and because these estimates are higher than values reported in the literature.
+Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature.
 
@@
 We then investigate using the extended set of HapMap3 variants proposed here (Methods).
-As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [10.8-13.7]) and lower $p$ (decrease of 11.5\% [10.7-12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
-This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1-8.2]).
-A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502-0.514] instead of 0.324 [0.320-0.329]), which is also reflected in a larger predictive performance ($r^2$ in test set of 0.516 [0.508-0.524] instead of 0.344 [0.335-0.353]).
+As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
+This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
+A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
 Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}).
+However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.
 
 Finally, we investigate different transformations to apply to some continuous phenotypes used here.
-Indeed, 49 of such phenotypes seems log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
+Indeed, 49 of the phenotypes used here seem log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
 However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
@@
 
-In this section, we use the new HapMap3+ set of variants constructed here. 
+In this section, we use the extended set of variants constructed here, HapMap3+. 
 We recall that we have defined 431 independent LD blocks for this set (Methods).
-We compute a per-block (local) $h^2$ estimate, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
+We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals as before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
 Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W.
@@
 As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
-Finally, we compute per-annotation heritability estimates, to investigate functional enrichment.
+Finally, we compute per-annotation heritability estimates to investigate functional enrichment.
 We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
-Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
+Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
 
@@
 
-[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
+\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
+
+\noindent[TODO: EXPORT LDREF FOR NEW HAPMAP3+]
 

### 2022-09-07 paper/paper-infer.tex
discu

@@
 
-[KEEP RG FOR DISCU + CITE EXISTING METHODS]
+LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the (local) SNP-heritability, polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
+It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
+We remind readers that it can also be used to infer the incertainty of individual polygenic scores \cite[]{ding2022large}.
+Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
+We also obtain widespread signatures of selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
+However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}.
+For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x (Figure \ref{fig:enrichment}).
 
-[BETTER SET OF VARIANTS, BUT STILL NOT PERFECT E.G. FOR PROTEIN LEVELS]
+Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
+This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
+Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
+However, we note that we are able to achieve an $r^2$ of 0.67 [TODO: UPDATE NUMBER] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
+It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
 
-[ALSO TALK ABOUT IMPUTATION]
+Our proposed method have limitations.
+First, when power is low, estimates of $\alpha$ and $p$ become less reliable.
+However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
+We think this is due to assortative mating making the estimates too large in the presence of many spouses such as for the UK Biobank [TODO: CITATIONS? YENGO2022? OTHERS?]. 
+Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue.
+[TODO: OTHERS?]
 
-[ALSO TALK ABOUT LDSC?]
-
-[FUTURE WORK + LIMITATION: OVERESTIMATED H2 WITH SMALL LD + NEED TO WORK ON CORRECTION]
-
-[ASSORTATIVE MATING FOR HEIGHT?]
-
-[VERY POLYGENIC, BUT NOT INFINITESIMAL -> THEREFORE LDPRED2-INF NEVER RECOMMENDED (LDPRED2 + MISSPEC)]
-
-[HM3\_PLUS AND MISSING HERITABILITY]
-
-[MENTION YI'S PAPER]
-
-[COMPARE FA ENRICHMENT TO YENGO2022]
+Nevertheless, LDpred2-auto users can now get much from running a single method only.
+The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
+As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, and multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}.
 
@@
 Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
-This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite{zeng2021widespread}. 
+This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite[]{zeng2021widespread}. 
 In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
@@
 
-%\clearpage
+\clearpage
 %\vspace*{3em}
@@
 
+\noindent[TODO: UPDATE LDPRED2 TUTORIAL TO INCLUDE INFERENCE]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-09-08 paper/paper-infer.tex
finished draft

@@
 Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
-These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.
+These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SuSiE (per-variant $p$), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,wang2020simple,zeng2021widespread,johnson2021estimation}.
 
@@
 Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
+An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
@@
 
-\begin{figure}[!h]
-	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
-	\caption{[TODO: ADD AN OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.8\textwidth]{overview}}
+	\caption{Overview of what LDpred2-auto can provide. For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large}. \label{fig:overview}}
 \end{figure}
@@
 Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
+To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
+To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
 
@@
 Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
-However, we note that we are able to achieve an $r^2$ of 0.67 [TODO: UPDATE NUMBER] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
+However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
 It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
@@
 The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
-As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, and multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}.
+As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
 

### 2022-09-27 paper/paper-infer.tex
add new affil

@@
 \title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6}}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}
 
@@
 \noindent$^{\text{\sf 1}}$National Centre for Register-based Research, Aarhus University, Aarhus, Denmark. \\
-\noindent$^{\text{\sf 2}}$Bioinformatics Interdepartmental Program, University of California, Los Angeles, CA. \\
-\noindent$^{\text{\sf 3}}$Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
-\noindent$^{\text{\sf 4}}$Department of Pathology and Laboratory Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
-\noindent$^{\text{\sf 5}}$Department of Computational Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA. \\
+\noindent$^{\text{\sf 2}}$Bioinformatics Interdepartmental Program, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 3}}$Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 4}}$Department of Pathology and Laboratory Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 5}}$Department of Computational Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
 \noindent$^{\text{\sf 6}}$Bioinformatics Research Centre, Aarhus University, Aarhus, Denmark. \\
+\noindent$^{\text{\sf 7}}$Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute, Cambridge, MA, USA. \\
 \noindent$^\ast$To whom correspondence should be addressed. \\
@@
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
-This research has been conducted using the UK Biobank Resource under Application Number 58024; authors thank all the participants of the UK Biobank for contributing to such useful data for Research.
+This research has been conducted using the UK Biobank Resource under Application Number 58024; authors thank all the UK Biobank participants for contributing to such useful data for Research.
 

### 2022-09-28 paper/paper-infer.tex
integrate Clara's comments

@@
 \title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Clara Albi\~nana,$^{\text{1}}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}
 
@@
 
-Most traits and diseases are heritable, if not all. 
+Most traits and diseases are heritable. 
 What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
@@
 These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
-Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
-Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
-These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SuSiE (per-variant $p$), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,wang2020simple,zeng2021widespread,johnson2021estimation}.
+Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
+Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome.
+These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$ that can inform about negative selection, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
 
-These two parameters are useful e.g.\ to determine how well we can predict a phenotype from using genetic variants alone \cite[]{daetwyler2008accuracy,dudbridge2013power}. 
-Such genetic predictors are called polygenic scores (PGS), and are getting closer to be included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
-LDpred2 is a very competitive polygenic score method that can directly builds PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
-LDpred2 uses the (SNP) heritability and polygenicity as parameters of its model.
+As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone. 
+Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
+LDpred2 is a widely used polygenic score method that can directly build PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model.
 In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.
 
-Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
+Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant for fine-mapping purposes), and $\alpha$.
 So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
@@
 
-For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}; we use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
+[TODO: INCLUDE SIMPLIFIED VERSIONS OF FIG S1-S3 HERE?]
+
+For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
 We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e.
@@
 Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
-Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled). 
-LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$.
+Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, in which case it prevents effect sizes from changing sign without going through 0 first).
+LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
+[TODO: EXPLAIN FILTERING OF CHAINS]
 
-LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
+First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
 Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
 When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
-The $\alpha$ estimate becomes unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
-Moreover, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
+Then, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
 Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).
 
-We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a thresholding corresponding to the prevalence.
+We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
@@
 As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
-Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature.
+Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
 
-We then investigate using two alternative LD references (Methods).
+Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
 Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
@@
 
-We then investigate using the extended set of HapMap3 variants proposed here (Methods).
+We then investigate using the extended set of HapMap3 variants proposed here, which includes \textasciitilde40\% more variants than the set of HapMap3 variants recommended to use for LDpred2 (Methods).
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
@@
 
-In this section, we use the extended set of variants constructed here, HapMap3+. 
-We recall that we have defined 431 independent LD blocks for this set (Methods).
+In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
 We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
@@
 Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
-Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
+Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
 To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
 Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W.
@@
 Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%).
+[TODO: RERUN FOR HEIGHT WITH JUMP\_SIGN]
 Therefore, we estimate that height has at least 50,000 causal variants.
-We also identify 1744 SNPs with a probability of being causal larger than 95\%.
+We also identify 1744 SNPs with a probability of being causal larger than 95\% (fine-mapping).
 As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
@@
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the (local) SNP-heritability, polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (global and local), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
 It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
-We remind readers that it can also be used to infer the incertainty of individual polygenic scores \cite[]{ding2022large}.
+We remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
 Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
@@
 However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}.
-For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x (Figure \ref{fig:enrichment}).
+For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
 
@@
 Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
-However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
+However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
 It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
 
-Our proposed method have limitations.
+Our proposed method has limitations.
 First, when power is low, estimates of $\alpha$ and $p$ become less reliable.
 However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
-We think this is due to assortative mating making the estimates too large in the presence of many spouses such as for the UK Biobank [TODO: CITATIONS? YENGO2022? OTHERS?]. 
+We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
 Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue.
@@
 The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
-As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
+As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
 
@@
 
-We construct and use the same phenotypes as in \cite{prive2021high}.
+We construct and use the same phenotypes as in \cite{prive2021high}. [TODO: PROVIDE THE LIST SOMEWHERE]
 About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
@@
 
-F.P.\ and B.J.V.\ are supported by a Lundbeck Foundation Fellowship (R335-2019-2339 to B.J.V.).
+F.P., C.A.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof.\ John McGrath), the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R102-A9118, R155-2014-1724, R248-2017-2003), and a Lundbeck Foundation Fellowship (R335-2019-2339 to B.J.V.).
 

### 2022-09-29 paper/paper-infer.tex
minor comments

@@
 
-Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant for fine-mapping purposes), and $\alpha$.
+Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probability for fine-mapping purposes), and $\alpha$.
 So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{overview}}
-	\caption{Overview of what LDpred2-auto can provide. For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large}. \label{fig:overview}}
+	\caption{Overview of what LDpred2-auto can now provide.
+For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large}.
+HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here, and now recommend to use for LDpred2 (Methods).
+CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting).
+\label{fig:overview}}
 \end{figure}
@@
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
-The only difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
+The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
 
@@
 We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations.
-50,000 of these will be used as test set, while the other 306,409 are used to run a GWAS using linear regression (function \texttt{big\_univLinReg}) for each of all 248 phenotypes and using 8 covariates (Methods).
-Then the new implementation of LDpred2-auto (Methods) is used.
+To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (see Methods for details).
+Then the new implementation of LDpred2-auto is used (see Methods).
 
@@
 
-We then investigate using the extended set of HapMap3 variants proposed here, which includes \textasciitilde40\% more variants than the set of HapMap3 variants recommended to use for LDpred2 (Methods).
+We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2 (Methods).
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
 This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
-A particularly much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
+In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
 Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now run in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.
+However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.
 
@@
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
-We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by one block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
+We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by a single block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
 
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M European individuals [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
-Using the method developed in \cite{prive2021using}, we first infer the ancestry proportions of individuals included in the meta-analysis as 81.9\% from N.W.
-Europe, 9.5\% from E.
-Europe, 6.5\% from Finland, 1.5\% of Ashkenazi ancestry, 0.3\% from S.W.
-Europe, and 0.2\% from W.
-Africa.
-We can therefore use the same N.W. European LD with these GWAS summary statistics. 
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
+We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in
+\cite{prive2021using}, and find that 81.9\% are from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi genetic ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
+We therefore use the same N.W. European LD matrix with these GWAS summary statistics. 
 Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).
@@
 Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
-We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% versus 54.1\%).
+We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% divided by 54.1\%).
 Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%).
@@
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (global and local), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
-It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
-We remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
-Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
-We also obtain widespread signatures of selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
-However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
+We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
+We also remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
+Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
+We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
+However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression (S-LDSC, \cite{finucane2018heritability}).
 For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
+This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
+We also note that this heritability partitioning is performed after running LDpred2-auto for each annotation independently, therefore, unlike S-LDSC, the LDpred2-auto heritability partitioning does not depend on the set of annotations used.
 
@@
 This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
-Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
+Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
 However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
-It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
+This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
 
 Our proposed method has limitations.
-First, when power is low, estimates of $\alpha$ and $p$ become less reliable.
+First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
 However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.

### 2022-09-29 paper/paper-infer.tex
typos

@@
 
-Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probability for fine-mapping purposes), and $\alpha$.
+Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities for fine-mapping purposes), and $\alpha$.
 So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
@@
 Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
-We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in
-\cite{prive2021using}, and find that 81.9\% are from N.W. Europe, 9.5\% from E. Europe, 6.5\% from Finland, 1.5\% of Ashkenazi genetic ancestry, 0.3\% from S.W. Europe, and 0.2\% from W. Africa. 
+We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
+Europe, 9.5\% from E.
+Europe, 6.5\% from Finland, 1.5\% of Ashkenazi genetic ancestry, 0.3\% from S.W.
+Europe, and 0.2\% from W.
+Africa.
 We therefore use the same N.W. European LD matrix with these GWAS summary statistics. 

### 2022-09-30 paper/paper-infer-supp.tex
save need discu
integrate Clara's comments
continue changes

@@
 	\caption{Estimates from LDpred2-auto for either the SNP heritability $h^2$ or the polygenicity $p$ for all 248 phenotypes defined from the UK Biobank.
+Estimates of $h^2$ are constrained to be at least 0.001, and at least $10^{-5}$ for $p$.
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
@@
 \label{fig:alt_LD}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
-	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
-Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here.
-Red dashed lines represent the 1:1 line.
-The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
-The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
-\label{fig:hm3_plus}}
 \end{figure}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
-\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations.
-The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
-\label{fig:enrichment}}
-\end{figure}
-
-\begin{figure}[p]
 	\centerline{\includegraphics[width=\textwidth]{median_local_h2}}
@@
 \caption{Per-variant median probabilities of being causal across 169 UKBB phenotypes.
+Variants with a median probability larger than 0.01 were mapped to genes using R package rsnps (only the largest per peak).
 The HapMap3+ set of variants is used.
@@
 \label{median_postp}}
+\end{figure}
+
+\FloatBarrier
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=\textwidth]{height_enrichment}}
+\caption{Heritability enrichment from LDpred2-auto for height across 50 functional annotations.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+\label{fig:enrichment}}
 \end{figure}

### 2022-09-30 paper/paper-infer.tex
continue changes

@@
 Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
-Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome.
-These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$ that can inform about negative selection, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
+A third parameter, often called $\alpha$ or $S$, can also be used to quantify the effect of negative selection on complex traits \cite[]{schoech2019quantification,zeng2021widespread}.
+Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
+These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
 
@@
 The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
-Then, LDpred2-auto can also infer per-variant probabilities of being causal and per-block (local) heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
 We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
@@
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
-The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2012better}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
+The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
 
@@
 Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
-When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).
+When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes [TODO: CHECK IF ARE THE SAME] for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).
 
-We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2 (Methods).
+We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2, to improve the coverage of this set (Methods).
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
@@
 However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
+Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:hm3_plus}}
+\end{figure}
 
@@
 In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
-We compute per-block (local) $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
+We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
@@
 To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
-To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
+To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
 
@@
 However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
-This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
+This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more localized set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
 
@@
 Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue.
-[TODO: OTHERS?]
 

### 2022-10-05 paper/paper-infer-supp.tex
save

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{est_h2_one}}
-\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
-Horizontal dashed lines represent the true simulated values.
-The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
-The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
-Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-\label{fig:simu_h2}}
-\end{figure}
 
@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_chains}}
+	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_runtimes}}
+	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_h2}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_r2}}
-	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
-These are stratified by the polygenicity estimated from LDpred2-auto.
-Green dashed lines represent the 1:1 line.
-The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept.
-The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
-\label{fig:ukbb_r2}}
-\end{figure}
-
-\begin{figure}[p]
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}
@@
 
-\begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{ukbb_chains}}
-\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
-\end{figure}
-
-\begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{ukbb_runtimes}}
-\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_jumpsign}}
+	\caption{LDpred2-auto estimates for 248 UKBB phenotypes, with or without option `allow\_jump\_sign' enabled.
+HapMap3+ variants are used.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:jumpsign}}
+\end{figure}
+
+\begin{figure}[p]
 	\centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_notransfo}}
 	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their raw versions.
+HapMap3+ variants are used.
 Red dashed lines represent the 1:1 line.
@@
 	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their rank-based inverse normal (RIN) transformed versions.
+HapMap3+ variants are used.
 Red dashed lines represent the 1:1 line.
@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{postp_height}}
+	\caption{Per-variant probabilities of being causal for height.
+These are provided by LDpred2-auto when using the external GWAS summary statistics of 1.6M European individuals from \citet{yengo2022saturated}.
+\label{fig:postp_height}}
+\end{figure}
+
+\begin{figure}[p]
 \centerline{\includegraphics[width=\textwidth]{height_enrichment}}

### 2022-10-05 paper/paper-infer.tex
save

@@
 
-	TODO
+LDpred2 is a widely used Bayesian method for building polygenic scores (PGS).
+LDpred2-auto can infer the two parameters from the LDpred model, $h^2$ and $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
+Here, we present a new version of LDpred2-auto, which adds a third parameter $\alpha$ to its model for modeling negative selection.
+Additional changes are also made to provide better sampling of these parameters.
+We then validate the inference of these three parameters.
+LDpred2-auto also provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes.
+We also derive a new formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto.
+Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average.
 
@@
 
-Most traits and diseases are heritable. 
-What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
-Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
-These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
-Understanding how heritable and polygenic a phenotype is can provide valuable clinical insight and inform future research efforts.
-A third parameter, often called $\alpha$ or $S$, can also be used to quantify the effect of negative selection on complex traits \cite[]{schoech2019quantification,zeng2021widespread}.
-Therefore, many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
+Most traits and diseases in humans are heritable.
+What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero impact on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
+Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme}.
+Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
+The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes [TODO: ADD CITATIONS].
+Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
 These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
@@
 
-[TODO: INCLUDE SIMPLIFIED VERSIONS OF FIG S1-S3 HERE?]
-
 For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
@@
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
-[TODO: EXPLAIN FILTERING OF CHAINS]
+Then only chains that provide top correlations $R \beta$ are kept.
 
@@
 
+\begin{figure}[!h]
+	\centerline{\includegraphics[width=0.95\textwidth]{est_h2_one}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
+Horizontal dashed lines represent the true simulated values.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+\label{fig:simu_h2}}
+\end{figure}
+
 We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
@@
 
-\subsection*{Genetic architectures of 248 phenotypes from the UK Biobank}
+\subsection*{More heritability and predictive accuracy with new set of variants}
 
-We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations.
+We use the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations. 
 To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (see Methods for details).
@@
 
-Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
-Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
-Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
-Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
-As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
-Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
-
-Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
-Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
-When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes [TODO: CHECK IF ARE THE SAME] for which none of the LDpred2-auto chains is usable (Figure \ref{fig:alt_LD}).
-
-We then investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde40\% more variants on top of HapMap3 variants recommended to use for LDpred2, to improve the coverage of this set (Methods).
+We investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the coverage of this set (Methods).
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
@@
 
+\subsection*{Genetic architectures of 248 phenotypes from the UK Biobank}
+
+Here we use the same individuals as in the previous section. We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
+Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
+Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
+Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
+Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
+As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
+Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_r2}}
+	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
+These are stratified by the polygenicity estimated from LDpred2-auto.
+Green dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+``F\_height'' and ``M\_height'' use females and males only, respectively.
+\label{fig:ukbb_r2}}
+\end{figure}
+
+Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
+Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
+When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
+
+Then, we investigate the effect of disabling ``allow\_jump\_sign'' (for extra robustness as shown in \cite{prive2021identifying}) on the estimates from LDpred2-auto.
+Note that we now use the HapMap3+ set of variants here.
+Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ normal sampling, Figure \ref{fig:jumpsign}).
+$h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
+As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here and normal sampling can be used (Figure \ref{fig:jumpsign}).
+
 Finally, we investigate different transformations to apply to some continuous phenotypes used here.
@@
 However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
-Note that we now use the HapMap3+ set of variants here.
+Note that we use the HapMap3+ set of variants here.
 We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries [TODO: ADD NUMBER OF VARIANTS] \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
 We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
@@
 
-Intercepts from LD Score regression are increasing with sample size: 1.01 (SE: 0.007) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
-SNP heritability estimates are of 65.5\% (SE: 2.8), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively.
-And of 60.2\% [95\% CI: 57.4, 63.1], 62.9\% [61.7, 64.0], and 54.1\% [53.8, 54.4] with LDpred2-auto, respectively.
+Intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
+SNP heritability estimates are of 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively.
+And of 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto, respectively.
 The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
-As expected, predictive performance $r^2$ are increasing with sample size, with 30.0\% [29.2, 30.8], 42.7\% [42.2, 43.1], and 46.9\% [46.8, 47.1], respectively.
+As expected, predictive performance $r^2$ are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
 Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
 We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% divided by 54.1\%).
-Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.1\% [1.0, 1.4], 2.2\% [2.0, 2.4], and 5.7\% [5.4, 6.0], consistent with results of simulations with a large polygenicity (p=10\%).
-[TODO: RERUN FOR HEIGHT WITH JUMP\_SIGN]
+Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%).
+These results are similar independent of whether ``allow\_jump\_sign'' is used or not.
 Therefore, we estimate that height has at least 50,000 causal variants.
-We also identify 1744 SNPs with a probability of being causal larger than 95\% (fine-mapping).
-As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.72 [-0.76, -0.68], -0.74 [-0.76, -0.71], and -0.78 [-0.81, -0.75], respectively.
+We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
+As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
 Finally, we compute per-annotation heritability estimates to investigate functional enrichment.
@@
 
-Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
+Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 37\% larger to offer a better coverage of the genome.
 This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
@@
 
-Authors thank members of the NCRR/QGG StatGen group for helpful discussions.
+Authors thank members of the NCRR/QGG StatGen group and Marc-Andr\'e Legault for helpful discussions.
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.

### 2022-10-05 paper/paper-infer.tex
provide pheno desc

@@
 
-[TODO: CHOOSE FIGURES TO HAVE IN MAIN TEXT]
-
 \subsection*{Validating the inference with simulations}
@@
 We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
-As for $\alpha$ estimates from LDpred2-auto, they remain highly consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
+As for $\alpha$ estimates from LDpred2-auto, they remain consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
 Finally, we compute per-annotation heritability estimates to investigate functional enrichment.
@@
 
-We construct and use the same phenotypes as in \cite{prive2021high}. [TODO: PROVIDE THE LIST SOMEWHERE]
+We construct and use the same phenotypes as in \cite{prive2021high}.
 About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
@@
 All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code}.
+Description of phenotypes is available at \url{https://github.com/privefl/paper-infer/tree/master/phenotype_description.xlsx} (also available in TSV format).
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
@@
 
-\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
+%\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
 

### 2022-10-05 paper/paper-infer.tex
no need xlsx
read again

@@
 Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
-The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes [TODO: ADD CITATIONS].
+The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite{speed2012improved}.
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
@@
 
-Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities for fine-mapping purposes), and $\alpha$.
+Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
 So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
@@
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
-We call this new preferred set of 1,444,196 SNPs HapMap3+.
+We call `HapMap3+' this new preferred set of 1,444,196 SNPs.
 
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.8\textwidth]{overview}}
+	\centerline{\includegraphics[width=0.82\textwidth]{overview}}
 	\caption{Overview of what LDpred2-auto can now provide.
-For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large}.
+For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}.
 HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here, and now recommend to use for LDpred2 (Methods).
@@
 
-First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha}).
+First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}, \ref{fig:simu_p} and \ref{fig:simu_alpha}).
 Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
@@
 
-\begin{figure}[!h]
-	\centerline{\includegraphics[width=0.95\textwidth]{est_h2_one}}
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{est_h2_one}}
 	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
@@
 We use the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations. 
-To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (see Methods for details).
-Then the new implementation of LDpred2-auto is used (see Methods).
+To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (Methods).
+Then the new implementation of LDpred2-auto is used (Methods).
 
@@
 Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, fitting with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}) instead of 4--12 hours before.
+However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}), instead of 4--12 hours before.
 
@@
 Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
-As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height and sitting height are they slightly overestimated (Figure \ref{fig:ukbb_r2}).
-Heritability estimates for these two traits are probably slightly overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
+As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height are they overestimated (Figure \ref{fig:ukbb_r2}).
+Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
 
@@
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
-We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 of them have at least 50\% of their heritability explained by a single block, and at least 80\% for 8 of them (Figure \ref{fig:protein_local_h2}).
+We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure \ref{fig:protein_local_h2}).
 
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 305K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 356K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
 We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
@@
 We therefore use the same N.W. European LD matrix with these GWAS summary statistics. 
-Note that we use the HapMap3+ set of 1,444,196 SNPs here, but that we use only 1,013,499 SNPs for the GWAS meta-analysis (those overlapping and passing quality control).
+Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs are overlapping and passing quality control.
 
 Intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
-SNP heritability estimates are of 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively.
-And of 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto, respectively.
-The reduced heritability estimation from the meta-analysis can be partly explained by three reasons: a reduced number of overlapping SNPs, more heterogeneity in the individuals from the meta-analysis compared to within the UK Biobank only, and less effect of assortative mating (which is strong in the UK Biobank due to the presence of many spouses, and probably contributes to overestimating the heritability of height there).
-As expected, predictive performance $r^2$ are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
-Note that the first two estimates (from UKBB) are slightly overestimated, and that the last one is based on a smaller number of overlapping SNPs.
-We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only (46.9\% divided by 54.1\%).
+SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
+As expected, predictive performance $r^2$ (estimated from the Gibbs sampler) are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
+Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
+We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore requiring an even larger sample size to be able to better predict height.
 Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%).
-These results are similar independent of whether ``allow\_jump\_sign'' is used or not.
 Therefore, we estimate that height has at least 50,000 causal variants.
+These results are similar independent of whether ``allow\_jump\_sign'' is used or not, which is surprising to us.
 We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
@@
 Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
+We remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
 We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
-We also remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
 Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
@@
 First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
-However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
+However, estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
 We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
-Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue.
+Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
+Future work could focus on correcting these issues.
 
-Nevertheless, LDpred2-auto users can now get much from running a single method only.
+Nevertheless, LDpred2-auto users can now get much from running a single method.
 The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
@@
 As covariates, for the subset of individuals previously defined, we first recompute PCs using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
-We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates.
+We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).
 
@@
 We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
-Here we propose to extend the set we have used for LDpred2 until then. 
-This extension aims at making sure many genetic variants are well tagged by the extended set.
+Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
 To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
@@
 In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters.
-This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly to data without the need of a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
+This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly without requiring a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
 Previously, $p$ was sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
@@
 In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters.
-This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration.
-We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of correlated variants.
+This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
+We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of variants.
 As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
@@
 The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
-All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code}.
-Description of phenotypes is available at \url{https://github.com/privefl/paper-infer/tree/master/phenotype_description.xlsx} (also available in TSV format).
+All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code} [TODO: UPDATE README].
+Descriptions of UKBB phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.

### 2022-10-05 paper/paper-infer.tex
typo

@@
 Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
-The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite{speed2012improved}.
+The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.

### 2022-10-10 paper/paper-infer.tex
add r2 formula

@@
 \end{equation}
-where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\gamma$ the effect sizes on the allele scale, $S$ the standard deviations of the genotypes, and $\beta$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
+where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\boldsymbol{\gamma}$ the effect sizes on the allele scale, $\boldsymbol{S}$ the standard deviations of the genotypes, and $\boldsymbol{\beta}$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
 In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters.
@@
 
-\subsection*{Inference of predictive performance from the Gibbs sampler}
+\subsection*{Inference of predictive performance $r^2$}
 
-[TODO: EXPLAIN HOW TO INFER R2]
-
-[MENTION PARTIAL COR?]
+To infer the out-of-sample predictive performance $r^2$ (and CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
+Intuitively, if prediction is perfect then $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are the same and $r^2 = h^2$; when power is very low, these two are uncorrelated and $r^2 \approx 0$.
+Note that this assumes the target sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics).
+Although we would have liked to, we do not provide any theoretical justification for this equation.
+Instead, we do check it using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
+These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
+The partial correlation (and 95\% CI) is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign).
 

### 2022-10-10 paper/paper-infer.tex
finish README for code
export code

@@
 The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
-All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code} [TODO: UPDATE README].
-Descriptions of UKBB phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
+Descriptions of UK Biobank phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
+LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper are available at \url{https://doi.org/10.6084/m9.figshare.21305061}.
+All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code}.
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
@@
 
-\noindent[TODO: EXPORT LDREF FOR NEW HAPMAP3+]
+%\noindent[TODO: EXPORT LDREF FOR NEW HAPMAP3+]
 
-\noindent[TODO: UPDATE LDPRED2 TUTORIAL TO INCLUDE INFERENCE]
+%\noindent[TODO: UPDATE LDPRED2 TUTORIAL TO INCLUDE INFERENCE]
 

### 2022-10-10 code/LPA-targeted.R

@@ -1,0 +1,46 @@
+library(bigsnpr)
+library(dplyr)
+
+fam <- snp_attach("data/UKBB_hm3.rds")$fam
+
+# The region used (GRCh37): 6:160952515-161087407 +/- 1 Mb = 6:159952515-162087407
+# + filter out variants with MAF < 0.001 and INFO < 0.5
+info <- bigreadr::fread2("UKBB/mfi/ukb_mfi_chr6_v3.txt")
+snp_id <- info %>%
+  filter(V6 > 1e-3, V8 > 0.5, V3 >= 159952515, V3 <= 162087407) %>%
+  with(paste(6, V3, V4, V5, sep = "_"))
+
+snp_readBGEN(
+  bgenfiles = "UKBB/bgen/ukb_imp_chr6_v3.bgen",
+  backingfile = "tmp-data/UKBB_LPA",
+  ind_row = fam$ind_bgen,
+  list_snp_id = list(snp_id),
+  ncores = nb_cores()
+)
+
+lpa <- snp_attach("tmp-data/UKBB_LPA.rds")
+G <- lpa$genotypes
+covar <- as.matrix(select(fam, -eid, -ind_csv, -ind_bgen, -set))
+y <- readRDS("data/ukbb-quant-pheno/log_lipoA.rds")[fam$ind_csv]
+length(ind_train <- which(!is.na(y) & fam$set == "train" & complete.cases(covar)))  # 232,233
+length(ind_col <- with(lpa$map, which(pmin(freq, 1 - freq) > 1e-3, info > 0.5)))  # 13,572
+
+mod <- runonce::save_run(
+  big_spLinReg(G, y[ind_train], ind.train = ind_train, ind.col = ind_col, K = 10,
+               covar.train = covar[ind_train, ],
+               pf.covar = rep(0, ncol(covar)),
+               power_scale = c(0, 0.5, 1),
+               power_adaptive = c(0, 0.5, 1.5),
+               lambda.min.ratio = 1e-6, nlam.min = 30,
+               n.abort = 3, ncores = nb_cores()),
+  file = "tmp-data/PLR_LPA.rds"
+) # 6H
+
+summary(mod)
+summary(mod)$message
+plot(mod)
+
+ind_test <- which(fam$set == "test")
+
+pred <- predict(mod, G, ind.row = ind_test, covar.row = matrix(0, length(ind_test), ncol(covar)))
+pcor(pred, y[ind_test], covar[ind_test, ])^2  # 0.6765771 0.6711782 0.6819082

### 2022-10-10 code/export-ldref.R

@@ -1,0 +1,47 @@
+library(bigsnpr)
+# library(bigreadr)
+# library(dplyr)
+# library(ggplot2)
+
+#### Filter data from allele frequency errors ####
+
+ukb <- snp_attach("data/UKBB_hm3_plus.rds")
+G <- ukb$genotypes
+map <- transmute(ukb$map,
+                 chr = as.integer(chromosome), pos = physical.pos,
+                 a0 = allele1, a1 = allele2, rsid)  # reversed somehow..
+vctrs::vec_duplicate_any(map[, 1:2])  # FALSE
+
+NCORES <- nb_cores()
+map$af_UKBB <- runonce::save_run(
+  big_colstats(G, ncores = NCORES)$sum / (2 * nrow(G)),
+  file = "tmp-data/af-ldref.rds")
+
+
+# Compute LD scores
+map$ld <- do.call('c', lapply(1:22, function(chr) {
+  cat(chr, ".. ", sep = "")
+  corr_chr <- readRDS(paste0("data/corr_hm3_plus/chr", chr, ".rds"))
+  Matrix::colSums(corr_chr^2)
+}))
+
+
+# Add block IDs
+block_size <- unlist(readRDS("data/corr_hm3_plus/all_final_grp.rds")$all_size)
+map$block_id <- rep(seq_along(block_size), block_size)
+
+
+# add positions in different builds
+liftOver <- runonce::download_file(
+  "http://hgdownload.cse.ucsc.edu/admin/exe/linux.x86_64/liftOver", "tmp-data")
+map$pos_hg18 <- snp_modifyBuild(map, liftOver, from = "hg19", to = "hg18")$pos
+map$pos_hg38 <- snp_modifyBuild(map, liftOver, from = "hg19", to = "hg38")$pos
+
+saveRDS(map, "map_hm3_plus.rds", version = 2)
+
+setwd("data/corr_hm3_plus")
+for(chr in 1:22) {
+  ld_file <- paste0("LD_with_blocks_chr", chr, ".rds")
+  system(paste("zip -0 ../../ldref_hm3_plus.zip", ld_file))
+}
+# 14.1 GB

### 2022-10-10 code/greedy-maxtag.cpp

@@ -1,0 +1,127 @@
+/******************************************************************************/
+
+#include <Rcpp.h>
+#include <unordered_set>
+using namespace Rcpp;
+using namespace std;
+
+// Enable C++11 via this plugin (Rcpp 0.10.3 or later)
+// [[Rcpp::plugins(cpp11)]]
+
+/******************************************************************************/
+
+double compute_add(const std::vector<size_t>& p,
+                   const IntegerVector& i,
+                   const NumericVector& x,
+                   const NumericVector& tagged,
+                   int j) {
+
+  double sum_add = 0;
+
+  for (size_t k = p[j]; k < p[j + 1]; k++) {
+
+    int    ind = i[k];
+    double val = x[k];
+
+    double add = val - tagged[ind];
+    if (add > 0) sum_add += add;
+  }
+
+  return sum_add;
+}
+
+/******************************************************************************/
+
+// [[Rcpp::export]]
+ListOf<NumericVector> greedy_maxtag(const std::vector<size_t>& p,
+                                    const IntegerVector& i,
+                                    const NumericVector& x0,
+                                    const LogicalVector& select,
+                                    const LogicalVector& exclude,
+                                    const NumericVector& sqrt_info,
+                                    double min_add,
+                                    bool remove_diag) {
+
+  NumericVector x = x0 * x0;
+  LogicalVector removed = Rcpp::clone(exclude);
+
+  int m = p.size() - 1;
+  int m_remain = Rcpp::sum(!removed);
+
+  NumericVector can_add(m, 0.0);
+  for (int j = 0; j < m; j++) {
+    size_t lo = p[j];
+    size_t up = p[j + 1];
+    for (size_t k = lo; k != up; k++) {
+      if (remove_diag && i[k] == j) {
+        x[k] = 0;
+      } else {
+        x[k] *= sqrt_info[j];
+        if (!removed[j]) can_add[j] += x[k];
+      }
+    }
+  }
+  can_add[select] = R_PosInf;
+
+  NumericVector tagged(m, 0.0), added(m, NA_REAL);
+
+  std::unordered_set<int> ind_has_changed, ind_could_change;
+
+  do {
+
+    int j_max = which_max(can_add);
+    if (can_add[j_max] < min_add) break;
+
+    // Rcout << j_max + 1 << " -> " << can_add[j_max] << std::endl;
+
+    added[j_max] = can_add[j_max];
[…]

### 2022-10-10 code/height-meta-1600k.R

@@ -1,0 +1,218 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+
+map <- transmute(snp_attach("data/UKBB_hm3_plus.rds")$map,
+                 chr = as.integer(chromosome), pos = physical.pos, rsid, af_val = freq,
+                 a0 = allele2, a1 = allele1)
+
+
+gz <- runonce::download_file(
+  "https://cnsgenomics.com/data/giant_2022/GIANT_HEIGHT_YENGO_2022_GWAS_SUMMARY_STATS_EUR.gz",
+  dir = "tmp-data")
+sumstats <- fread2(
+  gz,
+  select = c("CHR", "POS", "RSID", "EFFECT_ALLELE", "OTHER_ALLELE", "EFFECT_ALLELE_FREQ", "BETA", "SE", "N"),
+  col.names = c("chr", "pos", "rsid", "a0", "a1", "freq", "beta", "beta_se", "n_eff"))
+
+
+# Ancestry inference
+all_freq <- bigreadr::fread2(
+  runonce::download_file("https://figshare.com/ndownloader/files/31620968",
+                         dir = "tmp-data", fname = "ref_freqs.csv.gz"))
+projection <- bigreadr::fread2(
+  runonce::download_file("https://figshare.com/ndownloader/files/31620953",
+                         dir = "tmp-data", fname = "projection.csv.gz"))
+matched <- snp_match(sumstats, all_freq[1:5], return_flip_and_rev = TRUE) %>%
+  mutate(freq = ifelse(`_REV_`, 1 - freq, freq))
+
+res <- snp_ancestry_summary(
+  freq = 1 - matched$freq,  # need to switch ref alleles
+  info_freq_ref = all_freq[matched$`_NUM_ID_`, -(1:5)],
+  projection = projection[matched$`_NUM_ID_`, -(1:5)],
+  correction = c(1, 1, 1, 1.008, 1.021, 1.034, 1.052, 1.074, 1.099,
+                 1.123, 1.15, 1.195, 1.256, 1.321, 1.382, 1.443)
+)
+# Note that some ancestry groups from the reference are very close to one another,
+# and should be merged a posteriori.
+group <- colnames(all_freq)[-(1:5)]
+group[group %in% c("Scandinavia", "United Kingdom", "Ireland")]   <- "Europe (North West)"
[…]

### 2022-10-10 code/height-ukbb-100k.R

@@ -1,0 +1,132 @@
+library(bigsnpr)
+library(dplyr)
+
+pheno_file <- "data/ukbb-quant-pheno/height.rds"
+set <- "hm3_plus"
+
+# Run the GWAS with less people (100K only)
+ukbb <- snp_attach(paste0("data/UKBB_", set, ".rds"))
+G <- ukbb$genotypes
+ind_csv <- ukbb$fam$ind_csv
+covar <- ukbb$fam %>% select(PC1:date) %>% as.matrix()
+y <- readRDS(pheno_file)[ind_csv] + 0
+set.seed(1); ind.train <- sort(sample(which(complete.cases(covar) & !is.na(y)), 100e3))
+
+
+NCORES <- 14
+
+gwas <- runonce::save_run({
+
+  library(future.batchtools)
+  plan(batchtools_slurm(resources = list(
+    t = "12:00:00", c = NCORES + 2, mem = "50g",
+    name = basename(rstudioapi::getSourceEditorContext()$path))))
+  intervals <- bigparallelr::split_len(ncol(G), nb_split = 30)
+
+  furrr::future_map_dfr(rows_along(intervals), function(block) {
+    big_univLinReg(G, y[ind.train], ind.train = ind.train, covar.train = covar[ind.train, ],
+                   ind.col = seq(intervals[block, "lower"], intervals[block, "upper"]),
+                   ncores = NCORES)
+  })
+}, file = "tmp-data/GWAS-height-100K.rds")
+
+df_beta <- gwas %>%
+  transmute(beta = estim, beta_se = std.err, n_eff = length(ind.train))
+
+corr <- readRDS(paste0("data/corr_", set, ".rds"))
+
+
+# Heritability estimation of LD score regression
+# to be used as a starting value in LDpred2-auto
+(ldsc <- snp_ldsc2(corr, df_beta, blocks = 200, intercept = NULL, ncores = NCORES))
+#         int      int_se          h2       h2_se
+# 1.018185266 0.007567497 0.645516599 0.027348155
+
+
+# LDpred2-auto
+coef_shrink <- 0.95
+multi_auto <- runonce::save_run(
+  snp_ldpred2_auto(corr, df_beta, h2_init = pmax(ldsc[["h2"]], 0.001),
+                   vec_p_init = seq_log(1e-4, 0.2, length.out = 50),
[…]

### 2022-10-10 code/height-ukbb-305k.R

@@ -1,0 +1,145 @@
+library(bigsnpr)
+library(dplyr)
+
+pheno_file <- "data/ukbb-quant-pheno/height.rds"
+NCORES <- 13
+
+
+# Get GWAS and its sample size
+pheno <- sub("\\.rds$", "", basename(pheno_file))
+gwas <- purrr::map_dfr(paste0("GWAS/", pheno, "_hm3_plus_part", 1:30, ".rds"), readRDS)
+N <- as.integer(sub(".+, df = ([0-9]+), .+", "\\1", body(attr(gwas, "predict"))[2]))
+
+df_beta <- gwas %>%
+  transmute(beta = estim, beta_se = std.err, n_eff = N)
+
+
+corr <- readRDS("data/corr_hm3_plus.rds")
+
+
+# Heritability estimation of LD score regression
+# to be used as a starting value in LDpred2-auto
+(ldsc <- snp_ldsc2(corr, df_beta, blocks = 200, intercept = NULL, ncores = NCORES))
+#        int     int_se         h2      h2_se
+# 1.11317774 0.01532732 0.59707067 0.02219329
+
+
+# LDpred2-auto
+coef_shrink <- 0.95
+multi_auto <- runonce::save_run(
+  snp_ldpred2_auto(corr, df_beta, h2_init = pmax(ldsc[["h2"]], 0.001),
+                   vec_p_init = seq_log(1e-4, 0.2, length.out = 50),
+                   burn_in = 500, num_iter = 500, report_step = 20,
+                   allow_jump_sign = TRUE, shrink_corr = coef_shrink,
+                   ncores = NCORES),
+  file = "tmp-data/mod_height.rds"
+)
+
+# derive other results
+(range <- sapply(multi_auto, function(auto) diff(range(auto$corr_est))))
+(keep <- which(range > (0.95 * quantile(range, 0.95, na.rm = TRUE)) & range < 3))
+
+hist(all_h2    <- sapply(multi_auto[keep], function(auto) tail(auto$path_h2_est,    500)))
+unname(round(100 * quantile(all_h2, c(0.5, 0.025, 0.975)), 1))  # 63.2 [62.0, 64.4]
+hist(all_p     <- sapply(multi_auto[keep], function(auto) tail(auto$path_p_est,     500)))
+unname(round(100 * quantile(all_p, c(0.5, 0.025, 0.975)), 1))  # 2.3 [2.0, 2.5]
[…]

### 2022-10-10 code/prepare-corr-hm3-altpop.R

@@ -1,0 +1,129 @@
+library(dplyr)
+library(ggplot2)
+library(bigsnpr)
+library(bigreadr)
+NCORES <- nb_cores()
+
+list_snp_id <- with(snp_attach("data/UKBB_hm3.rds")$map,
+                    split(paste(chromosome, physical.pos, allele1, allele2, sep = "_"), chromosome))
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select    = c("eid", "22020-0.0",   paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+
+# Individuals still in data
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("UKBB/w58024_20220222.csv")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv) & df0$used_in_pca)
+
+# Closer to South Europe (than UK)
+PC_UKBB <- select(df0, PC1:PC16)[sub, ]
+all_centers <- read.csv(
+  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
+  stringsAsFactors = FALSE)
+center_italy <- unlist(all_centers[all_centers$Ancestry == "Italy", -1])
+all_sq_dist <- rowSums(sweep(PC_UKBB, 2, center_italy, '-')^2)
+closest_ind <- head(order(all_sq_dist), 10e3)
+hist(log(all_sq_dist), "FD", xlim = c(6, 9)); abline(v = log(all_sq_dist[tail(closest_ind, 1)]), col = "red")
+sub2 <- sub[closest_ind]
+length(sub2) # 10,000
+eid_UK <- df0$eid[sub2]
+
+# Read the data
+snp_readBGEN(
+  bgenfiles = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = 1:22),
+  list_snp_id = list_snp_id,
+  backingfile = "tmp-data/UKBB_hm3_altpop",
+  ind_row = match(df0$eid[sub2], sample$ID_2),
+  ncores = NCORES
+)
+
+
+rm(list = ls())
+
+library(bigsnpr)
[…]

### 2022-10-10 code/prepare-corr-hm3-plus.R

@@ -1,0 +1,83 @@
+library(bigsnpr)
+library(dplyr)
+ukb <- snp_attach("data/UKBB_hm3_plus.rds")
+G     <- ukb$genotypes
+CHR   <- as.integer(ukb$map$chromosome)
+POS   <- ukb$map$physical.pos
+POS2  <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = nb_cores())
+
+library(future.batchtools)
+NCORES <- 14
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("data/corr_hm3_plus")
+
+library(furrr)
+all_final_grp <- future_map_dfr(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  # LD for LDpred2
+  corr0 <- runonce::save_run(
+    snp_cor(G, ind.row = sort(sample(nrow(G), 20e3)), ind.col = ind.chr,
+            infos.pos = POS2[ind.chr], size = 3 / 1000, ncores = NCORES),
+    file = paste0("data/corr_hm3_plus/chr", chr, ".rds")
+  )
+
+  # find nearly independent LD blocks
+  m <- ncol(corr0)
+  (SEQ <- round(seq_log(m / 30, m / 5, length.out = 20)))
+  splits <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 200, max_size = SEQ, max_r2 = 0.2)
+
+  best_split <- splits %>%
+    arrange(cost2 * sqrt(5 + cost)) %>%
+    print() %>%
+    slice(1) %>%
+    print()
+
+  library(ggplot2)
+  plot_grid(
+    qplot(data = splits, perc_kept, cost, color = as.factor(max_size)) +
+      geom_point(data = best_split, size = 2, color = "black") +
+      theme_bw(12) +
+      theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
+      scale_x_continuous(limits = c(0, NA), breaks = seq(0, 1, by = 0.1)) +
+      labs(y = "Sum of squared correlations outside blocks",
+           x = "% of non-zero values kept", color = "Maximum block size"),
+    qplot(data = splits, cost2, cost, color = as.factor(max_size)) +
[…]

### 2022-10-10 code/prepare-corr-hm3-small.R

@@ -1,0 +1,83 @@
+library(bigsnpr)
+library(dplyr)
+ukb <- snp_attach("data/UKBB_hm3.rds")
+G     <- ukb$genotypes
+CHR   <- as.integer(ukb$map$chromosome)
+POS   <- ukb$map$physical.pos
+POS2  <- snp_asGeneticPos(CHR, POS, dir = "tmp-data", ncores = nb_cores())
+
+library(future.batchtools)
+NCORES <- 14
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("data/corr_hm3_small")
+
+library(furrr)
+all_final_grp <- future_map_dfr(1:22, function(chr) {
+
+  ind.chr <- which(CHR == chr)
+
+  # LD for LDpred2
+  corr0 <- runonce::save_run(
+    snp_cor(G, ind.row = sort(sample(nrow(G), 2000)), ind.col = ind.chr,
+            infos.pos = POS2[ind.chr], size = 3 / 1000, ncores = NCORES),
+    file = paste0("data/corr_hm3_small/chr", chr, ".rds")
+  )
+
+  # find nearly independent LD blocks
+  m <- ncol(corr0)
+  (SEQ <- round(seq_log(m / 30, m / 5, length.out = 20)))
+  splits <- snp_ldsplit(corr0, thr_r2 = 0.02, min_size = 200, max_size = SEQ, max_r2 = 0.2)
+
+  best_split <- splits %>%
+    arrange(cost2 * sqrt(5 + cost)) %>%
+    print() %>%
+    slice(1) %>%
+    print()
+
+  library(ggplot2)
+  plot_grid(
+    qplot(data = splits, perc_kept, cost, color = as.factor(max_size)) +
+      geom_point(data = best_split, size = 2, color = "black") +
+      theme_bw(12) +
+      theme(legend.position = "top") + guides(colour = guide_legend(nrow = 1)) +
+      scale_x_continuous(limits = c(0, NA), breaks = seq(0, 1, by = 0.1)) +
+      labs(y = "Sum of squared correlations outside blocks",
+           x = "% of non-zero values kept", color = "Maximum block size"),
+    qplot(data = splits, cost2, cost, color = as.factor(max_size)) +
[…]

### 2022-10-10 code/prepare-geno-hm3-plus.R

@@ -1,0 +1,106 @@
+NCORES <- 50
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 10, mem = "200g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+# plan("multisession", workers = 12)
+
+new_map <- furrr::future_map_dfr(1:22, function(chr) {
+
+  print(chr)
+
+  library(bigsnpr)
+  bkfile <- paste0("tmp-data/hm3_plus/ukbb_hm3_plus_chr", chr, ".bk")
+  rdsfile <- sub_bk(bkfile, ".rds")
+
+  if (!file.exists(rdsfile)) {
+
+    mfi <- bigreadr::fread2(paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt"))
+
+    library(dplyr)
+    snp_id <- mfi %>%
+      filter(V6 > 0.005) %>%
+      with(paste(chr, V3, V4, V5, sep = "_"))
+
+    df0 <- bigreadr::fread2(
+      "UKBB/ukb41181.csv",
+      select = c("eid", "22020-0.0", "22006-0.0"),
+      col.names = c("eid", "used_in_pca", "white_british")
+    )
+
+    # Individuals still in data
+    sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+    ind.indiv <- match(df0$eid, sample$ID_2)
+    id_to_rm <- scan("UKBB/w58024_20220222.csv")
+    ind.indiv[df0$eid %in% id_to_rm] <- NA
+    sub <- which(!is.na(ind.indiv) & df0$used_in_pca & is.na(df0$white_british))
+    length(sub) # 69,630
+
+    # Read dosage data
+    snp_readBGEN(
+      bgenfiles = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = chr),
+      list_snp_id = list(snp_id),
+      backingfile = sub_bk(bkfile),
+      ind_row = ind.indiv[sample(sub, 10e3)],
+      ncores = NCORES
+    )
+  }
+
+  ukbb <- snp_attach(rdsfile)
+  map <- dplyr::transmute(ukbb$map, chr = as.integer(chromosome), pos = physical.pos,
+                          a1 = allele1, a0 = allele2, freq, info)
+
[…]

### 2022-10-10 code/prepare-geno-hm3.R

@@ -1,0 +1,48 @@
+# Get variant info
+info <- readRDS("tmp-data/map_hm3_ldpred2.rds")
+list_snp_id <- with(info, split(paste(chr, pos, a0, a1, sep = "_"), chr))
+
+# same individuals as in simu
+library(bigsnpr)
+fam <- snp_attach("data/ukbb4simu.rds")$fam
+
+system.time(
+  rds <- snp_readBGEN(
+    bgenfiles   = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = 1:22),
+    list_snp_id = list_snp_id,
+    backingfile = "data/UKBB_hm3",
+    ind_row     = fam$ind_bgen,
+    ncores      = nb_cores()
+  )
+)
+
+ukbb <- snp_attach("data/UKBB_hm3.rds")
+
+obj.svd <- runonce::save_run(
+  snp_autoSVD(ukbb$genotypes, infos.chr = as.integer(ukbb$map$chromosome),
+              infos.pos = ukbb$map$physical.pos, ncores = nb_cores()),
+  file = "tmp-data/svd_subset.rds")
+
+plot(obj.svd)  # 4
+plot(obj.svd, type = "scores", scores = 1:6, coeff = 0.6)  # ok for 4
+
+library(dplyr)
+PC <- predict(obj.svd)[, 1:4] %>%
+  as.data.frame() %>%
+  setNames(paste0("PC", 1:4))
+
+other_covar <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select = c("34-0.0", "52-0.0", "22001-0.0", "21022-0.0", "189-0.0"),
+  col.names = c("year", "month", "sex", "age", "deprivation_index")
+) %>%
+  mutate(date = (year - 1900) + (month - 0.5) / 12,
+         year = NULL, month = NULL) %>%
+  slice(fam$ind_csv)
+
+set <- rep("train", nrow(fam))
+set.seed(1); set[sample(length(set), 50e3)] <- "test"
+
+ukbb$fam <- cbind(fam, set, PC, other_covar)
+
+snp_save(ukbb)

### 2022-10-10 code/prepare-geno-simu.R

@@ -1,0 +1,57 @@
+library(dplyr)
+
+# Get variant info
+info <- readRDS(runonce::download_file(
+  "https://figshare.com/ndownloader/files/36360900",
+  dir = "tmp-data", fname = "map_hm3_ldpred2.rds")) %>%
+  filter(chr %% 3 == 0)
+
+list_snp_id <- with(info, split(paste(chr, pos, a0, a1, sep = "_"), chr))
+
+
+# Get indiv info
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select    = c("eid", "22020-0.0",   paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+
+# Individuals still in data
+sample <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")[-1, ]
+ind.indiv <- match(df0$eid, sample$ID_2)
+id_to_rm <- scan("UKBB/w58024_20220222.csv")
+ind.indiv[df0$eid %in% id_to_rm] <- NA
+sub <- which(!is.na(ind.indiv) & df0$used_in_pca)
+
+# Genetically homogeneous
+dist <- bigutilsr::dist_ogk(as.matrix(df0[sub, -(1:2)]))
+hist(log(dist), "FD")
+sub2 <- sub[log(dist) < 4.5]
+length(sub2) # 356,409
+
+
+# Read dosage data
+library(bigsnpr)
+system.time(
+  snp_readBGEN(
+    bgenfiles = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = names(list_snp_id)),
+    list_snp_id = list_snp_id,
+    backingfile = "data/ukbb4simu",
+    ind_row = ind.indiv[sub2],
+    ncores = nb_cores()
+  )
+) # 14 min
+
+simu <- snp_attach("data/ukbb4simu.rds")
+simu$map$chromosome <- as.integer(simu$map$chromosome)
+simu$map$ld <- info$ld
+simu$fam <- data.frame(eid = df0$eid[sub2],
+                       ind_csv = sub2,
+                       ind_bgen = ind.indiv[sub2])
+snp_save(simu)
+
+set.seed(1)
+ind.gwas <- sort(sample(length(sub2), 200e3))
+ind.test <- setdiff(seq_along(sub2), ind.gwas)
+save(ind.gwas, ind.test, file = "data/ukbb4simu_ind.RData")

### 2022-10-10 code/prepare-phecodes.R

@@ -1,0 +1,107 @@
+#### From https://doi.org/10.1016/j.ajhg.2021.11.008 ####
+
+library(bigreadr)
+library(dplyr)
+csv <- "UKBB/ukb41181.csv"
+sex <- fread2(csv, select = "22001-0.0")[[1]]
+
+df_ICD10 <- fread2(csv, colClasses = "character", select = c(
+  paste0("40001-", 0:1, ".0"),     # death
+  paste0("40002-0.", 0:13),        # death
+  paste0("40002-1.", 0:13),        # death
+  paste0("40006-", 0:16, ".0"),    # cancer
+  paste0("41270-0.", 0:212),       # diagnosis
+  paste0("41201-0.", 0:21)         # external cause
+))
+
+# Non-cancer illness code, self-reported
+# coding609 <- fread2("UKBB/coding609.tsv")
+# df_self_reported <- fread2(csv, select = c(paste0("20002-0.", 0:33),
+#                                            paste0("20002-1.", 0:33),
+#                                            paste0("20002-2.", 0:33),
+#                                            paste0("20002-3.", 0:33))) %>%
+#   mutate_all(~ as.character(factor(., levels = coding609$coding,
+#                                    labels = coding609$meaning)))
+
+# Depression ever diagnosed by a professional
+df_dep <- fread2(csv, select = c(paste0("20544-0.", 1:16))) %>%
+  mutate_all(~ ifelse(. == 11, "F330", NA)) %>%
+  select_if(is.character)
+
+df_ICD9 <- fread2(csv, colClasses = "character", select = c(
+  paste0("40013-", 0:14, ".0"),    # cancer
+  paste0("41271-0.", 0:46)         # diagnosis
+))
+
+id_icd10_count <- bind_cols(df_ICD10, df_dep) %>%
+  mutate_all(~ ifelse(. == "", NA, .)) %>%
+  mutate(id = row_number()) %>%
+  tidyr::pivot_longer(-id, values_to = "code", values_drop_na = TRUE) %>%
+  group_by(id, code) %>%
+  summarise(count = n(), .groups = "drop") %>%
+  transmute(id, vocabulary_id = "ICD10", code, count)
+str(id_icd10_count)
+
+id_icd9_count <- df_ICD9 %>%
+  mutate_all(~ ifelse(. == "", NA, .)) %>%
+  mutate(id = row_number()) %>%
[…]

### 2022-10-10 code/prepare-pheno-fields.R

@@ -1,0 +1,750 @@
+#### From https://doi.org/10.1016/j.ajhg.2021.11.008 ####
+
+library(bigreadr)
+
+csv <- "UKBB/ukb41181.csv"
+
+sex <- fread2(csv, select = "22001-0.0")[[1]]
+
+
+#### Quantitative UKBB phenotypes ####
+
+bigassertr::assert_dir("data/ukbb-quant-pheno")
+
+#### Misc ####
+
+# Hand grip strength (mean of both) (46/47)
+y <- rowMeans(fread2(csv, select = c("46-0.0", "47-0.0")))
+y[y < 1] <- NA
+saveRDS(y, "data/ukbb-quant-pheno/hand_grip_strength.rds")
+
+# Waist circumference (48)
+y <- fread2(csv, select = "48-0.0")[[1]]
+y[log(y) < 3.5] <- NA
+saveRDS(log(y), "data/ukbb-quant-pheno/log_waist_circ.rds")
+
+# Hip circumference (49)
+y <- fread2(csv, select = "49-0.0")[[1]]
+y[log(y) < 4] <- NA
+saveRDS(log(y), "data/ukbb-quant-pheno/log_hip_circ.rds")
+
+# Standing height (50)
+y <- fread2(csv, select = "50-0.0")[[1]]
+y[y < 130] <- NA
+saveRDS(y, "data/ukbb-quant-pheno/height.rds")
+
+# sex specific
+yM <- y; yM[sex == 0] <- NA
+saveRDS(yM, "data/ukbb-quant-pheno/M_height.rds")
+yF <- y; yF[sex == 1] <- NA
+saveRDS(yF, "data/ukbb-quant-pheno/F_height.rds")
+
+# Pulse rate, automated reading (102)
+y <- fread2(csv, select = "102-0.0")[[1]]
+saveRDS(log(y), "data/ukbb-quant-pheno/log_pulse_rate.rds")
+
+# Average total household income before tax (738)
+y <- fread2(csv, select = "738-0.0")[[1]]
+y[y < 0] <- NA
+saveRDS(sqrt(y), "data/ukbb-quant-pheno/income.rds")
+
+# Time spent watching television (TV) or using computer (1070/1080)
+y <- fread2(csv, select = c("1070-0.0", "1080-0.0"))
+y[y == -10] <- 0; y[y < 0] <- NA
+y <- rowSums(y)
+y[y > 15] <- NA
[…]

### 2022-10-10 code/run-GWAS.R

@@ -1,0 +1,77 @@
+bigassertr::assert_dir("GWAS")
+
+pheno_files0 <- c(list.files("data/ukbb-quant-pheno", full.names = TRUE),
+                  list.files("data/ukbb-binary-pheno", full.names = TRUE))
+
+grid <- tidyr::expand_grid(
+  set = c("hm3", "hm3_plus"),
+  block = 1:30
+)
+
+NCORES <- 10
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 1, mem = "50g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+furrr::future_pwalk(grid, function(set, block) {
+
+  library(bigsnpr)
+  library(dplyr)
+  ukbb <- snp_attach(paste0("data/UKBB_", set, ".rds"))
+  G <- ukbb$genotypes
+  ind_csv <- ukbb$fam$ind_csv
+  covar <- ukbb$fam %>% select(PC1:date) %>% as.matrix()
+  is_train <- (ukbb$fam$set == "train") & complete.cases(covar)
+
+  intervals <- bigparallelr::split_len(ncol(G), nb_split = 30)
+  ind <- seq(intervals[block, "lower"], intervals[block, "upper"])
+
+  pheno_files <- if (set == "hm3_plus") {
+    pheno_files0
+  } else {
+    # investigate RINT and raw pheno for HM3+ only
+    pheno_files0 %>%
+      grep(pattern = "rint_", value = TRUE, invert = TRUE) %>%
+      grep(pattern = "raw_", value = TRUE, invert = TRUE)
+  }
+
+  lapply(pheno_files, function(pheno_file) {
+
+    pheno <- sub("\\.rds$", "", basename(pheno_file))
+    res_file <- paste0("GWAS/", pheno, "_", set, "_part", block, ".rds")
+
+    if (!file.exists(res_file)) {
+
+      y <- readRDS(pheno_file)[ind_csv] + 0
+      ind.train <- which(is_train & !is.na(y))
+
+      if (length(table(covar[ind.train, 1])) == 1) {
+        COVAR <- covar[, -1]  # rm sex as covariate
+      } else {
+        COVAR <- covar
+      }
+
+      gwas <- big_univLinReg(G, y[ind.train], ind.train = ind.train,
[…]

### 2022-10-10 code/run-ldpred2-cardio-proteins.R

@@ -1,0 +1,159 @@
+library(dplyr)
+urls <- readLines("https://zenodo.org/record/2615265#.YvuAdBxBy01") %>%
+  grep(value = TRUE, pattern = "type=\"application/gzip\" href=\"https://zenodo.org/record/2615265/files/.+\\.txt\\.gz\">") %>%
+  sub(pattern = ".+href=\"(https://zenodo.org/record/2615265/files/.+\\.txt\\.gz)\">", replacement = "\\1")
+
+bigassertr::assert_dir("cardio_proteins")
+
+NCORES <- 13
+library(future.batchtools)
+plan(batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "100g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+# future::plan("sequential")
+
+res <- furrr::future_map(urls, function(url) {
+
+  protein <- sub("\\.txt\\.gz$", "", basename(url))
+  res_file <- paste0("cardio_proteins/", protein, ".rds")
+  runonce::save_run(file = res_file, {
+
+    sumstats <- bigreadr::fread2(runonce::download_file(url, dir = "tmp-data/SCALLOP_CVD1")) %>%
+      transmute(chr = as.integer(sub("(.+):.+:.+_.+", "\\1", MarkerName)),
+                pos = as.integer(sub(".+:(.+):.+_.+", "\\1", MarkerName)),
+                across(c(a1 = Allele1, a0 = Allele2), toupper),
+                freq = Freq1, beta = Effect, beta_se = StdErr, n_eff = TotalSampleSize)
+
+    library(bigsnpr)
+    map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+      transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele2, a1 = allele1, freq)
+
+    (info_snp <- tibble::as_tibble(snp_match(sumstats, map)))
+
+    maf <- map$freq[info_snp$`_NUM_ID_`]
+    sd_val <- sqrt(2 * maf * (1 - maf))
+    sd_ss <- with(info_snp, 1 / sqrt(n_eff * beta_se^2 + beta^2))
+    is_bad <-
+      sd_ss < (0.6 * sd_val) | sd_ss > (sd_val + 0.1) | sd_ss < 0.05 | sd_val < 0.05
+    mean(is_bad)
+
+    df_beta <- info_snp[!is_bad, ] %>%
[…]

### 2022-10-10 code/run-ldpred2-ukbb.R

@@ -1,0 +1,639 @@
+library(bigsnpr)
+library(dplyr)
+
+SETS <- c(
+  "hm3",        # HM3 set from LDpred2
+  "hm3_plus",   # extending prev HM3 to fill gaps of tagging
+  "hm3_small",  # HM3 but with corr based on 2000 indiv only
+  "hm3_altpop"  # HM3 but with corr based on 10K indiv from around South Europe
+)
+
+# build SFBM (same for all pheno)
+for (set in SETS) {
+
+  corr <- runonce::save_run({
+
+    for (chr in 1:22) {
+
+      cat(chr, ".. ", sep = "")
+
+      if (set == "hm3") {
+        corr_chr <- readRDS(paste0("../misspec/ldref/LD_with_blocks_chr", chr, ".rds"))
+      } else {
+        corr_chr <- readRDS(paste0("data/corr_", set, "/LD_with_blocks_chr", chr, ".rds"))
+      }
+
+      if (chr == 1) {
+        corr <- as_SFBM(corr_chr, paste0("data/corr_", set), compact = TRUE)
+      } else {
+        corr$add_columns(corr_chr, nrow(corr))
+      }
+    }
+    corr
+  }, file = paste0("data/corr_", set, ".rds"))
+  print(dim(corr))
+}
+# 1054330 x 1054330
+# 1444196 x 1444196
+# 1054330 x 1054330
+# 1054330 x 1054330
+rm(corr)
+
+pheno_files <- c(list.files("data/ukbb-quant-pheno", full.names = TRUE),
+                 list.files("data/ukbb-binary-pheno", full.names = TRUE))
+
+bigassertr::assert_dir("ldpred2")
+
+grid <- tibble(pheno_file = pheno_files) %>%
+  # slice_sample(n = 20) %>%
+  mutate(is_alt_pheno = grepl("rint_", pheno_file) | grepl("raw_", pheno_file)) %>%
+  tidyr::expand_grid(set = SETS, jump_sign = c(TRUE, FALSE)) %>%
+  filter(ifelse(is_alt_pheno, set == "hm3_plus" & !jump_sign,    TRUE),
+         ifelse(jump_sign,    set == "hm3_plus" & !is_alt_pheno, TRUE)) %>%
+  mutate(res_file = paste0("ldpred2/", set, ifelse(jump_sign, "_jump", ""), "_",
+                           basename(pheno_file)),
+         is_alt_pheno = NULL) %>%
[…]

### 2022-10-10 code/simu-binary.R

@@ -1,0 +1,210 @@
+library(bigsnpr)
+
+#### Run simulations ####
+
+grid <- tidyr::expand_grid(
+  p     = signif(seq_log(0.1, 5e-4, 4), 1),
+  h2    = c(1, 3, 10, 30) / 100,
+  alpha = c(0, -0.5, -1),
+  # N     = c(20, 200) * 1000,
+  K     = c(0.02, 0.2),
+  num   = 1,
+  logistic = c(TRUE, FALSE)
+)
+
+library(future.batchtools)
+NCORES <- 13
+plan(workers = print(nrow(grid)) + 10, batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "60g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("results_simu")
+
+grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num, logistic) {
+
+  print(params <- paste(c(p, h2, alpha, K, num, logistic), collapse = "_"))
+  runonce::save_run(file = paste0("results_simu/res_binary_", params, ".rds"), {
+
+    corr <- readRDS("data/corr_simu.rds")
+
+    simu <- snp_attach("data/ukbb4simu.rds")
+    G <- simu$genotypes
+
+    # simu quantitative pheno
+    simu_pheno <- snp_simuPheno(G, h2 = h2, M = round(ncol(G) * p), K = K,
+                                alpha = alpha, ncores = NCORES)
+
+    # GWAS to get sumstats
+    load("data/ukbb4simu_ind.RData")
+    y <- simu_pheno$pheno[ind.gwas]
+    if (logistic) {
+      gwas <- big_univLogReg(G, y, ind.train = ind.gwas, ncores = NCORES)
+      Neff <- 4 / (1 / sum(y == 1) + 1 / sum(y == 0))
+    } else {
+      gwas <- big_univLinReg(G, y, ind.train = ind.gwas, ncores = NCORES)
+      Neff <- length(ind.gwas)
+    }
+    df_beta <- dplyr::transmute(gwas, beta = estim, beta_se = std.err, n_eff = Neff)
+
+    # ldsc estimate
+    (ldsc <- snp_ldsc2(corr, df_beta, blocks = 100, intercept = NULL, ncores = NCORES))
+
+    # LDpred2-auto
+    coef_shrink <- 0.95
[…]

### 2022-10-10 code/simu.R

@@ -1,0 +1,292 @@
+library(bigsnpr)
+
+#### Prepare LD matrix ####
+
+dim(runonce::save_run({
+
+  for (chr in seq(3, 22, by = 3)) {
+
+    print(chr)
+
+    corr0 <- readRDS(paste0("../misspec/ldref/LD_with_blocks_chr", chr, ".rds"))
+
+    if (chr == 3) {
+      corr <- as_SFBM(corr0, "data/corr_simu", compact = TRUE)
+    } else {
+      corr$add_columns(corr0, nrow(corr))
+    }
+  }
+
+  corr
+}, file = "data/corr_simu.rds"))
+# 322805 x 322805
+
+
+#### Run simulations ####
+
+grid <- tidyr::expand_grid(
+  p     = signif(seq_log(0.1, 5e-4, 4), 1),
+  h2    = c(1, 3, 10, 30) / 100,
+  alpha = c(0, -0.5, -1),
+  N     = c(20, 200) * 1000,
+  num   = 1,
+  jump_sign = c(FALSE, TRUE)
+)
+
+library(future.batchtools)
+NCORES <- 13
+plan(workers = print(nrow(grid)) + 10, batchtools_slurm(resources = list(
+  t = "12:00:00", c = NCORES + 2, mem = "60g",
+  name = basename(rstudioapi::getSourceEditorContext()$path))))
+
+bigassertr::assert_dir("results_simu")
+
+grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, N, num, jump_sign) {
+
+  print(params <- paste(c(p, h2, alpha, N, num, jump_sign), collapse = "_"))
+  runonce::save_run(file = paste0("results_simu/res_", params, ".rds"), {
+
+    corr <- readRDS("data/corr_simu.rds")
+
+    simu <- snp_attach("data/ukbb4simu.rds")
+    G <- simu$genotypes
+
+    load("data/ukbb4simu_ind.RData")
+    ind.gwas0 <- ind.gwas
+
+    # simu quantitative pheno
+    simu_pheno <- snp_simuPheno(G, h2 = h2, M = round(ncol(G) * p), alpha = alpha, ncores = NCORES)
+
+    # GWAS to get sumstats
+    ind.gwas <- sort(sample(ind.gwas0, N))
+    gwas <- big_univLinReg(G, simu_pheno$pheno[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
[…]

### 2022-10-21 code/example-with-provided-LD.R
add example script with new LD

@@ -1,0 +1,126 @@
+library(bigsnpr)
+library(bigreadr)
+
+NCORES <- 15  # TO MODIFY
+
+## Information for the variants provided in the LD reference
+map_ldref <- readRDS(runonce::download_file(
+  "https://figshare.com/ndownloader/files/37802721",
+  dir = "tmp-data", fname = "map_hm3_plus.rds"))
+
+## Breast cancer summary statistics
+
+tgz <- runonce::download_file(
+  "http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST004001-GCST005000/GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+  dir = "tmp-data", fname = "sumstats_BRCA.txt.gz")
+R.utils::gunzip(tgz, overwrite = FALSE, remove = FALSE)
+sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
+                   select = c("chr", "position_b37", "a0", "a1",
+                              "bcac_onco_icogs_gwas_beta",
+                              "bcac_onco_icogs_gwas_se"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se"))
+sumstats$n_eff <- 4 / (1 / 137045 + 1 / 119078)
+
+info_snp <- snp_match(sumstats, map_ldref)
+# 11,792,542 variants to be matched.
+# 110 ambiguous SNPs have been removed.
+# 1,411,710 variants have been matched; 0 were flipped and 6 were reversed.
+(info_snp <- tidyr::drop_na(tibble::as_tibble(info_snp)))
+
+# better to use af of GWAS and INFO scores as well (then can use 0.7 instead of 0.5 in L35)
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2))
+
+is_bad <-
+  sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) | sd_ss < 0.05 | sd_ldref < 0.05
+
+library(ggplot2)
+qplot(sd_ldref, sd_ss, color = is_bad, alpha = I(0.5)) +
+  theme_bigstatsr() +
+  coord_equal() +
+  scale_color_viridis_d(direction = -1) +
+  geom_abline(linetype = 2, color = "red") +
[…]

### 2022-10-21 paper/paper-infer.tex
cite SBayesRC

@@
 This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
+Another possible reason comes from using variants that capture the causal effects by LD, while these tagging variants may be annotated differently from the causal variants, which could cause functional enrichments to be diluted \cite[]{zheng2022leveraging}.
 We also note that this heritability partitioning is performed after running LDpred2-auto for each annotation independently, therefore, unlike S-LDSC, the LDpred2-auto heritability partitioning does not depend on the set of annotations used.
@@
 The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
-As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
+As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to use more variants and learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating,zheng2022leveraging}.
 

### 2023-03-15 paper/paper-infer-supp.tex
update simus

@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{est_p_one}}
+\centerline{\includegraphics[width=0.9\textwidth]{est_p_one}}
 \caption{Inferred polygenicity $p$ in simulations with continuous outcomes.
 Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{est_alpha_one}}
+\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one}}
 \caption{Inferred $\alpha$ in simulations with continuous outcomes.
 Horizontal dashed lines represent the true simulated values.
-Horizontal dotted lines represent boundaries imposed on the estimates.
+Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}).
+Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=\textwidth]{est_r2_one}}
+	\centering
+	\begin{subfigure}[b]{0.7\textwidth}
+		\includegraphics[width=\textwidth]{susie_002}
+		\caption{With $p=0.002$.}
+		\vspace*{1em}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.8\textwidth}
+		\includegraphics[width=\textwidth]{susie_02}
+		\vspace*{-2.6em}
+		\caption{With $p=0.02$.}
+		\vspace*{1em}		
+	\end{subfigure}
+	\caption{Calibration of per-variant posterior probabilities of being causal in two simulations with continuous outcomes, assuming $h^2=0.1$, $\alpha=-0.5$, and $N=10^5$.\label{fig:susie}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one}}
 \caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes.
 Green segments represent $r^2$ in the test set.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{est_binary_h2_one}}
+	\centerline{\includegraphics[width=0.82\textwidth]{est_binary_h2_one}}
 	\caption{Inferred SNP heritability $h^2$ in simulations with binary outcomes.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{est_binary_p_one}}
+	\centerline{\includegraphics[width=0.85\textwidth]{est_binary_p_one}}
 	\caption{Inferred polygenicity $p$ in simulations with binary outcomes. Horizontal dashed lines represent the true simulated values. {\NEFF} \label{fig:simu_p_bin}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{est_binary_alpha_one}}
+	\centerline{\includegraphics[width=0.82\textwidth]{est_binary_alpha_one}}
 	\caption{Inferred $\alpha$ in simulations with binary outcomes.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=\textwidth]{est_binary_r2_one}}
+	\centerline{\includegraphics[width=0.85\textwidth]{est_binary_r2_one}}
 	\caption{Inferred predictive performance $r^2$ in simulations with binary outcomes.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_chains}}
+	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_chains}}
 	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_runtimes}}
+	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_runtimes}}
 	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}

### 2023-03-15 paper/paper-infer.tex
update simus

@@
 \begin{document}
+\setlength{\baselineskip}{1.2\baselineskip}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section*{Introduction}
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.82\textwidth]{overview}}
+	\centerline{\includegraphics[width=0.85\textwidth]{overview}}
 	\caption{Overview of what LDpred2-auto can now provide.
@@
 
+\clearpage
 \section*{Results}
@@
 Finally, some Gaussian noise is added so that the full phenotype has a variance of 1.
-Then, a GWAS is run to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000) using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
-Finally, we run the new LDpred2-auto model (Methods) with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, in which case it prevents effect sizes from changing sign without going through 0 first).
+Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000), using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
+Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the new model including a third parameter $\alpha$  (new option \texttt{use\_MLE}, Methods).
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
-Then only chains that provide top correlations $R \beta$ are kept.
+Then only chains that provide top $R \beta$ (imputed marginal effect sizes) are kept.
 
@@
 Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
-When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$), and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
-The $\alpha$ estimate can become unreliable when power is too low, which can be detected by a small number of chains kept from LDpred2-auto (Figure \ref{fig:simu_alpha}).
+When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$) with LDpred2-auto, and smallish ($p\le0.02$) with SBayesS, and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
+SBayesS often overestimates the polygenicity.
+The $\alpha$ estimate of LDpred2-auto (with the new three-parameter model) can become unreliable or unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
 Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
-We recall that calibrated per-variant probabilities of being causal can be used for fine-mapping purposes  \cite[]{wang2020simple}.
-Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler, even when power is low (Figure \ref{fig:simu_r2}).
+We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIP) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
+LDpred2-auto provide PIPs that are more calibrated than e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block (Figure \ref{fig:susie}).
+Finally, LDpred2-auto (and SBayesS) can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low (Figure \ref{fig:simu_r2}).
 
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=\textwidth]{est_h2_one}}
-	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes.
-Horizontal dashed lines represent the true simulated values.
-The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept.
-The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes. Horizontal dashed lines represent the true simulated values. 
+	For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
+	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+	The 95\% confidence interval for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
 \label{fig:simu_h2}}
@@
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
-The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used when using $N_\text{eff}$ in LDpred2-auto or LD Score regression \cite[]{grotzinger2021pervasive}.
+The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.
 
@@
 
-\subsection*{Local heritability and poligenicity}
+\subsection*{Local heritability and polygenicity}
 
@@
 
+\clearpage
 \section*{Discussion}
@@
 
+\clearpage
 \section*{Materials and Methods}
@@
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-The latest version of R package bigsnpr can be installed from GitHub.%, and a recent enough version can be installed from CRAN [TODO: VERIFY].
+The latest version of R package bigsnpr can be installed from GitHub, and recent enough versions can be installed from CRAN.
 
 %\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
-
-%\noindent[TODO: EXPORT LDREF FOR NEW HAPMAP3+]
-
-%\noindent[TODO: UPDATE LDPRED2 TUTORIAL TO INCLUDE INFERENCE]
 

### 2023-03-15 code/compare-susie.R

@@ -1,0 +1,122 @@
+library(bigsnpr)
+NCORES <- nb_cores()
+
+load("data/ukbb4simu_ind.RData")
+ind.gwas0 <- ind.gwas
+
+simu <- snp_attach("data/ukbb4simu.rds")
+G <- simu$genotypes
+
+p <- 0.002
+# p <- 0.02
+h2 <- 0.1
+alpha <- -0.5
+N <- 100e3
+
+#### Run GWAS and prepare sumstats ####
+
+simu_pheno <- snp_simuPheno(G, h2 = h2, M = round(ncol(G) * p), alpha = alpha,
+                            ncores = NCORES)
+
+# GWAS to get sumstats
+ind.gwas <- sort(sample(ind.gwas0, N))
+gwas <- big_univLinReg(G, simu_pheno$pheno[ind.gwas], ind.train = ind.gwas, ncores = NCORES)
+df_beta <- dplyr::transmute(gwas, beta = estim, beta_se = std.err, n_eff = length(ind.gwas))
+
+
+# Run SuSiE-RSS
+
+blocks <- subset(readRDS("../misspec/ldref/map.rds"), chr %% 3 == 0)$group_id
+list_ind <- split(seq_along(blocks), blocks)
+hist(lengths(list_ind))
+
+bigassertr::assert_dir("tmp-data/corr_susie")
+bigparallelr::set_blas_ncores(NCORES)
+
+purrr::iwalk(list_ind, ~ {
+  print(.y)
+  res_file <- paste0("tmp-data/corr_susie/block_", .y, ".rds")
+  runonce::skip_run_if({
+    R <- big_cor(G, ind.col = .x, ind.row = sample(nrow(G), 20e3))[]
+    saveRDS(R, res_file)
+  }, files = res_file)
+})
+
+Z <- with(df_beta, beta / beta_se)
+N_gwas <- length(ind.gwas)
+system.time(
+  postp2 <- unlist(purrr::imap(list_ind, ~ {
+    print(.y)
+    R <- readRDS(paste0("tmp-data/corr_susie/block_", .y, ".rds"))
+    # print(isSymmetric(R))
+    susieR::susie_rss(Z[.x], R, n = N_gwas, L = 10)$pip
+  }))
+)
+#     user   system  elapsed
+# 3741.050 3300.362  897.720
+# 3479.697 3059.230  770.534
+
+
+# Run LDpred2-auto
+
[…]

### 2023-03-15 code/example-with-provided-LD.R

@@ -14,6 +14,7 @@ tgz <- runonce::download_file(
   "http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST004001-GCST005000/GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
   dir = "tmp-data", fname = "sumstats_BRCA.txt.gz")
 R.utils::gunzip(tgz, overwrite = FALSE, remove = FALSE)
+
 sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
                    select = c("chr", "position_b37", "a0", "a1",
                               "bcac_onco_icogs_gwas_beta",

### 2023-03-15 code/simu-binary.R

@@ -6,88 +6,96 @@ grid <- tidyr::expand_grid(
   p     = signif(seq_log(0.1, 5e-4, 4), 1),
   h2    = c(1, 3, 10, 30) / 100,
   alpha = c(0, -0.5, -1),
-  # N     = c(20, 200) * 1000,
   K     = c(0.02, 0.2),
-  num   = 1,
-  logistic = c(TRUE, FALSE)
+  num   = 1
 )
 
-library(future.batchtools)
 NCORES <- 13
+library(future.batchtools)
 plan(workers = print(nrow(grid)) + 10, batchtools_slurm(resources = list(
   t = "12:00:00", c = NCORES + 2, mem = "60g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
+# future::plan("multisession", workers = 14)
 
-bigassertr::assert_dir("results_simu")
+bigassertr::assert_dir("results_simu_binary")
 
-grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num, logistic) {
+grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num) {
 
-  print(params <- paste(c(p, h2, alpha, K, num, logistic), collapse = "_"))
-  runonce::save_run(file = paste0("results_simu/res_binary_", params, ".rds"), {
+  # p <- 0.02
+  # h2 <- 0.03
+  # alpha <- -0.5
+  # K <- 0.2
+  # num <- 1
+  print(params <- paste(c(p, h2, alpha, K, num), collapse = "_"))
 
-    corr <- readRDS("data/corr_simu.rds")
+  corr <- readRDS("data/corr_simu.rds")
+  simu <- snp_attach("data/ukbb4simu.rds")
+  G <- simu$genotypes
+  load("data/ukbb4simu_ind.RData")
 
-    simu <- snp_attach("data/ukbb4simu.rds")
-    G <- simu$genotypes
-
-    # simu quantitative pheno
-    simu_pheno <- snp_simuPheno(G, h2 = h2, M = round(ncol(G) * p), K = K,
-                                alpha = alpha, ncores = NCORES)
-
-    # GWAS to get sumstats
-    load("data/ukbb4simu_ind.RData")
-    y <- simu_pheno$pheno[ind.gwas]
-    if (logistic) {
-      gwas <- big_univLogReg(G, y, ind.train = ind.gwas, ncores = NCORES)
-      Neff <- 4 / (1 / sum(y == 1) + 1 / sum(y == 0))
[…]

### 2023-03-15 code/simu-methods.R

@@ -1,0 +1,150 @@
+################################################################################
+
+list2 <- function(...) setNames(list(...), sapply(substitute(list(...))[-1], deparse))
+
+r2_boot <- function(x, y) {
+  all_cor <- replicate(1000, {
+    ind <- sample(length(x), replace = TRUE)
+    cor(x[ind], y[ind])
+  })
+  cor <- c(cor(x, y), quantile(all_cor, c(0.025, 0.975)))
+  sign(cor) * cor^2
+}
+
+################################################################################
+
+run_ldpred2 <- function(jump_sign, use_mle, coef_shrink) {
+
+  multi_auto <- snp_ldpred2_auto(corr, df_beta, h2_init = pmax(ldsc[["h2"]], 0.001),
+                                 vec_p_init = seq_log(1e-4, 0.2, length.out = 50),
+                                 burn_in = 500, num_iter = 500, report_step = 20,
+                                 use_MLE = use_mle,
+                                 allow_jump_sign = jump_sign,
+                                 shrink_corr = coef_shrink,
+                                 ncores = NCORES)
+
+  # derive other results
+  (range <- sapply(multi_auto, function(auto) diff(range(auto$corr_est))))
+  (keep <- which(range > (0.95 * quantile(range, 0.95, na.rm = TRUE))))
+  if (length(keep) < 2) return(list(NULL))
+
+  (all_h2    <- sapply(multi_auto[keep], function(auto) tail(auto$path_h2_est,    500)))
+  (all_p     <- sapply(multi_auto[keep], function(auto) tail(auto$path_p_est,     500)))
+  (all_alpha <- sapply(multi_auto[keep], function(auto) tail(auto$path_alpha_est, 500)))
+
+  # inferred r2
+  bsamp <- lapply(multi_auto[keep], function(auto) auto$sample_beta)
+  all_r2 <- do.call("cbind", lapply(seq_along(bsamp), function(ic) {
+    # print(ic)
+    b1 <- bsamp[[ic]]
+    Rb1 <- apply(b1, 2, function(x)
[…]

### 2023-03-15 code/simu.R

@@ -2,7 +2,7 @@ library(bigsnpr)
 
 #### Prepare LD matrix ####
 
-dim(runonce::save_run({
+dim(corr <- runonce::save_run({
 
   for (chr in seq(3, 22, by = 3)) {
 
@@ -21,6 +21,42 @@ dim(runonce::save_run({
 }, file = "data/corr_simu.rds"))
 # 322805 x 322805
 
+# Same data, but format for GCTB
+binfile <- "tmp-data/corr_simu.ldm.sparse.bin"
+infofile <- sub("\\.bin$", ".info", binfile)
+runonce::skip_run_if(files = c(binfile, infofile), {
+
+  con <- file(binfile, open = "wb")
+  system.time(
+    info <- sapply(1:ncol(corr), function(j) {
+      if (j %% 1000 == 0) print(j)
+      one_col <- corr[, j]
+      I <- one_col@i
+      X <- one_col@x
+      L <- length(I)
+      writeBin(as.integer(I), con, size = 4)
+      writeBin(as.double(X),  con, size = 4)
+      c(I[1], I[L], L, sum(X))
+    })
+  ) # 35 min (for 322,805 variants)
+  close(con)
+
+  map <- subset(readRDS("../misspec/ldref/map.rds"), chr %in% seq(3, 22, by = 3))
+  stopifnot(nrow(map) == ncol(corr))
+  info2 <- dplyr::transmute(map, Chrom = chr, ID = rsid, GenPos = 0, PhysPos = pos,
+                            A1 = a1, A2 = a0, A2Freq = af_UKBB,
+                            Index = 0:(ncol(info) - 1),
+                            WindStart = info[1, ],
+                            WindEnd = info[2, ],
+                            WindSize = info[3, ],
+                            WindWidth = -1,
+                            N = 360e3,
+                            SamplVar = -1,
+                            LDsum = info[4, ])
+
+  bigreadr::fwrite2(info2, infofile, sep = " ", scipen = 50)
+})
+
 
 #### Run simulations ####
 
@@ -29,101 +65,68 @@ grid <- tidyr::expand_grid(
   h2    = c(1, 3, 10, 30) / 100,
   alpha = c(0, -0.5, -1),
   N     = c(20, 200) * 1000,
-  num   = 1,
-  jump_sign = c(FALSE, TRUE)
+  num   = 1
 )
 
[…]

### 2023-03-16 paper/paper-infer.tex
compare use_MLE in UKBB analyses

@@
 
-\subsection*{More heritability and predictive accuracy with new set of variants}
+\subsection*{Genetic architectures of 248 phenotypes from the UK Biobank}
 
@@
 To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (Methods).
-Then the new implementation of LDpred2-auto is used (Methods).
+We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} and \texttt{allow\_jump\_sign = FALSE}.
 
-We investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the coverage of this set (Methods).
-As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
-This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
-In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
-Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we recall that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}), instead of 4--12 hours before.
-
-\begin{figure}[!htb]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
-	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
-Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here.
-Red dashed lines represent the 1:1 line.
-The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
-The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
-\label{fig:hm3_plus}}
-\end{figure}
-
-\subsection*{Genetic architectures of 248 phenotypes from the UK Biobank}
-
-Here we use the same individuals as in the previous section. We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
 Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
@@
 
-Then, to investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
+To investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
 Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
 When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
+Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can get chains to converge (and therefore non-zero predictive performance) for the 7 (3 + 4) phenotypes before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
 
 Then, we investigate the effect of disabling ``allow\_jump\_sign'' (for extra robustness as shown in \cite{prive2021identifying}) on the estimates from LDpred2-auto.
-Note that we now use the HapMap3+ set of variants here.
 Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ normal sampling, Figure \ref{fig:jumpsign}).
 $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
-As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here and normal sampling can be used (Figure \ref{fig:jumpsign}).
+As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the normal LD reference) and normal sampling can be used in this case (Figure \ref{fig:jumpsign}).
 
@@
 However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
-Note that we use the HapMap3+ set of variants here.
 We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
-However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation, hinting that it probably makes sense to transform these phenotypes.
+However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation; it probably makes sense to transform these phenotypes.
 We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
 Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.
+
+\subsection*{More heritability and predictive accuracy with new set of variants}
+
+Here we use the same individuals as in the previous section.
+We investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the coverage of this set (Methods).
+As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
+This is consistent with higher predictive performance $r^2$ in the test set (increase of 6.1\% [4.1, 8.2]).
+In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
+Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
+However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}), instead of 4--12 hours before.
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
+Only 154 phenotypes with more than 25 chains kept when using the HapMap3 variants are represented here.
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate (in green) is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:hm3_plus}}
+\end{figure}
 
@@
 Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
+
+\subsection*{Application to other external GWAS summary statistics}
+
+A description of external (to UKBB) GWAS summary statistics used is provided in Table \ref{tab:sumstats}.
+Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
+
+\begin{table}[htb]
+	\centering
+	\begin{tabular}{|l||c|c|c||c|c|}
+		\hline
+		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with HapMap3+\end{tabular} \\
+		\hline
+		Asthma & \citet{demenais2018multiancestry} & ~~67,341.2 & ~~2,001,280 & ~~~946,092 \\
+		Breast cancer (BRCA) & \citet{michailidou2017association} & 254,862.6 & 11,792,542 & 1,411,710 \\
+		Coronary artery disease (CAD) & \citet{nikpay2015comprehensive} & 129,014.3 & ~~9,455,778 & 1,325,052 \\
+		Depression (MDD) [without UKBB] & \citet{wray2018genome} & 168,040.2 & ~~9,874,289 & 1,314,499 \\
+		Prostate cancer (PRCA) & \citet{schumacher2018association} & 135,316.1 & 20,370,946 & 1,411,710  \\
+		Type 1 diabetes (T1D) & \citet{censin2017childhood} & ~~13,497.6  & ~~8,996,866 & 1,127,489  \\
+		Type 2 diabetes (T2D) & \citet{scott2017expanded} & ~~72,143.0  & 12,056,346 & 1,408,283  \\
+		Vitamin D & \citet{jiang2018genome} & ~~79,366~~~ & ~~2,579,296 & 1,028,171 \\
+		\hline
+	\end{tabular}
+	\caption{Summary of external GWAS summary statistics used.  \label{tab:sumstats}}
+\end{table}
 
@@
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-The latest version of R package bigsnpr can be installed from GitHub, and recent enough versions can be installed from CRAN.
+The latest version of R package bigsnpr can be installed from GitHub, and recent enough versions can be installed from CRAN [TODO: REUPLOAD].
 

### 2023-03-16 code/run-GWAS.R

@@ -1,7 +1,7 @@
 bigassertr::assert_dir("GWAS")
 
-pheno_files0 <- c(list.files("data/ukbb-quant-pheno", full.names = TRUE),
-                  list.files("data/ukbb-binary-pheno", full.names = TRUE))
+pheno_files <- c(list.files("data/ukbb-quant-pheno", full.names = TRUE),
+                 list.files("data/ukbb-binary-pheno", full.names = TRUE))
 
 grid <- tidyr::expand_grid(
   set = c("hm3", "hm3_plus"),
@@ -26,15 +26,6 @@ furrr::future_pwalk(grid, function(set, block) {
 
   intervals <- bigparallelr::split_len(ncol(G), nb_split = 30)
   ind <- seq(intervals[block, "lower"], intervals[block, "upper"])
-
-  pheno_files <- if (set == "hm3_plus") {
-    pheno_files0
-  } else {
-    # investigate RINT and raw pheno for HM3+ only
-    pheno_files0 %>%
-      grep(pattern = "rint_", value = TRUE, invert = TRUE) %>%
-      grep(pattern = "raw_", value = TRUE, invert = TRUE)
-  }
 
   lapply(pheno_files, function(pheno_file) {
 

### 2023-03-16 code/run-ldpred2-ukbb.R

@@ -45,14 +45,12 @@ pheno_files <- c(list.files("data/ukbb-quant-pheno", full.names = TRUE),
 bigassertr::assert_dir("ldpred2")
 
 grid <- tibble(pheno_file = pheno_files) %>%
-  # slice_sample(n = 20) %>%
-  mutate(is_alt_pheno = grepl("rint_", pheno_file) | grepl("raw_", pheno_file)) %>%
-  tidyr::expand_grid(set = SETS, jump_sign = c(TRUE, FALSE)) %>%
-  filter(ifelse(is_alt_pheno, set == "hm3_plus" & !jump_sign,    TRUE),
-         ifelse(jump_sign,    set == "hm3_plus" & !is_alt_pheno, TRUE)) %>%
-  mutate(res_file = paste0("ldpred2/", set, ifelse(jump_sign, "_jump", ""), "_",
-                           basename(pheno_file)),
-         is_alt_pheno = NULL) %>%
+  # slice_sample(n = 10) %>%
+  tidyr::expand_grid(set = SETS, use_mle = c(TRUE, FALSE), jump_sign = c(TRUE, FALSE)) %>%
+  filter(set == "hm3" | !jump_sign) %>%
+  mutate(res_file = paste0("ldpred2/", set, ifelse(use_mle, "_MLE", "_noMLE"),
+                           ifelse(jump_sign, "_jump", "_nojump"),
+                           "_", basename(pheno_file))) %>%
   # filter(!file.exists(res_file)) %>%
   print(n = Inf)
 
@@ -62,16 +60,15 @@ library(future.batchtools)
 plan(workers = nrow(grid) + 10, batchtools_slurm(resources = list(
   t = "12:00:00", c = NCORES + 2, mem = "100g",
   name = basename(rstudioapi::getSourceEditorContext()$path))))
+# future::plan("multisession", workers = 14, gc = TRUE)
 
-# future::plan("multisession", workers = 14)
-
-grid$res <- furrr::future_pmap(grid, function(pheno_file, set, jump_sign, res_file) {
+res_grid <- furrr::future_pmap_dfr(grid, function(pheno_file, set, use_mle, jump_sign, res_file) {
 
   # pheno_file <- "data/ukbb-quant-pheno/log_lipoA.rds"
   # set <- "hm3_small"
   # res_file <- paste0("ldpred2/", set, "_", basename(pheno_file))
 
[…]

### 2023-03-23 paper/paper-infer-supp.tex
compare use_MLE in UKBB analyses
external sumstats

@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_chains}}
-	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
-\end{figure}
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_runtimes}}
-	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\FloatBarrier
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_h2}}
+	\centerline{\includegraphics[width=\textwidth]{ukbb_h2}}
 	\caption{SNP heritability estimates from either LDpred2-auto or LD Score regression for all 248 phenotypes defined from the UK Biobank.
@@
 
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_mle}}
+\caption{$r^2$ estimates (in the test set) for UKBB phenotypes, using LDpred2-auto with either the new sampling (MLE, 3-parameter model) or the previous one (2-parameter model).
+Red dashed lines represent the 1:1 line.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:compare_MLE}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	\caption{LDpred2-auto estimates for 248 UKBB phenotypes, with or without option `allow\_jump\_sign' enabled.
-HapMap3+ variants are used.
 Red dashed lines represent the 1:1 line.
@@
 	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their raw versions.
-HapMap3+ variants are used.
 Red dashed lines represent the 1:1 line.
@@
 	\caption{LDpred2-auto estimates for 49 log-transformed UKBB phenotypes, versus for their rank-based inverse normal (RIN) transformed versions.
-HapMap3+ variants are used.
 Red dashed lines represent the 1:1 line.
@@
 \label{fig:RINT}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_chains}}
+	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_runtimes}}
+	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
 \end{figure}
@@
 \FloatBarrier
+
+\begin{figure}[p]
+	\centering
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_Asthma}
+		\caption{Asthma}
+		\vspace*{3em}
+	\end{subfigure}
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_BrCa}
+		\caption{Breast cancer}
+		\vspace*{3em}		
+	\end{subfigure}
+\end{figure}
+\begin{figure}[p]\ContinuedFloat
+	\centering
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_CAD}
+		\caption{Coronary artery disease}
+		\vspace*{3em}		
+	\end{subfigure}
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_MDD}
+		\caption{Depression}
+		\vspace*{3em}		
+	\end{subfigure}
+\end{figure}
+\begin{figure}[p]\ContinuedFloat
+	\centering
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_PrCa}
+		\caption{Prostate cancer}
+		\vspace*{3em}		
+	\end{subfigure}
+	\begin{subfigure}[b]{0.85\textwidth}
+		\includegraphics[width=\textwidth]{ldpred2_T1D}
+		\caption{Type 1 diabetes}
+		\vspace*{3em}		
+	\end{subfigure}
+\end{figure}
+\begin{figure}[p]\ContinuedFloat
+	\centering
+	\begin{subfigure}[b]{0.8\textwidth}
+	\includegraphics[width=\textwidth]{ldpred2_T2D}
+	\caption{Type 2 diabetes}
+	\vspace*{3em}		
+	\end{subfigure}
+	\begin{subfigure}[b]{0.8\textwidth}
+	\includegraphics[width=\textwidth]{ldpred2_VitaminD}
+	\caption{Vitamin D}
+	\vspace*{1em}		
+	\end{subfigure}
+	\caption{LDpred2-auto results with external GWAS summary statistics. The UK Biobank is used as test set to compute $r^2$.\label{fig:sumstats}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\FloatBarrier
 \clearpage

### 2023-03-23 paper/paper-infer.tex
external sumstats

@@
 LDpred2-auto can infer the two parameters from the LDpred model, $h^2$ and $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
-Here, we present a new version of LDpred2-auto, which adds a third parameter $\alpha$ to its model for modeling negative selection.
-Additional changes are also made to provide better sampling of these parameters.
-We then validate the inference of these three parameters.
+Here, we present a new version of LDpred2-auto (as an option), which adds a third parameter $\alpha$ to its model for modeling negative selection.
+We then validate the inference of these three parameters (or two, when using the previous model).
 LDpred2-auto also provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes.
@@
 
-A description of external (to UKBB) GWAS summary statistics used is provided in Table \ref{tab:sumstats}.
+A description of the eight external (to UKBB) GWAS summary statistics used is provided in Table \ref{tab:sumstats}.
 Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
+We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}), and explore multiple values for parameter \texttt{coef\_shrink} (multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix).
+Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
+Results are presented in Figure \ref{fig:sumstats}.
+In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); it seems more useful to use this new set of variants for larger effective sample sizes.
+Results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is a bit surprising to us.
+Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance, except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
+Maximum $r^2$ are acheived at different regularization coefficients across phenotypes.
+As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
+This is also the case for estimates of $p$ when using \texttt{use\_MLE = TRUE}.
+Estimates of $\alpha$ are largely constant but can become very small in the case of robustness issues when using e.g.\ \texttt{use\_MLE = TRUE} and almost no regularization on the LD matrix (i.e.\ \texttt{coef\_shrink} close to 1).
 
@@
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
-We remind readers that it can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, useful for fine-mapping), and the selection-related parameter $\alpha$.
+We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
 We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
@@
 Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 37\% larger to offer a better coverage of the genome.
+Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to define this extended set of variants, we expect it is beneficial across diverse ancestries.
 This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
-Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
+Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% in UKBB analyses here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
 However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
 This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more localized set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
+When using external GWAS summary statistics, using HapMap3+ instead of HapMap3 variants was more beneficial for larger sample sizes.
 
@@
 We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
-Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
+Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
 Future work could focus on correcting these issues.
+Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
+Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric between GWAS summary statistics and the LD reference used.
 
@@
 To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
-We then compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3, and using all unrelated UKBB individuals excluding all White British (field 22006) to have a set of individuals from diverse ancestries.
+Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
 Finally, we design an algorithm which aims at maximizing the tagging of all these variants read.
@@
 This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to define this extended set of variants, we expect it is beneficial across diverse ancestries.
 
-\subsection*{New model and inference with LDpred2-auto}
+\subsection*{New optional model and inference with LDpred2-auto}
 
@@
 This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly without requiring a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
-Previously, $p$ was sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
+$h^2$ is estimated by $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
+As for $p$, it was sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
+Note a small change, we now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more variability in the sampling in order to account for a reduced effective number of independent variants.
 
-Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
-First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
+Here we provide a new model and sampling scheme for LDpred2-auto that can be optionally used by setting parameter \texttt{use\_MLE = TRUE} (otherwise it is run as described in the previous paragraph when using \texttt{use\_MLE = FALSE}).
+In this new option, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
 \begin{equation}\label{eq:new_model}
@@
 In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters.
-This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
-We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of variants.
-As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
-We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
-Note that this $h^2$ estimate can be restricted to e.g.\ variants from a single LD block to get estimates of local heritability.
+This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
+Note that we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}).
+Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
+Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\gamma}^T \boldsymbol{\gamma}$; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
 
@@
 These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
-The partial correlation (and 95\% CI) is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign).
+The partial correlation is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign).
+Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.
 
@@
 
-Authors thank members of the NCRR/QGG StatGen group and Marc-Andr\'e Legault for helpful discussions.
+Authors thank members of the NCRR/QGG StatGen group and Marc-Andr\'e Legault for helpful discussions, as well as reviewers for their useful comments.
 Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.

### 2023-03-23 code/prepare-corr-hm3-plus-regul.R

@@ -1,0 +1,75 @@
+library(bigsnpr)
+library(dplyr)
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map
+CHR <- as.integer(map$chromosome)
+POS <- map$physical.pos
+
+all_final_grp <- readRDS("data/corr_hm3_plus/all_final_grp.rds")
+grid <- all_final_grp %>%
+  select(all_size) %>%
+  mutate(chr = 1:22) %>%
+  tidyr::unnest_longer(all_size) %>%
+  mutate(
+    ind_group = {
+      id_group <- rep(seq_along(all_size), all_size)
+      split(seq_along(id_group), id_group)
+    },
+    id_group = row_number(),
+    all_size = NULL
+  ) %>%
+  # filter(!file.exists(paste0("data/corr_hm3_plus_regul/LD_part", id_group, ".rds"))) %>%
+  print()
+
+
+bigassertr::assert_dir("data/corr_hm3_plus_regul")
+
+library(furrr)
+plan("multisession", workers = 5)
+furrr::future_pwalk(grid, function(chr, ind_group, id_group) {
+
+  # id_group <- 1
+  # chr <- grid$chr[id_group]
+  # ind_group <- grid$ind_group[[id_group]]
+
+  library(Matrix)
+  corr_block <- runonce::save_run({
+    ind.chr <- which(CHR == chr)
+    ind.block <- match(ind_group, ind.chr)
+    readRDS(paste0("data/corr_hm3_plus/LD_with_blocks_chr", chr, ".rds"))[ind.block, ind.block]
+  }, file = paste0("data/corr_hm3_plus_regul/LD_part", id_group, ".rds"))
+
+  new_corr <- runonce::save_run({
+    corrT <- as(corr_block, "dgTMatrix")
+    POS2 <- snp_asGeneticPos(CHR[ind_group], POS[ind_group], dir = "tmp-data")
+    corrT@x <- corrT@x * exp(-0.5 * abs(POS2[corrT@i + 1L] - POS2[corrT@j + 1L]))
+    corrT %>%
+      as("dgCMatrix") %>%
+      as("symmetricMatrix")
+  }, file = paste0("data/corr_hm3_plus_regul/LDregul_part", id_group, ".rds"))
+
+}, .options = furrr_options(scheduling = FALSE))
+
+
[…]

### 2023-03-23 code/prepare-sumstats/Asthma.R

@@ -1,0 +1,54 @@
+# setwd("~/florian/ldpred2-inference/")
+
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Asthma ####
+
+zip <- runonce::download_file(
+  "http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST006001-GCST007000/GCST006862/TAGC_meta-analyses_results_for_asthma_risk.zip",
+  dir = "../sumstats")
+unzip(zip, exdir = "../sumstats", overwrite = FALSE)
+
+
+sumstats <- fread2(
+  "~/florian/sumstats/TAGC meta-analyses results for asthma risk/TAGC_Multiancestry_and_European-Ancestry_Meta-analyses_Results.tsv",
+  select = c("chr", "position", "alternate_allele", "reference_allele",
+             "European_ancestry_beta_fix", "European_ancestry_se_fix"),
+  col.names = c("chr", "pos", "a1", "a0", "beta", "beta_se"))
+
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 2,001,280 variants to be matched.
+# 0 ambiguous SNPs have been removed.
+# 946,092 variants have been matched; 0 were flipped and 0 were reversed.
+
+info_snp$n_eff <- 4 / (1 / 19954 + 1 / 107715)
+
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+is_bad <-
+  sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) | sd_ss < 0.05 | sd_ldref < 0.05
+
+ggplot(slice_sample(data.frame(sd_ldref, sd_ss, is_bad), n = 50e3)) +
+  geom_point(aes(sd_ldref, sd_ss, color = is_bad), alpha = 0.5) +
+  theme_bigstatsr() +
+  coord_equal() +
+  scale_color_viridis_d(direction = -1) +
+  geom_abline(linetype = 2, color = "red") +
[…]

### 2023-03-23 code/prepare-sumstats/BrCa.R

@@ -1,0 +1,57 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Breast cancer (BRCA) ####
+
+tgz <- runonce::download_file(
+  "http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST004001-GCST005000/GCST004988/oncoarray_bcac_public_release_oct17%20(1).txt.gz",
+  dir = "tmp-data", fname = "sumstats_BRCA.txt.gz")
+R.utils::gunzip(tgz, overwrite = FALSE, remove = FALSE)
+
+sumstats <- fread2("tmp-data/sumstats_BRCA.txt", na.strings = "NULL",
+                   select = c("chr", "position_b37", "a0", "a1",
+                              "bcac_onco_icogs_gwas_beta", "bcac_onco_icogs_gwas_se",
+                              "bcac_onco_icogs_gwas_eaf_controls"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se", "freq"))
+sumstats$n_eff <- 4 / (1 / 137045 + 1 / 119078)
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 11,792,542 variants to be matched.
+# 110 ambiguous SNPs have been removed.
+# Some duplicates were removed.
+# 1,411,710 variants have been matched; 0 were flipped and 6 were reversed.
+
+# comparing allele frequencies
+freq2 <- ifelse(info_snp$beta * sumstats$beta[info_snp$`_NUM_ID_.ss`] < 0,
+                1 - info_snp$freq, info_snp$freq)
+hist(diff <- freq2 - info_snp$af_UKBB, "FD", xlim = c(-0.08, 0.08))
+
+# better to use af of GWAS and INFO scores as well (then can use 0.7 instead of 0.5 in L35)
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+is_bad <-
[…]

### 2023-03-23 code/prepare-sumstats/CAD.R

@@ -1,0 +1,70 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Coronary artery disease (CAD) ####
+
+txt <- runonce::download_file(
+  "http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/GCST003001-GCST004000/GCST003116/cad.add.160614.website.txt",
+  dir = "tmp-data", fname = "sumstats_CAD.txt")
+sumstats <- fread2(txt,
+                   select = c("chr", "bp_hg19", "noneffect_allele",
+                              "effect_allele", "beta", "se_dgc", "p_dgc",
+                              "effect_allele_freq", "median_info", "n_studies", "het_pvalue"),
+                   col.names = c("chr", "pos", "a0", "a1", "beta", "beta_se",
+                                 "p", "freq", "info", "n_studies", "het_pvalue"))
+
+study_size <- c(
+  # from Supp Table 1
+  "5719 6545","206 259","278 312","505 1021","392 410","1010 3998",
+  "1628 368","2083 2048","1216 653","658 5841","1802 466","2099 2690",
+  "634 1608","1207 1288","1061 1467","1089 1147","877 2187",
+  "2700 2758","361 2778","487 1381","758 3337","2791 3757","2095 503",
+  "933 468","2905 2998","947 1008","1294 1529","843 318","933 468",
+  "119 830","631 334","836 761","426 594","814 5999","322 857",
+  "1926 2938","4651 4452","4380 3929","1535 772","1007 22286","402 448",
+  "745 1389","397 2474","506 5335","259 4202","334 3446","2034 3210","454 8443")
+study_size2 <- fread2(text = study_size, col.names = c("Nca", "Nco"))
+nrow(study_size2)   # 48
+colSums(study_size2)   # Nca: 61289 -- Nco: 126310  (slightly different from 60801 -- 123504)
+(sumstats$n_eff <- sum(4 / rowSums(1 / study_size2)))  # 129014.3
+4 / (1 / 60801 + 1 / 123504)  # 162972.6
+
[…]

### 2023-03-23 code/prepare-sumstats/MDD.R

@@ -1,0 +1,74 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### MDD ####
+
+gz <- runonce::download_file(
+  "https://figshare.com/ndownloader/files/34427408",
+  dir = "../sumstats", fname = "daner_pgc_mdd_meta_w2_no23andMe_rmUKBB.gz")
+sumstats <- fread2(gz,
+                   select = c("CHR", "BP", "A1", "A2", "OR", "SE",
+                              "INFO", "FRQ_A_45396", "FRQ_U_97250", "Neff_half"),
+                   col.names = c("chr", "pos", "a1", "a0", "or", "beta_se",
+                                 "info", "freq1", "freq2", "Neff_half")) %>%
+  as_tibble() %>%
+  mutate(beta = log(or), or = NULL, chr = as.integer(chr),
+         freq = (freq1 * 45396 + freq2 * 97250) / (45396 + 97250),
+         freq1 = NULL, freq2 = NULL,
+         n_eff = 2 * Neff_half, Neff_half = NULL,
+         info = pmin(info, 1)) %>%
+  filter(n_eff > (0.5 * max(n_eff)),
+         info > 0.4) %>%
+  print()
+
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 9,343,125 variants to be matched.
+# 76 ambiguous SNPs have been removed.
+# 1,314,499 variants have been matched; 0 were flipped and 612,915 were reversed.
+
+
+info_snp2 <- info_snp %>%
+  mutate(sd_af = sqrt(2 * freq * (1 - freq)),
+         sd_ss = 2 / sqrt(n_eff * beta_se^2 + beta^2),
+         sd_ss2 = sd_ss / sqrt(info))
+
+info_snp2$freq2 <- ifelse(info_snp2$beta * sumstats$beta[info_snp2$`_NUM_ID_.ss`] < 0,
+                          1 - info_snp2$freq, info_snp2$freq)
+
+qplot(af_UKBB, freq2, alpha = I(0.5),
+      data = slice_sample(info_snp2, n = 100e3)) +
+  theme_bigstatsr() +
+  coord_equal() +
[…]

### 2023-03-23 code/prepare-sumstats/PrCa.R

@@ -1,0 +1,64 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Prostate cancer (PRCA) ####
+
+zip <- runonce::download_file(
+  "http://practical.icr.ac.uk/blog/wp-content/uploads/uploadedfiles/oncoarray/MetaSummaryData/meta_v3_onco_euro_overall_ChrAll_1_release.zip",
+  dir = "tmp-data", fname = "sumstats_PRCA.zip")
+unzip(zip, exdir = "tmp-data", overwrite = FALSE)
+
+sumstats <- fread2(
+  "tmp-data/meta_v3_onco_euro_overall_ChrAll_1_release.txt",
+  select = c("Chr", "position", "Allele1", "Allele2", "Effect", "StdErr",
+             "Pvalue", "Freq1", "OncoArray_imputation_r2"),
+  col.names = c("chr", "pos", "a1", "a0", "beta", "beta_se", "p", "freq", "info")
+) %>%
+  mutate(a0 = toupper(a0), a1 = toupper(a1)) %>%
+  filter(info > 0.4, beta_se > 0)
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 11,792,542 variants to be matched.
+# 110 ambiguous SNPs have been removed.
+# Some duplicates were removed.
+# 1,411,710 variants have been matched; 0 were flipped and 6 were reversed.
+
+sizes <- bigreadr::fread2(text = c("44825\t27904","20219\t20440","1854\t1894","3650\t3940",
+                                   "474\t482","1458\t512","2068\t2993", "4600\t2941"))
+info_snp$n_eff <- print(sum(4 / rowSums(1 / sizes)))  # 135316.1
+4 / (1 / 79148 + 1 / 61106)  # 137933.1
+
+# comparing allele frequencies
+freq2 <- ifelse(info_snp$beta * sumstats$beta[info_snp$`_NUM_ID_.ss`] < 0,
+                1 - info_snp$freq, info_snp$freq)
+hist(diff <- freq2 - info_snp$af_UKBB, "FD", xlim = c(-0.08, 0.08))
+
[…]

### 2023-03-23 code/prepare-sumstats/T1D.R

@@ -1,0 +1,56 @@
+# setwd("~/florian/ldpred2-inference/")
+
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Type 1 diabetes (T1D) ####
+
+# Download at https://datadryad.org/stash/dataset/doi:10.5061/dryad.ns8q3
+# unzip("../sumstats/doi_10.5061_dryad.ns8q3__v1.zip", exdir = "../sumstats/T1D")
+
+# Affymetrix
+sumstats <- fread2(
+  paste0("../sumstats/T1D/meta_chr_", 1:22),
+  select = c("chromosome", "position", "a0", "a1",
+             "info_score.A", "info_score.I", "beta.meta", "se.meta"),
+  col.names = c("chr", "pos", "a0", "a1", "info1", "info2", "beta", "beta_se")) %>%
+  filter(info1 > 0.4, info2 > 0.4, beta_se > 0) %>%
+  select(-info1, -info2)
+
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 8,531,891 variants to be matched.
+# 79 ambiguous SNPs have been removed.
+# Some duplicates were removed.
+# 1,127,489 variants have been matched; 519 were flipped and 1,769 were reversed.
+
+info_snp$n_eff <- 4 / (1 / 1930 + 1 / (1455 + 1490 + 1884)) + 4 / (1 / 3983 + 1 / 3999)
+
+
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+is_bad <-
+  sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) | sd_ss < 0.1 | sd_ldref < 0.05
+
+ggplot(slice_sample(data.frame(sd_ldref, sd_ss, is_bad), n = 50e3)) +
+  geom_point(aes(sd_ldref, sd_ss, color = is_bad), alpha = 0.5) +
+  theme_bigstatsr() +
+  coord_equal() +
[…]

### 2023-03-23 code/prepare-sumstats/T2D.R

@@ -1,0 +1,46 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Type 2 diabetes (T2D) ####
+
+# Download from http://diagram-consortium.org/downloads.html
+# DIAGRAM 1000G GWAS meta-analysis Stage 1 Summary statistics
+# Published in Scott et al (2017)
+# unzip("../sumstats//METAANALYSIS_DIAGRAM_SE1.zip", exdir = "../sumstats")
+sumstats <- fread2("../sumstats//METAANALYSIS_DIAGRAM_SE1.txt")
+sumstats <- tidyr::separate(sumstats, "Chr:Position", c("chr", "pos"), convert = TRUE)
+names(sumstats) <- c("chr", "pos", "a1", "a0", "beta", "beta_se", "p", "N")
+sumstats$n_eff <- 72143
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 12,056,346 variants to be matched.
+# 89 ambiguous SNPs have been removed.
+# 1,408,283 variants have been matched; 154 were flipped and 650,482 were reversed.
+
+sd_ldref <- with(info_snp, sqrt(2 * af_UKBB * (1 - af_UKBB)))
+sd_ss <- with(info_snp, 2 / sqrt(n_eff * beta_se^2 + beta^2))
+
+is_bad <-
+  sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) | sd_ss < 0.05 | sd_ldref < 0.05
+
+ggplot(slice_sample(data.frame(sd_ldref, sd_ss, is_bad), n = 50e3)) +
+  geom_point(aes(sd_ldref, sd_ss, color = is_bad), alpha = 0.5) +
+  theme_bigstatsr() +
+  coord_equal() +
+  scale_color_viridis_d(direction = -1) +
+  geom_abline(linetype = 2, color = "red") +
+  labs(x = "Standard deviations derived from allele frequencies of the LD reference",
+       y = "Standard deviations derived from the summary statistics",
+       color = "Removed?")
+
+mean(is_bad)  # 0.02798159
+
[…]

### 2023-03-23 code/prepare-sumstats/VitaminD.R

@@ -1,0 +1,51 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+map <- snp_attach("data/UKBB_hm3_plus.rds")$map %>%
+  transmute(chr = as.integer(chromosome), pos = physical.pos, a0 = allele1, a1 = allele2,
+            af_UKBB = freq)
+
+
+#### Vitamin D ####
+
+# See google drive link in https://doi.org/10.1038/s41467-017-02662-2
+# Should have used Allele1 -> a1 and Allele2 -> a0
+sumstats <- fread2("../sumstats/upload_25HydroxyVitaminD_QC.METAANALYSIS1.txt",
+                   select = c("Chr", "Pos", "MarkerName", "Allele1", "Allele2",
+                              "Effect", "StdErr", "SAMPLESIZE", "HetPVal"),
+                   col.names = c("chr", "pos", "rsid", "a1", "a0",
+                                 "beta", "beta_se", "n_eff", "het_pval")) %>%
+  mutate_at(c("a0", "a1"), toupper)
+
+info_snp <- as_tibble(bigsnpr::snp_match(sumstats, map))
+# 2,543,566 variants to be matched.
+# 35 ambiguous SNPs have been removed.
+# 1,028,171 variants have been matched; 6,370 were flipped and 498,902 were reversed.
+
+hist(-log10(info_snp$het_pval), "FD")
+
+info_snp2 <- info_snp %>%
+  mutate(sd_ldref = sqrt(2 * af_UKBB * (1 - af_UKBB)),
+         sd_ss = 1 / sqrt(n_eff * beta_se^2 + beta^2),
+         sd_ss = sd_ss / quantile(sd_ss, 0.99) * sqrt(0.5))
+
+info_snp2$is_bad <- with(info_snp2, het_pval < 1e-5 |
+                           sd_ss < (0.5 * sd_ldref) | sd_ss > (sd_ldref + 0.1) |
+                           sd_ss < 0.05 | sd_ldref < 0.05)
+
+ggplot(slice_sample(info_snp2, n = 50e3)) +
+  geom_point(aes(sd_ldref, sd_ss, color = is_bad), alpha = 0.5) +
+  theme_bigstatsr() +
+  coord_equal() +
+  scale_color_viridis_d(direction = -1) +
+  geom_abline(linetype = 2, color = "red") +
+  labs(x = "Standard deviations derived from allele frequencies of the LD reference",
[…]

### 2023-03-23 code/run-ldpred2-sumstats.R

@@ -1,0 +1,211 @@
+library(bigsnpr)
+library(dplyr)
+
+grid <- tibble::tribble(
+  ~ pheno,    ~ pheno_file,
+  "CAD",      "data/ukbb-binary-pheno/411.4.rds",
+  "PrCa",     "data/ukbb-binary-pheno/185.rds",
+  "BrCa",     "data/ukbb-binary-pheno/174.1.rds",
+  "T1D",      "data/ukbb-binary-pheno/250.1.rds",
+  "T2D",      "data/ukbb-binary-pheno/250.2.rds",
+  "MDD",      "data/ukbb-binary-pheno/296.2.rds",
+  "Asthma",   "data/ukbb-binary-pheno/495.rds",
+  "VitaminD", "data/ukbb-quant-pheno/log_vitaminD.rds"
+) %>%
+  mutate(K = purrr::map_dbl(pheno_file, ~ {
+    y <- readRDS(.)
+    `if`(is.logical(y), mean(y, na.rm = TRUE), NA) })) %>%
+  print() %>%
+  #   pheno    pheno_file                                    K
+  # 1 CAD      data/ukbb-binary-pheno/411.4.rds        0.0609
+  # 2 PrCa     data/ukbb-binary-pheno/185.rds          0.0508
+  # 3 BrCa     data/ukbb-binary-pheno/174.1.rds        0.0731
+  # 4 T1D      data/ukbb-binary-pheno/250.1.rds        0.00867
+  # 5 T2D      data/ukbb-binary-pheno/250.2.rds        0.0609
+  # 6 MDD      data/ukbb-binary-pheno/296.2.rds        0.0756
+  # 7 Asthma   data/ukbb-binary-pheno/495.rds          0.0763
+  # 8 VitaminD data/ukbb-quant-pheno/log_vitaminD.rds NA
+  tidyr::expand_grid(set = c("hm3", "hm3_plus", "hm3_plus_regul"),
+                     mle = c(TRUE, FALSE),
+                     coef_shrink = seq(0.3, 1, by = 0.1)) %>%
+  mutate(res_file = paste0("ldpred2-ext/", pheno, "_", set, ifelse(mle, "", "_noMLE"),
+                           "_", coef_shrink, ".rds")) %>%
+  # filter(!file.exists(res_file)) %>%
+  relocate(K, .after = last_col())
+
+bigassertr::assert_dir("ldpred2-ext")
+
+NCORES <- 13
[…]

### 2023-03-24 paper/paper-infer-supp.tex
reread

@@
 \centerline{\includegraphics[width=\textwidth]{postp_calib}}
-\caption{Calibration of per-variant posterior probabilities of being causal in simulations with continuous outcomes.
+\caption{Calibration of per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities) in simulations with continuous outcomes.
+These probabilities are binned on a log-scale, and mean in each bin is compared to the proportion of simulated causal variants in the bin (with 95\% CI).
 Green dashed lines represent the 1:1 line.
@@
 \caption{Calibration of per-block heritability estimates in simulations with continuous outcomes. 
-Green dashed lines represent the 1:1 line.
-Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented.
-\label{fig:simu_h2_calib}}
+Local $h^2$ estimate for each block is compared to the true simulated local heritability in the block.
+Green dashed lines represent the 1:1 line. Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented. \label{fig:simu_h2_calib}}
 \end{figure}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.82\textwidth]{est_binary_h2_one}}
+	\centerline{\includegraphics[width=0.83\textwidth]{est_binary_h2_one}}
 	\caption{Inferred SNP heritability $h^2$ in simulations with binary outcomes.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{est_binary_p_one}}
+	\centerline{\includegraphics[width=0.9\textwidth]{est_binary_p_one}}
 	\caption{Inferred polygenicity $p$ in simulations with binary outcomes. Horizontal dashed lines represent the true simulated values. {\NEFF} \label{fig:simu_p_bin}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.82\textwidth]{est_binary_alpha_one}}
+	\centerline{\includegraphics[width=0.83\textwidth]{est_binary_alpha_one}}
 	\caption{Inferred $\alpha$ in simulations with binary outcomes.
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{est_binary_r2_one}}
+	\centerline{\includegraphics[width=0.88\textwidth]{est_binary_r2_one}}
 	\caption{Inferred predictive performance $r^2$ in simulations with binary outcomes.
@@
 The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+``use\_mle: TRUE'' corresponds to using the new 3-parameter model and sampling scheme (Methods).
 \label{fig:ukbb_h2}}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_chains}}
-	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes. \label{fig:ukbb_chains}}
+	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes.
+``Use MLE'' corresponds to using the new 3-parameter model and sampling scheme (Methods).
+\label{fig:ukbb_chains}}
 \end{figure}
@@
 	\centerline{\includegraphics[width=0.95\textwidth]{ukbb_runtimes}}
-	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes. For each phenotype, 50 chains were used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
+	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes.
+	``Use MLE'' corresponds to using the new 3-parameter model and sampling scheme (Methods). For each phenotype, 50 chains are used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
 \end{figure}
@@
 	\centering
-	\begin{subfigure}[b]{0.8\textwidth}
+	\begin{subfigure}[b]{0.75\textwidth}
 	\includegraphics[width=\textwidth]{ldpred2_T2D}
@@
 	\end{subfigure}
-	\begin{subfigure}[b]{0.8\textwidth}
+	\begin{subfigure}[b]{0.75\textwidth}
 	\includegraphics[width=\textwidth]{ldpred2_VitaminD}
@@
 	\end{subfigure}
-	\caption{LDpred2-auto results with external GWAS summary statistics. The UK Biobank is used as test set to compute $r^2$.\label{fig:sumstats}}
+	\caption{LDpred2-auto results with external GWAS summary statistics.
+We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}, where setting to \texttt{TRUE} uses the new model), and explore multiple values for parameter \texttt{coef\_shrink} (multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix).
+The UK Biobank is used as test set to compute $r^2$.\label{fig:sumstats}}
 \end{figure}

### 2023-03-24 paper/paper-infer.tex
reread

@@
 We then validate the inference of these three parameters (or two, when using the previous model).
-LDpred2-auto also provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes.
+We also show that LDpred2-auto provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes.
 We also derive a new formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto.
-Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average.
+Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average, in UK Biobank analyses.
 
@@
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
-These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$ used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (per-variant $p$ used for fine-mapping, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
+These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$, also called posterior inclusion probabilities PIP, used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (PIP, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
 
-As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone. 
+As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, where $r^2_\text{max} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$ ($N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants).
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
@@
 
-Here we extend LDpred2-auto to make it a highly reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
-So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
+Here we extend LDpred2-auto and show it is a reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities PIP used for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
+So, on top of providing competitive PGS, LDpred2-auto can also provide all these estimates of genetic architecture. 
 Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
@@
 Finally, some Gaussian noise is added so that the full phenotype has a variance of 1.
-Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000), using a simple linear regression implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
+Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
 Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the new model including a third parameter $\alpha$  (new option \texttt{use\_MLE}, Methods).
@@
 When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
-When power is low, the polygenicity can be overestimated when the true value is very small (e.g.\ $p=0.0005$) with LDpred2-auto, and smallish ($p\le0.02$) with SBayesS, and underestimated when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
-SBayesS often overestimates the polygenicity.
-The $\alpha$ estimate of LDpred2-auto (with the new three-parameter model) can become unreliable or unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
+When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
+SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
+The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become unreliable or unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
 Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
@@
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} and \texttt{allow\_jump\_sign = FALSE}.
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the new 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE}.
 
-Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a smaller polygenicity (Figure \ref{fig:ukbb_h2}).
+Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure \ref{fig:ukbb_h2}).
 Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
 Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
-Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}), which is consistent with widespread negative selection.
+Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}).
 As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height are they overestimated (Figure \ref{fig:ukbb_r2}).
@@
 The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
-``F\_height'' and ``M\_height'' use females and males only, respectively.
+``F\_height'' and ``M\_height'' use females and males only, respectively (in both GWAS and test sets).
 \label{fig:ukbb_r2}}
@@
 To investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
-Using a smaller number of individuals for computing the LD results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
+Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
 When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
-Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can get chains to converge (and therefore non-zero predictive performance) for the 7 (3 + 4) phenotypes before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
+Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can get chains to converge (and therefore non-zero predictive performance) for the 7 (3 + 4) phenotypes mentioned before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
 
@@
 Here we use the same individuals as in the previous section.
-We investigate using the extended set of HapMap3 variants proposed here, HapMap3+, which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the coverage of this set (Methods).
+We investigate using the extended set of HapMap3 variants proposed here, HapMap3+ (Methods), which includes \textasciitilde37\% more variants on top of HapMap3 variants recommended to use for LDpred2 (i.e.\ 1,054,330 + 389,866 variants), to improve the genome coverage of this set.
 As expected, compared to HapMap3, higher $h^2$ (average increase of 12.3\% [95\% CI: 10.8, 13.7]) and lower $p$ (decrease of 11.5\% [10.7, 12.3]) estimates are obtained with this extended set HapMap3+ (Figure \ref{fig:hm3_plus}).
@@
 Africa.
-We therefore use the same N.W. European LD matrix with these GWAS summary statistics. 
+For this set of external GWAS summary statistics, we therefore use the same N.W. European LD matrix as used in UKBB analyses. 
 Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs are overlapping and passing quality control.
 
-Intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, as expected \cite[]{loh2018mixed}.
+Intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, which is unfortunately expected \cite[]{loh2018mixed}.
 SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
@@
 Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
-We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore requiring an even larger sample size to be able to better predict height.
+We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore an even larger sample size is required to be able to better predict height.
 Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%).
 Therefore, we estimate that height has at least 50,000 causal variants.
-These results are similar independent of whether ``allow\_jump\_sign'' is used or not, which is surprising to us.
+These results are similar irrespective of whether ``allow\_jump\_sign'' is used or not, which is surprising to us.
 We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
@@
 
-A description of the eight external (to UKBB) GWAS summary statistics used is provided in Table \ref{tab:sumstats}.
+A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these were not partly derived from UKBB individuals.
 Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
@@
 Results are presented in Figure \ref{fig:sumstats}.
-In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); it seems more useful to use this new set of variants for larger effective sample sizes.
+In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); it seems more useful to use this new set of variants for larger sample sizes.
 Results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is a bit surprising to us.
-Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance, except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
+Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
 Maximum $r^2$ are acheived at different regularization coefficients across phenotypes.
@@
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, useful for fine-mapping), and the selection-related parameter $\alpha$.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities PIP, useful for fine-mapping), and the selection-related parameter $\alpha$.
 We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
 We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
+Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
 Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
@@
 Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to define this extended set of variants, we expect it is beneficial across diverse ancestries.
-This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
+This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability, i.e.\ the difference between the family-based heritability and the SNP-based heritability.
 Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% in UKBB analyses here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
@@
 Future work could focus on correcting these issues.
-Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
+Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the new sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
 Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric between GWAS summary statistics and the LD reference used.
+Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by.
 
@@
 In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
-In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters.
+In the new LDpred2-auto, to add some sampling to these two parameters, we first sample causal variants with replacement before computing the maximum likelihood estimators. 
 This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.

### 2023-06-01 paper/paper-infer-supp.tex
save

@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{est_p_one}}
-\caption{Inferred polygenicity $p$ in simulations with continuous outcomes.
-Horizontal dashed lines represent the true simulated values.
+	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one_200K}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=200K. Horizontal dashed lines represent the true simulated values. 
 		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
-``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-\label{fig:simu_p}}
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+\label{fig:simu_h2_large}}
 \end{figure}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one}}
-\caption{Inferred $\alpha$ in simulations with continuous outcomes.
+\centerline{\includegraphics[width=0.9\textwidth]{est_p_one_20K}}
+\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=20K.
+Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{est_p_one_200K}}
+	\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=200K.
+Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p_large}}
+\end{figure}
+
+\begin{figure}[p]
+\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one_20K}}
+\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=20K.
 Horizontal dashed lines represent the true simulated values.
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-\label{fig:simu_alpha}}
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one_200K}}
+	\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=200K.
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}).
+Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha_large}}
 \end{figure}

### 2023-06-01 paper/paper-infer.tex
save

@@
 \title{Inferring disease architecture and predictive ability\\with LDpred2-auto}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Clara Albi\~nana,$^{\text{1}}$ Bogdan Pasaniuc,$^{\text{2,3,4,5}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,6,7}}$}
+
+\author{Florian Priv\'e,$^{\text{1,}*}$ Clara Albi\~nana,$^{\text{1}}$ Julyan Arbel,$^{\text{2}}$ Bogdan Pasaniuc,$^{\text{3,4,5,6}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,7,8}}$}
 
@@
 \noindent$^{\text{\sf 1}}$National Centre for Register-based Research, Aarhus University, Aarhus, Denmark. \\
-\noindent$^{\text{\sf 2}}$Bioinformatics Interdepartmental Program, University of California, Los Angeles, CA, USA. \\
-\noindent$^{\text{\sf 3}}$Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
-\noindent$^{\text{\sf 4}}$Department of Pathology and Laboratory Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
-\noindent$^{\text{\sf 5}}$Department of Computational Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
-\noindent$^{\text{\sf 6}}$Bioinformatics Research Centre, Aarhus University, Aarhus, Denmark. \\
-\noindent$^{\text{\sf 7}}$Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute, Cambridge, MA, USA. \\
+\noindent$^{\text{\sf 2}}$Univ.\ Grenoble Alpes, Inria, CNRS, Grenoble INP, LJK, Grenoble, France. \\
+\noindent$^{\text{\sf 3}}$Bioinformatics Interdepartmental Program, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 4}}$Department of Human Genetics, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 5}}$Department of Pathology and Laboratory Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 6}}$Department of Computational Medicine, David Geffen School of Medicine, University of California, Los Angeles, CA, USA. \\
+\noindent$^{\text{\sf 7}}$Bioinformatics Research Centre, Aarhus University, Aarhus, Denmark. \\
+\noindent$^{\text{\sf 8}}$Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute, Cambridge, MA, USA. \\
 \noindent$^\ast$To whom correspondence should be addressed. \\
@@
 What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero impact on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
-Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme}.
+Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
 Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
@@
 
-As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, where $r^2_\text{max} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$ ($N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants).
+As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, where $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$ ($N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants).
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
@@
 For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}.
-HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here, and now recommend to use for LDpred2 (Methods).
+HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend to use for LDpred2 when power is sufficient.
 CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting).
@@
 
+[TODO: MAKE SURE TO CITE / INCLUDE ALL NEW FIGURES]
+
 For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
@@
 the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
-This function first picks a proportion $p$ of causal variants at random, sample effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scale the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix.
-Finally, some Gaussian noise is added so that the full phenotype has a variance of 1.
-Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
+This function first picks a proportion $p$ of causal variants at random, samples effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scales the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix.
+Finally, some Gaussian noise is added so that the final phenotype has a variance of 1.
+Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
 Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the new model including a third parameter $\alpha$  (new option \texttt{use\_MLE}, Methods).
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
-Then only chains that provide top $R \beta$ (imputed marginal effect sizes) are kept.
+Then we keep only chains that provide top imputed marginal scaled effect sizes $\check{\beta} = R \bar{\beta}$, where $R$ is the correlation matrix and $\bar{\beta}$ are the PGS scaled effect sizes [TODO: MOVE TO ITS OWN METHOD SECTION, ALSO INTRODUCING NEW AND PREVIOUS].
 
-First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2}, \ref{fig:simu_p} and \ref{fig:simu_alpha}).
-Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figure \ref{fig:simu_h2}).
-When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
+Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
+When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (i.e.\ CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
 When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
 SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
-The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become unreliable or unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
+The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
+Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (``LDpred2\_newfilter'', Methods); results remain practically unchanged (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
+
 Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
-We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIP) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
+We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
 LDpred2-auto provide PIPs that are more calibrated than e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block (Figure \ref{fig:susie}).
-Finally, LDpred2-auto (and SBayesS) can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low (Figure \ref{fig:simu_r2}).
+Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low (Figure \ref{fig:simu_r2}).
+We show it works with results from SBayesS's Gibbs sampler as well (Figure \ref{fig:simu_r2}).
 
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one}}
-	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes. Horizontal dashed lines represent the true simulated values. 
+	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one_20K}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=20K. Horizontal dashed lines represent the true simulated values. 
 	For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
-	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
+	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-	The 95\% confidence interval for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+	The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
 The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
@@
 We use the same 356,409 unrelated individuals of Northwestern European ancestry as in the simulations. 
-To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using 8 covariates (Methods).
+To form the test set, we randomly select 50,000 of these, while the other 306,409 are used to run a GWAS using linear regression (with function \texttt{big\_univLinReg} from R package bigstatsr) for each of all 248 phenotypes and using eight covariates (Methods).
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a subset of the 356K used before), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a random subset of the 356K used before for which height is recorded), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
 We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
@@
 Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
-We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}), and explore multiple values for parameter \texttt{coef\_shrink} (multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix).
+We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}).
+Because of the increased mismatch between external GWAS summary statistics and the LD reference we use (compared to simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
 Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
 Results are presented in Figure \ref{fig:sumstats}.
-In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); it seems more useful to use this new set of variants for larger sample sizes.
-Results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is a bit surprising to us.
+In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); therefore it seems more useful to use this new set of variants for larger sample sizes.
 Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
-Maximum $r^2$ are acheived at different regularization coefficients across phenotypes.
+Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
+However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is a bit surprising to us.
 As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
@@
 
+[TODO: VERIFY I MENTIONED ANCESTRY FOR EG CAD]
+
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
@@
 We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
-Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
-Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
+Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
+Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$, \cite{prive2020ldpred2}).
 We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
@@
 Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 37\% larger to offer a better coverage of the genome.
-Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to define this extended set of variants, we expect it is beneficial across diverse ancestries.
+Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to construct this extended set of variants, we expect this new set to be beneficial across diverse ancestries.
 This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability, i.e.\ the difference between the family-based heritability and the SNP-based heritability.
@@
 Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
-Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\gamma}^T \boldsymbol{\gamma}$; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
+Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\gamma}^T \boldsymbol{\gamma}$ [TODO: MODIFY, THIS IS WRONG NOTATIONS]; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
 
@@
 To infer the out-of-sample predictive performance $r^2$ (and CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
-Intuitively, if prediction is perfect then $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are the same and $r^2 = h^2$; when power is very low, these two are uncorrelated and $r^2 \approx 0$.
-Note that this assumes the target sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics).
-Although we would have liked to, we do not provide any theoretical justification for this equation.
-Instead, we do check it using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
+Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
+Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
+These previous works have shown that $r^2$ can be approximated by $\big(\boldsymbol{\bar{\beta}}^T \boldsymbol{\hat{\beta}}_\text{test}\big)^2 / \big(\boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}\big)$ where $\boldsymbol{\bar{\beta}}$ and $\boldsymbol{\hat{\beta}}_\text{test}$ are respectively the predictive effects from the training set and the marginal effects from the test set (after scaling).
+Note that $E[\boldsymbol{\beta_1}] = \boldsymbol{\bar{\beta}}$, $E[\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}] = \boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}$, and, when the test sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics), $E[\boldsymbol{R} \boldsymbol{\beta_2}] \approx \boldsymbol{\hat{\beta}}_\text{test}$.
+Therefore, we propose to use $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ as a sample of $r^2$.
+This can also be computed for a specific chain by taking two samples $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ that are far enough on the same chain to remove the possible autocorrelation.
+This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
+
+In this paper, we check this new $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
 These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
@@
 Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.
+
+\subsection*{Post-filtering of chains in LDpred2-auto}
+
+Because running a Gibbs sampler with so many variants and with possible mismatches between e.g.\ the GWAS summary statistics and the LD reference used, we have always recommended to run multiple chains in LDpred2-auto, and post-filter some of them as quality control \cite[]{prive2020ldpred2}.
+We originally proposed to filter chains by keeping the ones providing PGS with the largest variances. 
+Then we tested an almost equivalent and simpler alternative \cite[]{prive2021identifying}, which is to keep only chains that provide top imputed marginal scaled effect sizes $\check{\beta} = R \bar{\beta}$, where $R$ is the correlation matrix and $\bar{\beta}$ are the PGS scaled effect sizes.
+This is the default post-filtering of LDpred2-auto chains that we use here, for both prediction and inference.
+
+Moreover, here we test two other filtering criteria in first simulations based on continuous outcomes (and call this ``LDpred2\_newfilter''). 
+First, for prediction, we test to filter on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, which is an estimate of $r^2$ (cf.\ previous Methods section).
+Second, for inference, we filter on some convergence criterion.
+The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}.
+We have found this statistic to perform badly when e.g.\ one parameter get stuck and is constant; in this case, the chain do not mix well, but we do get a perfect Rhat of 1.0.
+However, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat while not suffering from the previous issue.
+We therefore chose to use this statistic, and average the three statistics computed for $h^2$, $p$, and $\alpha$ for each chain.
+We use a threshold of 4 above which we filter out the chain, because we have found that a value of 4 for this statistic often corresponds to a value close to 1.05 for the split-Rhat.
 
@@
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-The latest version of R package bigsnpr can be installed from GitHub, and recent enough versions can be installed from CRAN [TODO: REUPLOAD].
+The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.
 

### 2023-06-02 paper/paper-infer-supp.tex
reread part 1

@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p}}
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p}}
 \end{figure}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p_large}}
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p_large}}
 \end{figure}
@@
 Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
-For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}).
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha}}
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha}}
 \end{figure}
@@
 Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
-For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}).
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha_large}}
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha_large}}
 \end{figure}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one}}
-\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes.
+\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one_20K}}
+\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=20K.
 Green segments represent $r^2$ in the test set.
 For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
-``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
+``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
 \label{fig:simu_r2}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one_200K}}
+	\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=200K.
+Green segments represent $r^2$ in the test set.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+\label{fig:simu_r2_large}}
 \end{figure}

### 2023-06-02 paper/paper-infer.tex
reread part 1

@@
 LDpred2-auto can infer the two parameters from the LDpred model, $h^2$ and $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
+The main aim of this paper is to properly validate the use of LDpred2-auto for inference of multiple parameters.
 Here, we present a new version of LDpred2-auto (as an option), which adds a third parameter $\alpha$ to its model for modeling negative selection.
@@
 Most traits and diseases in humans are heritable.
-What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero impact on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
+What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
 Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
@@
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
-These methods include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$, also called posterior inclusion probabilities PIP, used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (PIP, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
+These methods (non-exhaustively) include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (PIPs, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
 
-As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, where $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$ ($N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants).
+As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
-LDpred2 is a widely used polygenic score method that can directly build PGS using summary statistics results from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
+LDpred2 is a widely used polygenic score method that can directly build PGS using resulting summary statistics from genome-wide associations studies (GWAS), making it highly applicable \cite[]{prive2020ldpred2,pain2021evaluation,kulm2021systematic}.
 LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model.
-In LDpred2-auto, it can directly estimate these parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.
+LDpred2-auto, one variant of LDpred2, can directly estimate these two parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.
 
+The main aim of this paper is to properly validate the use of LDpred2-auto for inference of multiple parameters.
 Here we extend LDpred2-auto and show it is a reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities PIP used for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
@@
 An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
-Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
-We call `HapMap3+' this new preferred set of 1,444,196 SNPs.
+Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
+We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend to use it when power is sufficient.
 
@@
 
-[TODO: MAKE SURE TO CITE / INCLUDE ALL NEW FIGURES]
-
 For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
-We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e.
-the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
+We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e.\ the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
 This function first picks a proportion $p$ of causal variants at random, samples effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scales the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix.
@@
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
-Then we keep only chains that provide top imputed marginal scaled effect sizes $\check{\beta} = R \bar{\beta}$, where $R$ is the correlation matrix and $\bar{\beta}$ are the PGS scaled effect sizes [TODO: MOVE TO ITS OWN METHOD SECTION, ALSO INTRODUCING NEW AND PREVIOUS].
+Then some of these chains are filtered out for quality control (Methods).
 
@@
 When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (i.e.\ CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+LDpred2\_nojump looks very reliable for estimating $h^2$.
 When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
@@
 We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
-LDpred2-auto provide PIPs that are more calibrated than e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block (Figure \ref{fig:susie}).
-Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low (Figure \ref{fig:simu_r2}).
-We show it works with results from SBayesS's Gibbs sampler as well (Figure \ref{fig:simu_r2}).
+LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure \ref{fig:susie}).
+Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
 
@@
 
-To investigate whether estimates from LDpred2-auto are robust to misspecifications, we test using two alternative LD references (Methods).
+To investigate whether estimates from LDpred2-auto are robust to some misspecifications, we test using two alternative LD references (Methods).
 Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
 When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
-Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can get chains to converge (and therefore non-zero predictive performance) for the 7 (3 + 4) phenotypes mentioned before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
+Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can always get some chains not to diverge (and therefore get non-zero predictive performance) for the seven (three and four) phenotypes mentioned before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
 
@@
 Indeed, 49 of the phenotypes used here seem log-normally distributed or heavy-tailed (when visualizing their histogram); we therefore log-transform them.
-However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference from LDpred2-auto.
+However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference and prediction from LDpred2-auto.
 We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
@@
 We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
-For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one block only.
+For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
 We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure \ref{fig:protein_local_h2}).
@@
 Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
-To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
-To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
+To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a small negative correlation of -5.5\% with the probabilities of being causal.
+To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a small correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
 
@@
 
-Here we run three LDpred2-auto models for height, one based on 100K UKBB individuals (as a random subset of the 356K used before for which height is recorded), one from the same 305K UKBB individuals used before, and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries across 1,373,020 variants \cite[]{yengo2022saturated}.
+Here we run three LDpred2-auto models for height, one from the same 305K training UKBB individuals used before (with available height, out of 306K), one based on 100K UKBB individuals (as a random subset of the previous 305K), and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries \cite[]{yengo2022saturated}.
 We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
@@
 For this set of external GWAS summary statistics, we therefore use the same N.W. European LD matrix as used in UKBB analyses. 
-Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs are overlapping and passing quality control.
+Note that we use the HapMap3+ set of 1,444,196 SNPs here, however, for the GWAS meta-analysis, only 1,013,499 SNPs (out of 1,373,020) are overlapping with HapMap3+ and passing quality control.
 
-Intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M, which is unfortunately expected \cite[]{loh2018mixed}.
+As expected \cite[]{loh2018mixed}, intercepts from LD Score regression are increasing with sample size: 1.02 (SE: 0.008) with N=100K, 1.11 (0.015) with N=305K, and 2.31 (0.068) with N=1.6M.
 SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
@@
 Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
-We emphasize that, even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore an even larger sample size is required to be able to better predict height.
+Even though there are 1.6M individuals in the meta-analysis, the predictive performance corresponds to around 87\% of the SNP heritability only, therefore an even larger sample size is required to be able to better predict height.
 Polygenicity estimates from LDpred2-auto are increasing with sample size with 1.2\% [1.0, 1.5], 2.3\% [2.0, 2.5], and 5.9\% [5.6, 6.3], consistent with results of simulations with a large polygenicity (p=10\%).
@@
 
-A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these were not partly derived from UKBB individuals.
+A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these do not include UKBB individuals.
 Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
 We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}).
-Because of the increased mismatch between external GWAS summary statistics and the LD reference we use (compared to simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
+Because of the increased mismatch between external GWAS summary statistics and the LD reference we use here (compared to in simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
 Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
@@
 Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
-However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is a bit surprising to us.
+However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is surprising to us.
 As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
 This is also the case for estimates of $p$ when using \texttt{use\_MLE = TRUE}.
-Estimates of $\alpha$ are largely constant but can become very small in the case of robustness issues when using e.g.\ \texttt{use\_MLE = TRUE} and almost no regularization on the LD matrix (i.e.\ \texttt{coef\_shrink} close to 1).
+Estimates of $\alpha$ are largely constant but can become very small (capped at -1.5) in the case of robustness issues when using e.g.\ \texttt{use\_MLE = TRUE} and almost no regularization on the LD matrix (i.e.\ \texttt{coef\_shrink} close to 1).
 
@@
 	\end{tabular}
-	\caption{Summary of external GWAS summary statistics used.  \label{tab:sumstats}}
+	\caption{Summary of external GWAS summary statistics used.
+These do not include UKBB individuals.
+Note that some of them may contain a substantial amount of non-European genetic ancestry (e.g.\ 20\%).
+\label{tab:sumstats}}
 \end{table}
@@
 Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
-Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\gamma}^T \boldsymbol{\gamma}$ [TODO: MODIFY, THIS IS WRONG NOTATIONS]; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
+Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
 

### 2023-06-02 paper/paper-infer.tex
reread rest

@@
 LDpred2-auto can infer the two parameters from the LDpred model, $h^2$ and $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
-The main aim of this paper is to properly validate the use of LDpred2-auto for inference of multiple parameters.
+The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
 Here, we present a new version of LDpred2-auto (as an option), which adds a third parameter $\alpha$ to its model for modeling negative selection.
@@
 
-The main aim of this paper is to properly validate the use of LDpred2-auto for inference of multiple parameters.
+The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
 Here we extend LDpred2-auto and show it is a reliable method for estimating $h^2$ (global and local), $p$ (also per-variant probabilities PIP used for fine-mapping purposes), and $\alpha$ (by extending its model to also include this third parameter).
@@
 For simulations, we use the UK Biobank imputed data \cite[]{bycroft2018uk}. We use 356,409 individuals of Northwestern European ancestry and 322,805 SNPs across seven chromosomes (Methods).
-We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP-heritability $h^2$, the polygenicity $p$ (i.e.\ the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
+We first simulate continuous phenotypes using function \texttt{snp\_simuPheno} from R package bigsnpr \cite[]{prive2017efficient}, varying three parameters: the SNP heritability $h^2$, the polygenicity $p$ (i.e.\ the proportion of causal variants), and the parameter $\alpha$ in equation \eqref{eq:new_model} that controls the relationship between minor allele frequencies and expected effect sizes.
 This function first picks a proportion $p$ of causal variants at random, samples effect sizes $\gamma$ using the variance component parameterized by $\alpha$ and then scales the effect sizes so that the genetic component $G \gamma$ has a variance $h^2$, where $G$ is the genotype matrix.
@@
 
-[TODO: VERIFY I MENTIONED ANCESTRY FOR EG CAD]
-
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP-heritability (both genome-wide, local, and for annotations), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities PIP, useful for fine-mapping), and the selection-related parameter $\alpha$.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
 We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
@@
 Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 37\% larger to offer a better coverage of the genome.
-Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to construct this extended set of variants, we expect this new set to be beneficial across diverse ancestries.
-This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability, i.e.\ the difference between the family-based heritability and the SNP-based heritability.
+Since we used individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
+Increasing the genome coverage enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability, i.e.\ the difference between the family-based heritability and the SNP-based heritability.
 Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% in UKBB analyses here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
@@
 When using external GWAS summary statistics, using HapMap3+ instead of HapMap3 variants was more beneficial for larger sample sizes.
+Our intuition and conclusion is that using more variants is beneficial when power is sufficient; however, when power is low (e.g.\ small $N$, small $h^2$, and/or large $p$), it may be actually be detrimental.
 
@@
 Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the new sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
-Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric between GWAS summary statistics and the LD reference used.
+Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
 Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by.
@@
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
-These individuals are unrelated and have passed some quality control including removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
+These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
@@
 The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
-As covariates, for the subset of individuals previously defined, we first recompute PCs using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
+As covariates, we first recompute PCs for the homogeneous subset of individuals previously defined using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
 We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).
@@
 We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
-We design two other LD matrices: one using a smaller subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
+We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
 We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
@@
 To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
+There are around 11.5M such variants.
 Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
@@
 Note that we only allow non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs, of which we compute the correlation between variants (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
-Since we use individuals of diverse ancestries for computing pairwise variant correlations that are used to define this extended set of variants, we expect it is beneficial across diverse ancestries.
+This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
 
@@
 where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\boldsymbol{\gamma}$ the effect sizes on the allele scale, $\boldsymbol{S}$ the standard deviations of the genotypes, and $\boldsymbol{\beta}$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
-In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters.
+In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters (as in LDpred2-grid). 
 This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly without requiring a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
 $h^2$ is estimated by $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
-As for $p$, it was sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
-Note a small change, we now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more variability in the sampling in order to account for a reduced effective number of independent variants.
+As for $p$, it is sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
+Note a small change, we now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more variability in the sampling in order to account for a reduced effective number of (independent) variants.
 
@@
 This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
-Note that we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}).
+Note that we still estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}). 
 Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
@@
 
-To infer the out-of-sample predictive performance $r^2$ (and CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
+To infer the out-of-sample predictive performance $r^2$ (and its CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
 Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
@@
 
-Moreover, here we test two other filtering criteria in first simulations based on continuous outcomes (and call this ``LDpred2\_newfilter''). 
-First, for prediction, we test to filter on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, which is an estimate of $r^2$ (cf.\ previous Methods section).
-Second, for inference, we filter on some convergence criterion.
-The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}.
-We have found this statistic to perform badly when e.g.\ one parameter get stuck and is constant; in this case, the chain do not mix well, but we do get a perfect Rhat of 1.0.
-However, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat while not suffering from the previous issue.
+Moreover, here we test two other filtering criteria in the first simulations based on continuous outcomes (and call this ``LDpred2\_newfilter''). 
+First, for prediction, we test filtering on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ within each chain, which is an estimate of $r^2$ (cf.\ previous Methods section).
+Second, for inference, we filter on some convergence criterion. The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}. 
+However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain do not mix well, but we do get a perfect Rhat of 1.0. 
+Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat while not suffering from the previous issue.
 We therefore chose to use this statistic, and average the three statistics computed for $h^2$, $p$, and $\alpha$ for each chain.

### 2023-06-02 code/simu-methods.R

@@ -1,6 +1,6 @@
 ################################################################################
 
-list2 <- function(...) setNames(list(...), sapply(substitute(list(...))[-1], deparse))
+list2 <- function(...) rlang::dots_list(..., .named = TRUE)
 
 r2_boot <- function(x, y) {
   all_cor <- replicate(1000, {
@@ -9,6 +9,26 @@ r2_boot <- function(x, y) {
   })
   cor <- c(cor(x, y), quantile(all_cor, c(0.025, 0.975)))
   sign(cor) * cor^2
+}
+
+conv_stat <- function(multi_auto, which, Niter) {
+  # Two-Sample Cramer-von Mises Statistic -- Eq (6) of 10.1214/aoms/1177704477
+  sapply(multi_auto, function(auto) {
+    x <- tail(auto[[paste0("path_", which, "_est")]], Niter)
+    if (anyNA(x)) return(NA_real_)
+    r <- rank(x, na.last = "keep", ties.method = "average")
+    N <- floor(length(x) / 2)
+    M <- length(x) - N
+    t <- (N / M * crossprod(sort(r[1:N]) - 1:N * (N + M) / N) +
+            M / N * crossprod(sort(r[1:M + N]) - 1:M * (N + M) / M)) / (N + M)^2
+    t[[1]]
+  })
+}
+
+conv_stat_all <- function(multi_auto, Niter) {
+  rowMeans(
+    sapply(c("h2", "p", "alpha"), conv_stat, multi_auto = multi_auto, Niter = Niter),
+    na.rm = TRUE)
 }
 
 ################################################################################
@@ -74,6 +94,94 @@ run_ldpred2 <- function(jump_sign, use_mle, coef_shrink) {
         all_local_h2, all_true_h2)
 }
 
+run_ldpred2_new <- function(jump_sign, use_mle, coef_shrink) {
+
+  res0 <- list2(jump_sign, use_mle)
+
+  multi_auto <- snp_ldpred2_auto(corr, df_beta, h2_init = pmax(ldsc[["h2"]], 0.001),
+                                 vec_p_init = seq_log(1e-4, 0.2, length.out = 50),
+                                 burn_in = 500, num_iter = 500, report_step = 20,
+                                 use_MLE = use_mle,
+                                 allow_jump_sign = jump_sign,
[…]

### 2023-06-02 code/simu.R

@@ -113,6 +113,9 @@ grid$res <- furrr::future_pmap(grid[1:5], function(p, h2, alpha, N, num) {
   # Run methods
   source('code/simu-methods.R', local = TRUE)
   list(
+    LDpred2_newfilter = runonce::save_run(
+      run_ldpred2_new(jump_sign = FALSE, use_mle = TRUE, coef_shrink = 0.95),
+      file = paste0("results_simu/LDpred2_newfilter_", params, ".rds")),
     LDpred2_noMLE = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = FALSE, coef_shrink = 0.95),
       file = paste0("results_simu/LDpred2_noMLE_", params, ".rds")),
@@ -140,20 +143,23 @@ grid2 <- grid %>%
   tidyr::unnest_wider("res") %>%
   tidyr::pivot_longer(-(1:5), names_to = "Method") %>%
   tidyr::unnest_wider("value") %>%
-  mutate(n_keep = purrr::map_dbl(all_h2, ~ `if`(is.null(.), 0, NCOL(.))),
-         Method = factor(Method, levels = Method[1:5]),
+  mutate(r2 = lapply(r2, unname),
+         n_keep_infer = purrr::map_dbl(all_h2, ~ `if`(is.null(.), 0, NCOL(.))),
+         n_keep_pred  = purrr::map_dbl(all_r2, ~ `if`(is.null(.), 0, NCOL(.) / 500 * 20)),
+         Method = factor(Method, levels = Method[1:6]),
          p = as.factor(p)) %>%
-  relocate(n_keep, .after = Method) %>%
+  relocate(n_keep_infer, n_keep_pred, .after = Method) %>%
   print(n = 100)
 
 
 ggplot(filter(grid2, grepl("LDpred2", Method))) +
-  geom_histogram(aes(n_keep, fill = Method), color = "black", size = 0.1,
+  geom_histogram(aes(n_keep_infer, fill = Method), color = "black", size = 0.1,
                  breaks = seq(0, 50, by = 2)) +
   bigstatsr::theme_bigstatsr(0.8) +
   facet_grid(p ~ N + h2, labeller = label_both) +
   labs(x = "Number of chains kepts") +
-  scale_y_continuous(breaks = 0:10 * 2)
+  scale_y_continuous(breaks = 0:10 * 2) +
+  theme(legend.position = "top")
 
 
[…]

### 2023-06-08 paper/paper-infer-supp.tex
small modifs

@@
 The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 \label{fig:simu_h2_large}}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p}}
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2 models represent the number of chains kept (out of 50).
+\label{fig:simu_p}}
 \end{figure}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_p_large}}
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2 models represent the number of chains kept (out of 50).
+\label{fig:simu_p_large}}
 \end{figure}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha}}
+The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2 models represent the number of chains kept (out of 50).
+\label{fig:simu_alpha}}
 \end{figure}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).\label{fig:simu_alpha_large}}
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2 models represent the number of chains kept (out of 50).
+\label{fig:simu_alpha_large}}
 \end{figure}
@@
 \centerline{\includegraphics[width=\textwidth]{postp_calib}}
-\caption{Calibration of per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities) in simulations with continuous outcomes.
+\caption{LDpred2-auto calibration of per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities) in simulations with continuous outcomes.
 These probabilities are binned on a log-scale, and mean in each bin is compared to the proportion of simulated causal variants in the bin (with 95\% CI).
@@
 Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented.
+Colors represent the number of chains kept (out of 50).
 \label{fig:simu_postp_calib}}
@@
 \centerline{\includegraphics[width=\textwidth]{local_h2_calib}}
-\caption{Calibration of per-block heritability estimates in simulations with continuous outcomes. 
+\caption{LDpred2-auto calibration of per-block heritability estimates in simulations with continuous outcomes. 
 Local $h^2$ estimate for each block is compared to the true simulated local heritability in the block.
-Green dashed lines represent the 1:1 line. Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented. \label{fig:simu_h2_calib}}
+Green dashed lines represent the 1:1 line.
+Only results for $\alpha = -0.5$ and \texttt{allow\_jump\_sign = FALSE} are represented.
+Colors represent the number of chains kept (out of 50).
+\label{fig:simu_h2_calib}}
 \end{figure}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 \label{fig:simu_r2}}
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 \label{fig:simu_r2_large}}
@@
 The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 {\NEFF} All $h^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}.
@@
 	\centerline{\includegraphics[width=0.9\textwidth]{est_binary_p_one}}
-	\caption{Inferred polygenicity $p$ in simulations with binary outcomes. Horizontal dashed lines represent the true simulated values. {\NEFF} \label{fig:simu_p_bin}}
+	\caption{Inferred polygenicity $p$ in simulations with binary outcomes.
+Horizontal dashed lines represent the true simulated values.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
+{\NEFF} \label{fig:simu_p_bin}}
 \end{figure}
@@
 Horizontal dotted lines represent boundaries imposed on the estimates.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 {\NEFF} \label{fig:simu_alpha_bin}}
@@
 Green segments represent $r^2$ in the test set.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 {\NEFF} All $r^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}; the ones from the test set are transformed with K and K.
@@
 ``use\_mle: TRUE'' corresponds to using the new 3-parameter model and sampling scheme (Methods).
+Colors represent the number of chains kept for LDpred2-auto (out of 50).
 \label{fig:ukbb_h2}}
@@
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+Colors represent the number of chains kept for LDpred2-auto (out of 50).
 \label{fig:ukbb_h2_p}}
@@
 The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the estimates from the iterations (after burn-in) of the chains kept.
+Colors represent the number of chains kept for LDpred2-auto (out of 50).
 We only show phenotypes for which there are more than 25 chains kept, because simulations have shown that $\alpha$ estimates are unreliable when a small number of chains is kept (Figure \ref{fig:simu_alpha}).

### 2023-06-08 paper/paper-infer.tex
small modifs

@@
 LDpred2 is a widely used Bayesian method for building polygenic scores (PGS).
-LDpred2-auto can infer the two parameters from the LDpred model, $h^2$ and $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
+LDpred2-auto can infer the two parameters from the LDpred model, the SNP heritability $h^2$ and polygenicity $p$, so that it does not require an additional validation dataset to choose best-performing parameters.
 The main aim of this paper is to properly validate the use of LDpred2-auto for inferring multiple genetic parameters.
-Here, we present a new version of LDpred2-auto (as an option), which adds a third parameter $\alpha$ to its model for modeling negative selection.
+Here, we present a new version of LDpred2-auto that adds an optional third parameter $\alpha$ to its model, for modeling negative selection.
 We then validate the inference of these three parameters (or two, when using the previous model).
@@
 
-As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
+As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
@@
 So, on top of providing competitive PGS, LDpred2-auto can also provide all these estimates of genetic architecture. 
-Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
+Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without requiring an independent test set.
 An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.85\textwidth]{overview}}
+	\centerline{\includegraphics[width=0.9\textwidth]{overview}}
 	\caption{Overview of what LDpred2-auto can now provide.
@@
 The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+Colors for LDpred2 models represent the number of chains kept (out of 50).
 \label{fig:simu_h2}}
@@
 Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
-The main difference is that the $h^2$ and $r^2$ estimates need to be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.
+The main difference is that the $h^2$ and $r^2$ estimates must be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.
 
@@
 The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+Colors represent the number of chains kept (out of 50).
 ``F\_height'' and ``M\_height'' use females and males only, respectively (in both GWAS and test sets).
@@
 
-Then, we investigate the effect of disabling ``allow\_jump\_sign'' (for extra robustness as shown in \cite{prive2021identifying}) on the estimates from LDpred2-auto.
+Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} (for extra robustness as shown in \cite{prive2021identifying}) on the estimates from LDpred2-auto.
 Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ normal sampling, Figure \ref{fig:jumpsign}).
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_plus}}
+	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_compare_hm3_plus}}
 	\caption{LDpred2-auto estimates for UKBB phenotypes using either the HapMap3 or HapMap3+ sets of variants.
@@
 Therefore, we estimate that height has at least 50,000 causal variants.
-These results are similar irrespective of whether ``allow\_jump\_sign'' is used or not, which is surprising to us.
+These results are similar irrespective of whether \texttt{allow\_jump\_sign} is used or not, which is surprising to us.
 We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
@@
 		\hline
+		\hline
 		Asthma & \citet{demenais2018multiancestry} & ~~67,341.2 & ~~2,001,280 & ~~~946,092 \\
@@
 These do not include UKBB individuals.
-Note that some of them may contain a substantial amount of non-European genetic ancestry (e.g.\ 20\%).
+Note that some of them may contain a substantial amount of non-European genetic ancestry (e.g.\ >20\%) for CAD.
 \label{tab:sumstats}}
@@
 
-Our proposed method has limitations.
-First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
+Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
+Therefore, we do recommend to use the 3-parameter model (with $\alpha$), but only when power is sufficient.
 However, estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
@@
 The HapMap3 variants generally provide a good coverage of the whole genome.
-We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ both 'A' and 'T' or 'C' and 'G'), and which includes SNPs only (e.g.\ no indel).
+We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
 Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
@@
 Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
-This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite[]{zeng2021widespread}. 
-In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
+This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is denoted by $S$ instead \cite[]{zeng2021widespread}. 
+In SBayesS, $\alpha$ and $\sigma_\beta^2$ are estimated by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
 In the new LDpred2-auto, to add some sampling to these two parameters, we first sample causal variants with replacement before computing the maximum likelihood estimators. 
@@
 Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
-Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
+Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the vector of scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
 
@@
 Second, for inference, we filter on some convergence criterion. The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}. 
-However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain do not mix well, but we do get a perfect Rhat of 1.0. 
+However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain does not mix well, but a perfect Rhat of 1.0 is obtained. 
 Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat while not suffering from the previous issue.

### 2023-06-08 paper/paper-infer.tex
finished

@@
 The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_h2}}
@@
 
-Because running a Gibbs sampler with so many variants and with possible mismatches between e.g.\ the GWAS summary statistics and the LD reference used, we have always recommended to run multiple chains in LDpred2-auto, and post-filter some of them as quality control \cite[]{prive2020ldpred2}.
+Because a Gibbs sampler can be unstable, with so many variants and with possible mismatches between e.g.\ the GWAS summary statistics and the LD reference used, we have always recommended to run multiple chains in LDpred2-auto, and post-filter some of them as quality control \cite[]{prive2020ldpred2}.
 We originally proposed to filter chains by keeping the ones providing PGS with the largest variances. 
@@
 However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain does not mix well, but a perfect Rhat of 1.0 is obtained. 
-Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat while not suffering from the previous issue.
+Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat statistic while not suffering from the previous issue.
 We therefore chose to use this statistic, and average the three statistics computed for $h^2$, $p$, and $\alpha$ for each chain.

### 2023-07-12 paper/paper-infer.tex
start answering reviews

@@
 
-Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} (for extra robustness as shown in \cite{prive2021identifying}) on the estimates from LDpred2-auto.
+Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
 Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ normal sampling, Figure \ref{fig:jumpsign}).
@@
 As for $\alpha$ estimates from LDpred2-auto, they remain consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
-Finally, we compute per-annotation heritability estimates to investigate functional enrichment.
+Finally, we compute per-annotation heritability estimates from LDpred2-auto results to investigate functional enrichment.
 We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
-Heritability enrichments are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
+Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
 
@@
 We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
-However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression (S-LDSC, \cite{finucane2018heritability}).
+However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC, \cite{finucane2018heritability}).
 For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
@@
 Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by.
+LDpred2-auto can also totally miss some causal variants when they are poorly tagged by the set of variants used.
+Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
 
@@
 Note that we still estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}). 
-Note that this $h^2$ estimation can be restricted to e.g.\ variants from a single LD block to get local heritability estimates.
+Note that, to get local heritability estimates (e.g.\ for a single LD block), this $h^2$ estimation ($\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$) is simply restricted to the variants from this LD block.
 Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the vector of scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.

### 2023-07-18 paper/paper-infer.tex
add tagging results

@@
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
-We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend to use it when power is sufficient.
+We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend to use it when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
 
@@
 For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}.
-HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend to use for LDpred2 when power is sufficient.
+HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend to use for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
 CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting).
@@
 Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 37\% larger to offer a better coverage of the genome.
-Since we used individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
+Since we used individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, this new set of variants is beneficial across diverse ancestries (Methods).
 Increasing the genome coverage enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability, i.e.\ the difference between the family-based heritability and the SNP-based heritability.
@@
 Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
+Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (resp.\ 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).
 

### 2023-07-21 paper/paper-infer.tex
autocor
clarify

@@
 Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
-When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (i.e.\ CIs are small, and do not contain the true parameter value), and LDpred2\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
-LDpred2\_nojump looks very reliable for estimating $h^2$.
+When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (i.e.\ CIs are small, and do not contain the true parameter value), and LDpred2-auto\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+LDpred2-auto\_nojump looks very reliable for estimating $h^2$.
 When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
@@
 The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
-Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (``LDpred2\_newfilter'', Methods); results remain practically unchanged (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
+Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (``LDpred2-auto\_newfilter'', Methods); results remain practically unchanged (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
 
@@
 Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}).
-As for the inferred predictive performance $r^2$, they are highly consistent with the ones derived from the test set; only for standing height are they overestimated (Figure \ref{fig:ukbb_r2}).
+As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figure \ref{fig:ukbb_r2}).
 Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
@@
 To investigate whether estimates from LDpred2-auto are robust to some misspecifications, we test using two alternative LD references (Methods).
-Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$), while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
+Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$) with LDpred2-auto, while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
 When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
@@
 SNP heritability estimates are 64.6\% (SE: 2.7), 59.7\% (2.2), and 39.2\% (1.7) with LD Score regression, respectively, and 60.2\% [95\% CI: 57.2, 63.2], 63.2\% [62.0, 64.4], and 54.2\% [53.9, 54.5] with LDpred2-auto.
-As expected, predictive performance $r^2$ (estimated from the Gibbs sampler) are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
+As expected, estimated predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto) are increasing with sample size, with 29.6\% [28.7, 30.5], 42.7\% [42.2, 43.1], and 47.0\% [46.8, 47.1], respectively.
 Note that these $r^2$ estimates are probably overestimated by the same margin as the (SNP heritability) $h^2$ estimates, and correspond to \textasciitilde49\%, \textasciitilde67.5\%, and \textasciitilde87\% of $h^2$, respectively.
@@
 LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
-Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
+Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
 We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
@@
 Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
-Therefore, we do recommend to use the 3-parameter model (with $\alpha$), but only when power is sufficient.
-However, estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
+Therefore, we recommend to use the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient.
+We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates.
+When still obtaining a large $p$ estimate ($>1\%$) with this option, we recommend rerunning LDpred2-auto without this option to get a less conservative estimate (cf.\ simulations).
+However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
 We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
@@
 This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
+Note that \cite{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
+Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.
 
@@
 
-Moreover, here we test two other filtering criteria in the first simulations based on continuous outcomes (and call this ``LDpred2\_newfilter''). 
+Moreover, here we test two other filtering criteria in the first simulations based on continuous outcomes (and call this ``LDpred2-auto\_newfilter''). 
 First, for prediction, we test filtering on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ within each chain, which is an estimate of $r^2$ (cf.\ previous Methods section).

### 2023-08-31 paper/paper-infer.tex
answer some reviews

@@
 The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
+In this model, the expected variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
@@
 However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
-We think this is likely due to assortative mating \cite[]{border2022assortative,yengo2022saturated}. 
+We think this is likely due to assortative mating, which causes GWAS effects to be inflated \cite[]{border2022assortative,herzig2023model}, which causes our estimates to be inflated as well. 
 Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.

### 2023-08-31 code/validate-hm3-plus.R
validate HM3+

@@ -1,0 +1,97 @@
+library(bigsnpr)
+library(dplyr)
+
+df0 <- bigreadr::fread2(
+  "UKBB/ukb41181.csv",
+  select    = c("eid", "22020-0.0",   paste0("22009-0.", 1:16)),
+  col.names = c("eid", "used_in_pca", paste0("PC", 1:16))
+)
+PC_UKBB <- select(df0, PC1:PC16) %>%
+  as.matrix()
+
+POPS <- c("United Kingdom", "Iran", "India", "China", "Nigeria")
+all_centers <- read.csv(
+  "https://raw.githubusercontent.com/privefl/UKBB-PGS/main/pop_centers.csv",
+  stringsAsFactors = FALSE) %>%
+  filter(Ancestry %in% POPS)
+all_sq_dist <- apply(all_centers[-1], 1, function(one_center) {
+  rowSums(sweep(PC_UKBB, 2, one_center, '-')^2)
+})
+thr_sq_dist <- max(dist(all_centers[-1])^2) * 0.002 / 0.16
+group <- apply(all_sq_dist, 1, function(x) {
+  ind <- which.min(x)
+  if (isTRUE(x[ind] < thr_sq_dist)) all_centers$Ancestry[ind] else NA
+})
+group[-which(df0$used_in_pca == 1)] <- NA  # unrelated and QCed
+table(group, exclude = NULL)
+# China      India     Iran    Nigeria   United Kingdom       <NA>
+#  1769       6019     1246       5017           374515     113939
+
+
+snpid_hm3_plus <- with(
+  snp_attach("data/UKBB_hm3_plus.rds")$map,
+  paste(chromosome, physical.pos, allele1, allele2, sep = "_"))
+
+eid_bgen <- bigreadr::fread2("UKBB/ukb58024_imp_chr1_v3_s487296.sample")$ID_2[-1]
+
+chr <- 22
+NCORES <- nb_cores()
+
+all_tagging <- lapply(POPS, function(pop) {
+
+  print(pop)
+
+  # pop <- POPS[1]
+
+  res_file <- paste0("tmp-data/tagging_chr", chr, "_", substr(pop, 1, 3), ".rds")
+
+  tagging <- runonce::save_run(file = res_file, {
+
+    ind <- which(group == pop)
+    if (length(ind) > 10e3) ind <- sample(ind, 10e3)
+
+
[…]

### 2023-08-31 paper/paper-infer-supp.tex
finished
validate CIs

@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one_200K}}
+	\centerline{\includegraphics[width=0.8\textwidth]{est_h2_one_20K}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=20K. Horizontal dashed lines represent the true simulated values. 
+		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu_h2}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{est_h2_one_200K}}
 	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=200K. Horizontal dashed lines represent the true simulated values. 
@@
 The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_h2_large}}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{est_p_one_20K}}
+\centerline{\includegraphics[width=0.75\textwidth]{est_p_one_20K}}
 \caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=20K.
@@
 The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_p}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{est_p_one_200K}}
+	\centerline{\includegraphics[width=0.75\textwidth]{est_p_one_200K}}
 	\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=200K.
@@
 		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_p_large}}
@@
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one_20K}}
+\centerline{\includegraphics[width=0.78\textwidth]{est_alpha_one_20K}}
 \caption{Inferred $\alpha$ in simulations with continuous outcomes and N=20K.
@@
 The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_alpha}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.8\textwidth]{est_alpha_one_200K}}
+	\centerline{\includegraphics[width=0.78\textwidth]{est_alpha_one_200K}}
 	\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=200K.
@@
 		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_alpha_large}}
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{est_h2_ten_20K}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=20K, in 10 different runs (``num''). Horizontal dashed lines represent the true simulated values. 
+		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_h2}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.8\textwidth]{est_h2_ten_200K}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=200K, in 10 different runs (``num''). Horizontal dashed lines represent the true simulated values. 
+		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_h2_large}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.82\textwidth]{est_p_ten_20K}}
+	\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=20K, in 10 different runs (``num'').
+Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_p}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.82\textwidth]{est_p_ten_200K}}
+	\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=200K, in 10 different runs (``num'').
+Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_p_large}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.73\textwidth]{est_alpha_ten_20K}}
+	\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=20K, in 10 different runs (``num'').
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_alpha}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.73\textwidth]{est_alpha_ten_200K}}
+	\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=200K, in 10 different runs (``num'').
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_alpha_large}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}[p]
-\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one_20K}}
+\centerline{\includegraphics[width=0.78\textwidth]{est_r2_one_20K}}
 \caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=20K.
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_r2}}
@@
 \begin{figure}[p]
-	\centerline{\includegraphics[width=0.9\textwidth]{est_r2_one_200K}}
+	\centerline{\includegraphics[width=0.78\textwidth]{est_r2_one_200K}}
 	\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=200K.
@@
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 \label{fig:simu_r2_large}}
@@
 
-%[ ADD THE SAME, BUT AVERAGING OVER 10 SIMUS -> CI OF THE MEAN ]
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.88\textwidth]{est_r2_ten_20K}}
+	\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=20K, in 10 different runs (``num'').
+Green segments represent $r^2$ in the test set.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_r2}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=0.88\textwidth]{est_r2_ten_200K}}
+	\caption{Inferred predictive performance $r^2$ in simulations with continuous outcomes and N=200K, in 10 different runs (``num'').
+Green segments represent $r^2$ in the test set.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu10_r2_large}}
+\end{figure}
 
@@
 The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 {\NEFF} All $h^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}.
@@
 Horizontal dashed lines represent the true simulated values.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 {\NEFF} \label{fig:simu_p_bin}}
@@
 Horizontal dotted lines represent boundaries imposed on the estimates.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 {\NEFF} \label{fig:simu_alpha_bin}}
@@
 Green segments represent $r^2$ in the test set.
-Colors for LDpred2 models represent the number of chains kept (out of 50).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
 {\NEFF} All $r^2$ estimates are transformed to the liability scale with K\_pop=K and K\_GWAS either K (the simulated prevalence) or 0.5 (when using Neff) using function \texttt{coef\_to\_liab} from R package bigsnpr \cite{lee2012better,prive2017efficient}; the ones from the test set are transformed with K and K.

### 2023-08-31 paper/paper-infer.tex
export results
validate CIs

@@
 
-First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
-Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
-When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of the heritability (i.e.\ CIs are small, and do not contain the true parameter value), and LDpred2-auto\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2_main}--\ref{fig:simu_alpha_main} and \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
+Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2_main}, \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
+When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of $h^2$ (i.e.\ CIs are small, and do not contain the true parameter value, Figure \ref{fig:simu10_h2}), and LDpred2-auto\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
 LDpred2-auto\_nojump looks very reliable for estimating $h^2$.
@@
 SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
-The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto, while estimates from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figure \ref{fig:simu_alpha}).
-Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (``LDpred2-auto\_newfilter'', Methods); results remain practically unchanged (Figures \ref{fig:simu_h2} and \ref{fig:simu_h2_large}--\ref{fig:simu_alpha_large}).
+The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
+Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures \ref{fig:simu10_p} and \ref{fig:simu10_alpha}).
+Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures \ref{fig:simu10_h2}--\ref{fig:simu10_alpha_large}), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
+Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_newfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
 
-Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
-We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
-LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure \ref{fig:susie}).
-Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
-
-\begin{figure}[p]
-	\centerline{\includegraphics[width=0.85\textwidth]{est_h2_one_20K}}
-	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=20K. Horizontal dashed lines represent the true simulated values. 
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{est_h2_one_main}}
+	\caption{Inferred SNP heritability $h^2$ in simulations with continuous outcomes and N=20K.
+More results can be found in the Supplementary Materials.
+Horizontal dashed lines represent the true simulated values.
 		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
@@
 Colors for LDpred2-auto models represent the number of chains kept (out of 50).
-\label{fig:simu_h2}}
+\label{fig:simu_h2_main}}
 \end{figure}
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{est_p_one_main}}
+	\caption{Inferred polygenicity $p$ in simulations with continuous outcomes and N=20K.
+More results can be found in the Supplementary Materials.
+Horizontal dashed lines represent the true simulated values.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $p$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu_p_main}}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centerline{\includegraphics[width=\textwidth]{est_alpha_one_main}}
+	\caption{Inferred $\alpha$ in simulations with continuous outcomes and N=20K.
+More results can be found in the Supplementary Materials.
+Horizontal dashed lines represent the true simulated values.
+Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
+Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
+		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $\alpha$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
+Colors for LDpred2-auto models represent the number of chains kept (out of 50).
+\label{fig:simu_alpha_main}}
+\end{figure}
+
+Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
+LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure \ref{fig:susie}).
+Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
+CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when $h^2$ and $N$ are small (Figures \ref{fig:simu10_r2} and \ref{fig:simu10_r2_large}).
 
@@
 
-\begin{table}[htb]
+\begin{table}[!htb]
 	\centering
@@
 Descriptions of UK Biobank phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
+Simulation and UKBB results are available at \url{https://github.com/privefl/paper-infer/tree/main/results}.
 LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper are available at \url{https://doi.org/10.6084/m9.figshare.21305061}.
-All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/master/code}.
+All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/main/code}.
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
@@
 
-\clearpage
-%\vspace*{3em}
+%\clearpage
+\vspace*{3em}
 

### 2023-08-31 code/simu.R

@@ -1,4 +1,5 @@
 library(bigsnpr)
+library(dplyr)
 
 #### Prepare LD matrix ####
 
@@ -65,8 +66,9 @@ grid <- tidyr::expand_grid(
   h2    = c(1, 3, 10, 30) / 100,
   alpha = c(0, -0.5, -1),
   N     = c(20, 200) * 1000,
-  num   = 1
-)
+  num   = 1:10
+) %>%
+  filter(num == 1 | alpha == -0.5)
 
 NCORES <- 13
 library(future.batchtools)
@@ -133,9 +135,43 @@ grid$res <- furrr::future_pmap(grid[1:5], function(p, h2, alpha, N, num) {
 })
 
 
+library(dplyr)
+
+grid %>%
+  # filter(num == 1) %>%
+  tidyr::unnest_wider("res") %>%
+  tidyr::pivot_longer(-(1:5), names_to = "Method") %>%
+  tidyr::unnest_wider("value") %>%
+  # head(20) %>%
+  rowwise() %>%
+  mutate(
+    r2 = list(unname(r2)),
+    n_keep_infer = `if`(grepl("LDpred2", Method), `if`(is.null(all_h2), 0, NCOL(all_h2)), NA),
+    n_keep_pred  = `if`(grepl("LDpred2", Method), `if`(is.null(all_r2), 0, NCOL(all_r2) / 500 * 20), NA),
+    r2_est = list(`if`(Method == "LDSc", NULL,
+                       unname(quantile(all_r2, probs = c(0.5, 0.025, 0.975))))),
+    h2_est = list(
+      `if`(Method == "LDSc", all_h2[["h2"]] + c(0, -1.96, 1.96) * all_h2[["h2_se"]],
+           unname(quantile(all_h2, probs = c(0.5, 0.025, 0.975))))
+    ),
+    p_est = list(`if`(Method == "LDSc", NULL,
+                      unname(quantile(all_p, probs = c(0.5, 0.025, 0.975))))),
+    alpha_est = list(`if`(Method %in% c("LDSc", "LDpred2_noMLE"), NULL,
+                          unname(quantile(all_alpha, probs = c(0.5, 0.025, 0.975))))),
+  ) %>%
+  relocate(n_keep_infer, n_keep_pred, .after = use_mle) %>%
+  ungroup() %>%
+  mutate(Method = factor(Method, levels = Method[1:6],
+                         labels = sub("LDpred2_", "LDpred2-auto_", Method[1:6]))) %>%
[…]

### 2023-08-31 code/simu-binary.R
just changed method name

@@ -97,13 +97,44 @@ grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num) {
   tibble::tibble(logistic = c(FALSE, TRUE), res = list(res_lin, res_log))
 })
 
-vec_to_liab <- function(x, K, logistic) {
-  x * purrr::map2_dbl(K, logistic, ~ coef_to_liab(.x, `if`(.y, 0.5, .x)))
-}
+
+library(dplyr)
+
+grid %>%
+  tidyr::unnest_wider(res) %>%
+  tidyr::unnest(cols = c(logistic, res)) %>%
+  tidyr::unnest_wider(res) %>%
+  tidyr::pivot_longer(-(1:6), names_to = "Method") %>%
+  tidyr::unnest_wider("value") %>%
+  # head(20) %>%
+  rowwise() %>%
+  mutate(
+    r2 = list(unname(r2)),
+    n_keep = `if`(grepl("LDpred2", Method), `if`(is.null(all_h2), 0, NCOL(all_h2)), NA),
+    r2_est = list(`if`(Method == "LDSc", NULL,
+                       unname(quantile(all_r2, probs = c(0.5, 0.025, 0.975))))),
+    h2_est = list(
+      `if`(Method == "LDSc", all_h2[["h2"]] + c(0, -1.96, 1.96) * all_h2[["h2_se"]],
+           unname(quantile(all_h2, probs = c(0.5, 0.025, 0.975))))
+    ),
+    p_est = list(`if`(Method == "LDSc", NULL,
+                      unname(quantile(all_p, probs = c(0.5, 0.025, 0.975))))),
+    alpha_est = list(`if`(Method %in% c("LDSc", "LDpred2_noMLE"), NULL,
+                          unname(quantile(all_alpha, probs = c(0.5, 0.025, 0.975))))),
+  ) %>%
+  relocate(n_keep, .after = use_mle) %>%
+  ungroup() %>%
+  mutate(Method = factor(Method, levels = Method[1:4],
+                         labels = sub("LDpred2_", "LDpred2-auto_", Method[1:4]))) %>%
+  select(-starts_with("all"), -postp) %>%
+  tidyr::unnest_wider(where(is.list), names_sep = "_") %>%
+  mutate(across(r2_1:alpha_est_3, ~ signif(., digits = 4))) %>%
+  print(n = 20) %>%
[…]

### 2023-08-31 paper/paper-infer-supp.tex
split in two figures

@@
 \begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_r2_noMLE}}
+	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
+These are stratified by the polygenicity estimated from LDpred2-auto.
+Green dashed lines represent the 1:1 line.
+The 95\% confidence interval for the LDpred2-auto estimate is obtained from the 2.5\% and 97.5\% quantiles of all the $r^2$ estimates from the iterations (after burn-in) of the chains kept.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+Colors represent the number of chains kept (out of 50).
+``F\_height'' and ``M\_height'' use females and males only, respectively (in both GWAS and test sets).
+\label{fig:ukbb_r2_noMLE}}
+\end{figure}
+
+\begin{figure}[p]
 \centerline{\includegraphics[width=\textwidth]{ukbb_compare_hm3_small}}

### 2023-08-31 paper/paper-infer.tex
split in two figures

@@
 Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}).
-As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figure \ref{fig:ukbb_r2}).
+As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and \ref{fig:ukbb_r2_noMLE}).
 Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
@@
 \begin{figure}[!htb]
-	\centerline{\includegraphics[width=0.9\textwidth]{ukbb_r2}}
+	\centerline{\includegraphics[width=0.8\textwidth]{ukbb_r2_MLE}}
 	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.

### 2023-08-31 paper/paper-infer-supp.tex
add dense fine-mapping

@@
 
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{manhattan_height_one_region}}
+	\caption{GWAS of height in a restricted region containing the most significant HapMap3+ variant for height. \label{manhattan_height}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{pip_height_one_region}}
+	\caption{Posterior inclusion probabilities (PIPs) for the variants in Figure \ref{manhattan_height}. 
+	L is the maximum number of causal variants in SuSiE-RSS, and max(p) is the maximum value of the estimated $p$ in LDpred2-auto.
+	Both panels A and B represent the same data, but colors and facets are changed. \label{pip_height}}
+\end{figure}
+
 \FloatBarrier

### 2023-08-31 paper/paper-infer.tex
add dense fine-mapping

@@
 
+Finally, since the HapMap3+ variants still represent a rather small proportion of common variants, we showcase running LDpred2-auto using a more dense set of variants from a small region, as usually done for fine-mapping.
+We identify the most significant HapMap3+ variant for height, rs2871960 on chromosome 3, and read all the variants within a 500kb distance that have a MAF larger than 0.005 and INFO score larger than 0.5 in the UK Biobank; there are 3881 such variants.
+Then, we perform a GWAS for these variants using the same training individuals as used before (Figure \ref{manhattan_height}).
+Finally, using the resulting GWAS summary statistics and an LD reference from the same set of individuals, we run both SuSiE-RSS and LDpred2-auto (\_nojump).
+We test two different values for L, the maximum number of causal variants in SuSiE-RSS (10 and 100), and three different values for the maximum value of the estimated $p$ (1, 0.01, and 0.001).
+Per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities, PIPs) are very similar with both methods, especially when restricting $p$ in LDpred2-auto, which is similar to what SuSiE-RSS assumes (more conservative).
+
 \subsection*{Application to height}
@@
 Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
+For now, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').
 

### 2023-08-31 code/height-one-region.R

@@ -1,0 +1,185 @@
+library(bigsnpr)
+library(dplyr)
+
+NCORES <- 13
+
+
+fam <- snp_attach(paste0("data/UKBB_hm3_plus.rds"))$fam %>%
+  filter(set == "train")
+
+# Get top hit of GWAS in HM3+
+gwas0 <- purrr::map_dfr(paste0("GWAS/height_hm3_plus_part", 1:30, ".rds"), readRDS)
+ind_top <- which.max(abs(gwas0$score))
+map_hm3_plus <- readRDS("map_hm3_plus.rds")
+stopifnot(nrow(map_hm3_plus) == nrow(gwas0))
+
+map_hm3_plus[ind_top, ]
+# chr       pos a0    a1    rsid      af_UKBB    ld block_id  pos_hg18  pos_hg38
+#   3 141121814 A     C     rs2871960   0.448  27.1       50 142604504 141402972
+
+# Use 1 Mb region around the top variant
+pos <- map_hm3_plus$pos[ind_top]
+chr <- map_hm3_plus$chr[ind_top]
+mfi <- bigreadr::fread2(paste0("UKBB/mfi/ukb_mfi_chr", chr, "_v3.txt"))
+snp_id <- mfi %>%
+  filter(V6 > 0.005, V8 > 0.5, between(V3, pos - 500e3, pos + 500e3)) %>%
+  with(paste(chr, V3, V4, V5, sep = "_"))
+length(snp_id)  # 3881
+
+# Read dosage data
+snp_readBGEN(
+  bgenfiles = glue::glue("UKBB/bgen/ukb_imp_chr{chr}_v3.bgen", chr = chr),
+  list_snp_id = list(snp_id),
+  backingfile = "tmp-data/height_one_region",
+  ind_row = fam$ind_bgen,
+  ncores = NCORES
+)
+
+
+ukbb <- snp_attach("tmp-data/height_one_region.rds")
+G <- ukbb$genotypes
+y <- readRDS("data/ukbb-quant-pheno/height.rds")[fam$ind_csv]
+covar <- fam %>% select(PC1:date) %>% as.matrix()
+ind.train <- which(!is.na(y) & complete.cases(covar))
+length(ind.train)  # 305338
+
+gwas <- big_univLinReg(G, y[ind.train], ind.train = ind.train,
+                       covar.train = covar[ind.train, ], ncores = NCORES)
+
+lpval <- predict(gwas)
[…]

### 2023-09-01 paper/paper-infer.tex
more on model comparisons

@@
 Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
-The distribution of causal effects relative to their allele frequencies is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
-In this model, the expected variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
+As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
+In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
@@
 LDpred2 is a Bayesian approach that uses the SNP heritability $h^2$ and polygenicity $p$ as parameters of its model.
-LDpred2-auto, one variant of LDpred2, can directly estimate these two parameters from the data, making it applicable even when no validation data is available for tuning these two model hyper-parameters \cite[]{prive2020ldpred2}.
+LDpred2-auto, one variant of LDpred2, can directly estimate these two parameters from GWAS summary statistics, making it applicable even when no validation data is available for tuning these two model parameters \cite[]{prive2020ldpred2}.
 
@@
 Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without requiring an independent test set.
-An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
+An overview of what LDpred2-auto can now provide is presented in Figure \ref{fig:overview}.
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
@@
 Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
-CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when $h^2$ and $N$ are small (Figures \ref{fig:simu10_r2} and \ref{fig:simu10_r2_large}).
+CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when both $h^2$ and $N$ are small (Figures \ref{fig:simu10_r2} and \ref{fig:simu10_r2_large}).
 
@@
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the new 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE}.
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the new 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}).
 
@@
 Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
-Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ normal sampling, Figure \ref{fig:jumpsign}).
+Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure \ref{fig:jumpsign}).
 $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
-As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the normal LD reference) and normal sampling can be used in this case (Figure \ref{fig:jumpsign}).
+As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W.
+European LD reference) and a standard sampling can be used in this case (Figure \ref{fig:jumpsign}).
 
@@
 We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
-Except for bilirubin and lipoprotein(a) concentration, generally higher $h^2$ estimates and predictive performance $r^2$ are obtained with the RIN-transformation than the log-transformation.
+Except for bilirubin and lipoprotein(a) concentration, higher $h^2$ estimates and predictive performance $r^2$ are generally obtained with the RIN-transformation than the log-transformation.
 
-\subsection*{More heritability and predictive accuracy with new set of variants}
+\subsection*{More heritability and predictive accuracy with the new set of variants}
 
@@
 We identify the most significant HapMap3+ variant for height, rs2871960 on chromosome 3, and read all the variants within a 500kb distance that have a MAF larger than 0.005 and INFO score larger than 0.5 in the UK Biobank; there are 3881 such variants.
-Then, we perform a GWAS for these variants using the same training individuals as used before (Figure \ref{manhattan_height}).
+Then, we perform a GWAS for these variants using the same 305,338 training individuals as used before (Figure \ref{manhattan_height}).
 Finally, using the resulting GWAS summary statistics and an LD reference from the same set of individuals, we run both SuSiE-RSS and LDpred2-auto (\_nojump).
@@
 We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
-Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K and slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
+Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K, and of slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
 
@@
 Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
+The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure \ref{fig:ukbb_r2_vs}).
 Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
@@
 These do not include UKBB individuals.
-Note that some of them may contain a substantial amount of non-European genetic ancestry (e.g.\ >20\%) for CAD.
+Note that some of them may contain a substantial amount of non-European genetic ancestry (e.g.\ >20\% for CAD).
 \label{tab:sumstats}}
@@
 Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
-We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large}.
+We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
 We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
@@
 When using external GWAS summary statistics, using HapMap3+ instead of HapMap3 variants was more beneficial for larger sample sizes.
-Our intuition and conclusion is that using more variants is beneficial when power is sufficient; however, when power is low (e.g.\ small $N$, small $h^2$, and/or large $p$), it may be actually be detrimental.
+Our intuition and conclusion is that using more variants is beneficial when power is sufficient; however, when power is low (e.g.\ small $N$, small $h^2$, and/or large $p$), it may actually be detrimental.
 
 Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
-Therefore, we recommend to use the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient.
+Therefore, we recommend to use the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
 We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates.
@@
 Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
-For now, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').
+For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').
 
@@
 The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions.
-As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to use more variants and learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating,zheng2022leveraging}.
+As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to use more variants and to leverage functional annotations \cite[]{zhang2021improved,marquez2021incorporating,zheng2022leveraging}.
 

### 2023-09-01 code/run-ldpred2-ukbb.R

@@ -222,6 +222,17 @@ median(purrr::map_dbl(filter(grid2, set == "hm3_plus")$time, 3)) /
   median(purrr::map_dbl(filter(grid2, set == "hm3")$time, 3))
 # 1.45
 
+grid2 %>%
+  mutate(pheno = sub("\\.rds$", "", basename(pheno_file))) %>%
+  relocate(pheno, .before = 0) %>%
+  select(-c(pheno_file, postp, res_file, time, local_h2)) %>%
+  relocate(r2, .after = r2_est) %>%
+  tidyr::unnest_wider(h2_ldsc_est:r2, names_sep = "_") %>%
+  mutate(across(h2_ldsc_est_1:r2_3, signif)) %>%
+  print() %>%
+  bigreadr::fwrite2("results_infer_ukbb.csv")
+
+
 grid3 <- grid2 %>%
   filter(n_keep > 0) %>%
   tidyr::unnest_wider("r2", names_sep = "_") %>%
@@ -235,6 +246,7 @@ grid3 <- grid2 %>%
 grid3_hm3 <- filter(grid3, set == "hm3")
 
 grid3_hm3 %>%
+  filter(!use_mle) %>%
   mutate(h2 = cut(h2_est_1, c(0, 0.05, 0.1, 0.2, 1))) %>%
   ggplot(aes(r2_1, r2_est_1, color = n_keep)) +
   scale_color_gradient(high = "#0072B2", low = "#D55E00", limits = c(0, 50)) +
@@ -246,14 +258,31 @@ grid3_hm3 %>%
                 position = position_dodge(width = 0.9), width = 0) +
   ggrepel::geom_label_repel(aes(label = sub("\\.rds$", "", basename(pheno_file))),
                             color = "purple", min.segment.length = 0, force = 500, size = 3, seed = 4,
-                            data = mutate(filter(grid3_hm3, grepl("height", pheno_file)),
+                            data = mutate(filter(grid3_hm3, grepl("height", pheno_file), !use_mle),
                                           h2 = cut(h2_est_1, c(0, 0.05, 0.1, 0.2, 1)))) +
   facet_wrap(~ use_mle + h2, scales = "free", labeller = label_both, nrow = 2) +
   geom_point(size = 1.5) +
-  theme(legend.position = "top", legend.key.width = unit(40, "pt")) +
[…]

### 2023-09-01 paper/paper-infer.tex
formatting
final typos

@@
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$, on average, in UK Biobank analyses.
-We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend to use it when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
+We call `HapMap3+' this extended set of 1,444,196 SNPs, and recommend using it when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
 
@@
 For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}.
-HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend to use for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
+HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
 CI means ``confidence interval'' (often called ``credible interval'' in a Bayesian setting).
@@
 
-\clearpage
+%\clearpage
 \section*{Results}
@@
 
-\clearpage
+%\clearpage
 \section*{Discussion}
@@
 Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
-Therefore, we recommend to use the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
+Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
 We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates.
@@
 Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by.
-LDpred2-auto can also totally miss some causal variants when they are poorly tagged by the set of variants used.
+LDpred2-auto can also miss some causal variants when they are poorly tagged by the set of variants used.
 Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
@@
 
-\clearpage
+%\clearpage
 \section*{Materials and Methods}
@@
 
+\vspace*{3em}
 %\clearpage
-\vspace*{3em}
 

### 2023-10-02 paper/paper-infer-supp.tex
more on model comparisons
avoid "new" if possible

@@
 		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
-		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``altfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
-``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``altfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
-		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``altfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
-For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``altfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
@@
 		For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and 
-		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
-		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+		``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 Horizontal dotted lines represent boundaries imposed on the LDpred2-auto estimates.
-For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}).
 Note that ``LDpred2\_noMLE'' (\texttt{use\_MLE = FALSE}) does not infer $\alpha$.
@@
 For LDpred2-auto, suffix ``nojump''/``jump'' refers to using \texttt{allow\_jump\_sign = FALSE/TRUE} (and \texttt{use\_MLE = TRUE}), and
-	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``newfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
+	``noMLE'' refers to using \texttt{use\_MLE = FALSE} (and \texttt{allow\_jump\_sign = FALSE}), and ``altfilter'' is similar to ``nojump'' but uses a different post-filtering of chains (Methods).
 Note that the recommended option is to use \texttt{allow\_jump\_sign = FALSE} \cite[]{prive2021identifying}.
@@
 The 95\% confidence interval for the LD Score regression estimate is obtained from $\pm$1.96 times its standard error.
-``use\_mle: TRUE'' corresponds to using the new 3-parameter model and sampling scheme (Methods).
+``use\_mle: TRUE'' corresponds to using the extended 3-parameter model and sampling scheme (Methods).
 Colors represent the number of chains kept for LDpred2-auto (out of 50).
@@
 \centerline{\includegraphics[width=0.85\textwidth]{ukbb_compare_mle}}
-\caption{$r^2$ estimates (in the test set) for UKBB phenotypes, using LDpred2-auto with either the new sampling (MLE, 3-parameter model) or the previous one (2-parameter model).
+\caption{$r^2$ estimates (in the test set) for UKBB phenotypes, using LDpred2-auto with either the extended model and sampling (MLE, 3-parameter model) or the previous one (2-parameter model).
 Red dashed lines represent the 1:1 line.
@@
 	\caption{Distribution of the number of LDpred2-auto chains kept across 248 UKBB phenotypes.
-``Use MLE'' corresponds to using the new 3-parameter model and sampling scheme (Methods).
+``Use MLE'' corresponds to using the extended 3-parameter model and sampling scheme (Methods).
 \label{fig:ukbb_chains}}
@@
 	\caption{Distribution of LDpred2-auto runtimes across 248 UKBB phenotypes.
-	``Use MLE'' corresponds to using the new 3-parameter model and sampling scheme (Methods). For each phenotype, 50 chains are used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
+	``Use MLE'' corresponds to using the extended 3-parameter model and sampling scheme (Methods). For each phenotype, 50 chains are used, parallelized over 13 cores. \label{fig:ukbb_runtimes}}
 \end{figure}
@@
 	\caption{LDpred2-auto results with external GWAS summary statistics.
-We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}, where setting to \texttt{TRUE} uses the new model), and explore multiple values for parameter \texttt{coef\_shrink} (multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix).
+We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}, where setting to \texttt{TRUE} uses the extended model), and explore multiple values for parameter \texttt{coef\_shrink} (multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix).
 The UK Biobank is used as test set to compute $r^2$.\label{fig:sumstats}}
+\end{figure}
+
+\begin{figure}[p]
+	\centerline{\includegraphics[width=\textwidth]{ukbb_r2_vs}}
+	\caption{Predictive performance $r^2$ obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
+The red dashed line represent the 1:1 line.
+The 95\% confidence interval for $r^2$ in the test set is obtained from bootstrap.
+\label{fig:ukbb_r2_vs}}
 \end{figure}

### 2023-10-02 paper/paper-infer.tex
reorder sections
avoid "new" if possible

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\clearpage
+\section*{Materials and Methods}
+
+\subsection*{Data for simulations}
+
+For simulations, we use the UK Biobank imputed (BGEN) data, read as allele dosages with function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{bycroft2018uk,prive2017efficient}. 
+We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
+Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
+We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
+These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
+To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
+This results in 356,409 individuals of Northwestern European ancestry.
+We randomly sample 200,000 individuals to form a training set (to run the GWAS), and use the remaining individuals to form a test set (to evaluate the predictive models).
+
+\subsection*{Data for the UK Biobank analyses}
+
+We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 individuals of Northwestern European ancestry as in the simulations.
+We randomly sample 50,000 individuals to form a test set (to evaluate the predictive models), and use the remaining individuals to form a training set (to run the GWAS).
+
+We construct and use the same phenotypes as in \cite{prive2021high}.
+About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
+The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
+As covariates, we first recompute PCs for the homogeneous subset of individuals previously defined using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
+We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).
+
+We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
+We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
+We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
+We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').
+
+\subsection*{Extending the set of HapMap3 variants used}
+
+The HapMap3 variants generally provide a good coverage of the whole genome.
+We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
+Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
+To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
+There are around 11.5M such variants.
+Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
+Finally, we design an algorithm which aims at maximizing the tagging of all these variants read.
+We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
+We start by including all previously used HapMap3 variants.
+Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
+Note that we only allow non-ambiguous SNPs to be included.
+This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
+Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (resp.\ 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).
+
+\subsection*{Optional extended model and inference with LDpred2-auto}
+
+LDpred2 originally assumed the following model for effect sizes,
+\begin{equation}\label{eq:prev_model}
+\beta_j = S_j \gamma_j \sim \left\{
+\begin{array}{ll}
+\mathcal N\left(0,~\dfrac{h^2}{M p}\right) & \mbox{with probability $p$,} \\
+0 & \mbox{otherwise,}
+\end{array}
+\right.
+\end{equation}
+where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\boldsymbol{\gamma}$ the effect sizes on the allele scale, $\boldsymbol{S}$ the standard deviations of the genotypes, and $\boldsymbol{\beta}$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
+In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters (as in LDpred2-grid). 
+This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly without requiring a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
+$h^2$ is estimated by $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
+As for $p$, it is sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
+Note a small change, we now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more variability in the sampling in order to account for a reduced effective number of (independent) variants.
+
+Here we provide a extended model and sampling scheme for LDpred2-auto that can be optionally used by setting parameter \texttt{use\_MLE = TRUE} (otherwise it is run as described in the previous paragraph when using \texttt{use\_MLE = FALSE}).
+In this new option, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
+\begin{equation}\label{eq:new_model}
+\beta_j = S_j \gamma_j \sim \left\{
+\begin{array}{ll}
+\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
+0 & \mbox{otherwise.}
+\end{array}
+\right.
+\end{equation}
+Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
+This extended model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is denoted by $S$ instead \cite[]{zeng2021widespread}. 
+In SBayesS, $\alpha$ and $\sigma_\beta^2$ are estimated by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
+When using this 3-parameter model in LDpred2-auto, in order to add some sampling to these two parameters, we first sample causal variants with replacement before computing the maximum likelihood estimators.
+This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
+Note that we still estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}). 
+Note that, to get local heritability estimates (e.g.\ for a single LD block), this $h^2$ estimation ($\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$) is simply restricted to the variants from this LD block.
+Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the vector of scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
+
+\subsection*{Inference of predictive performance $r^2$}
+
+To infer the out-of-sample predictive performance $r^2$ (and its CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
+Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
+Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
+These previous works have shown that $r^2$ can be approximated by $\big(\boldsymbol{\bar{\beta}}^T \boldsymbol{\hat{\beta}}_\text{test}\big)^2 / \big(\boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}\big)$ where $\boldsymbol{\bar{\beta}}$ and $\boldsymbol{\hat{\beta}}_\text{test}$ are respectively the predictive effects from the training set and the marginal effects from the test set (after scaling).
+Note that $E[\boldsymbol{\beta_1}] = \boldsymbol{\bar{\beta}}$, $E[\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}] = \boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}$, and, when the test sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics), $E[\boldsymbol{R} \boldsymbol{\beta_2}] \approx \boldsymbol{\hat{\beta}}_\text{test}$.
+Therefore, we propose to use $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ as a sample of $r^2$.
+This can also be computed for a specific chain by taking two samples $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ that are far enough on the same chain to remove the possible autocorrelation. 
+This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
+Note that \cite{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
+Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.
+
+In this paper, we check this new $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
+These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
+The partial correlation is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign).
+Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.
+
+\subsection*{Post-filtering of chains in LDpred2-auto}
+
+Because a Gibbs sampler can be unstable, with so many variants and with possible mismatches between e.g.\ the GWAS summary statistics and the LD reference used, we have always recommended to run multiple chains in LDpred2-auto, and post-filter some of them as quality control \cite[]{prive2020ldpred2}.
+We originally proposed to filter chains by keeping the ones providing PGS with the largest variances. 
+Then we tested an almost equivalent and simpler alternative \cite[]{prive2021identifying}, which is to keep only chains that provide top imputed marginal scaled effect sizes $\check{\beta} = R \bar{\beta}$, where $R$ is the correlation matrix and $\bar{\beta}$ are the PGS scaled effect sizes.
+This is the default post-filtering of LDpred2-auto chains that we use here, for both prediction and inference.
+
+Moreover, here we test two alternative filtering criteria in the first simulations based on continuous outcomes (and call this ``LDpred2-auto\_altfilter''). 
+First, for prediction, we test filtering on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ within each chain, which is an estimate of $r^2$ (cf.\ previous Methods section).
+Second, for inference, we filter on some convergence criterion. The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}. 
+However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain does not mix well, but a perfect Rhat of 1.0 is obtained. 
+Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat statistic while not suffering from the previous issue.
+We therefore chose to use this statistic, and average the three statistics computed for $h^2$, $p$, and $\alpha$ for each chain.
+We use a threshold of 4 above which we filter out the chain, because we have found that a value of 4 for this statistic often corresponds to a value close to 1.05 for the split-Rhat.
+
 %\clearpage
@@
 Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
-Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the new model including a third parameter $\alpha$  (new option \texttt{use\_MLE}, Methods).
+Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods).
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
@@
 SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
-The $\alpha$ estimate of LDpred2-auto (with the new 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
+The $\alpha$ estimate of LDpred2-auto (with the extended 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
 Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures \ref{fig:simu10_p} and \ref{fig:simu10_alpha}).
 Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures \ref{fig:simu10_h2}--\ref{fig:simu10_alpha_large}), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
-Finally, we also investigate new ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_newfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
+Finally, we also investigate alternative ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_altfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
 
@@
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the new 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}).
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}).
 
@@
 Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
-We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the new  or previous sampling and model (via parameter \texttt{use\_MLE}).
+We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}).
 Because of the increased mismatch between external GWAS summary statistics and the LD reference we use here (compared to in simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
@@
 In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); therefore it seems more useful to use this new set of variants for larger sample sizes.
-Using the new sampling scheme and model (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
+Using the extended model and sampling scheme (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
 The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure \ref{fig:ukbb_r2_vs}).
@@
 Future work could focus on correcting these issues.
-Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the new sampling and model), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
+Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the extended model and sampling), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
 Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
@@
 As future research directions, we are interested in using LDpred2-auto for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to use more variants and to leverage functional annotations \cite[]{zhang2021improved,marquez2021incorporating,zheng2022leveraging}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\clearpage
-\section*{Materials and Methods}
-
-\subsection*{Data for simulations}
-
-For simulations, we use the UK Biobank imputed (BGEN) data, read as allele dosages with function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{bycroft2018uk,prive2017efficient}. 
-We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Since we run lots of different models, we restrict the simulations to chromosomes 3, 6, 9, 12, 15, 18 and 21, resulting in a set of 322,805 SNPs.
-We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
-These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
-To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
-This results in 356,409 individuals of Northwestern European ancestry.
-We randomly sample 200,000 individuals to form a training set (to run the GWAS), and use the remaining individuals to form a test set (to evaluate the predictive models).
-
-\subsection*{Data for the UK Biobank analyses}
-
-We use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}, and the same 356,409 individuals of Northwestern European ancestry as in the simulations.
-We randomly sample 50,000 individuals to form a test set (to evaluate the predictive models), and use the remaining individuals to form a training set (to run the GWAS).
-
-We construct and use the same phenotypes as in \cite{prive2021high}.
-About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
-The other half consists of phenotypes defined in UKBB fields based on manual curation \cite[]{prive2021high}.
-As covariates, we first recompute PCs for the homogeneous subset of individuals previously defined using function \texttt{snp\_autoSVD} from R package bigsnpr and keep four PCs based on visual inspection \cite[]{prive2017efficient,prive2020efficient}.
-We also use sex (field 22001), age (field 21022), birth date (combining fields 34 and 52) and deprivation index (field 189) as additional covariates (to a total of eight).
-
-We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
-We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
-We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
-We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').
-
-\subsection*{Extending the set of HapMap3 variants used}
-
-The HapMap3 variants generally provide a good coverage of the whole genome.
-We recall that the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2} is a subset of the original set of HapMap3 variants, which does not include duplicated positions (e.g.\ multi-allelic variants), nor ambiguous variants (e.g.\ `A' and `T', or `C' and `G'), and which includes SNPs only (e.g.\ no indel).
-Here we propose to extend this set of 1,054,330 HapMap3 variants to make sure many genetic variants are well tagged by the extended set.
-To design this new set, we first read all variants from the UK Biobank (UKBB) with a minor allele frequency (MAF) larger than 0.005 in the whole data (i.e.\ the MAF from the MFI files). 
-There are around 11.5M such variants.
-Then we restrict to unrelated UKBB individuals which are \textit{not} listed as White British (field 22006) and use these individuals of diverse ancestries \cite[]{bycroft2018uk,prive2021using} to compute all pairwise correlations between variants within a 1 Mb distance, restricting to squared correlations larger than 0.3.
-Finally, we design an algorithm which aims at maximizing the tagging of all these variants read.
-We want to maximize $\sum_j \max_{k \in \text{HapMap3+}} r_{j,k}^2$, where $j$ spans the whole set of variants read, while $k$ spans the variants kept in the new set, which we call HapMap3+.
-We start by including all previously used HapMap3 variants.
-Then, for the sake of simplicity, we use a greedy approach, where we repeatedly include the variant which increases this sum most, until no variant improves it by more than 2. 
-Note that we only allow non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
-Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
-Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (resp.\ 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).
-
-\subsection*{New optional model and inference with LDpred2-auto}
-
-LDpred2 originally assumed the following model for effect sizes,
-\begin{equation}\label{eq:prev_model}
-\beta_j = S_j \gamma_j \sim \left\{
-\begin{array}{ll}
-\mathcal N\left(0,~\dfrac{h^2}{M p}\right) & \mbox{with probability $p$,} \\
-0 & \mbox{otherwise,}
-\end{array}
-\right.
-\end{equation}
-where $p$ is the proportion of causal variants, $M$ the number of variants, $h^2$ the (SNP) heritability, $\boldsymbol{\gamma}$ the effect sizes on the allele scale, $\boldsymbol{S}$ the standard deviations of the genotypes, and $\boldsymbol{\beta}$ the effects of the scaled genotypes \cite[]{prive2020ldpred2}.
-In LDpred2-auto, $p$ and $h^2$ are directly estimated within the Gibbs sampler, as opposed to testing several values of $p$ and $h^2$ from a grid of hyper-parameters (as in LDpred2-grid). 
-This makes LDpred2-auto a method free of hyper-parameters which can therefore be applied directly without requiring a validation dataset to choose best-performing hyper-parameters \cite[]{prive2020ldpred2}.
-$h^2$ is estimated by $\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler.
-As for $p$, it is sampled from $\text{Beta}(1 + M_c, 1 + M - M_c)$, where $M_c = \sum_j(\beta_j \neq 0)$.
-Note a small change, we now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more variability in the sampling in order to account for a reduced effective number of (independent) variants.
-
-Here we provide a new model and sampling scheme for LDpred2-auto that can be optionally used by setting parameter \texttt{use\_MLE = TRUE} (otherwise it is run as described in the previous paragraph when using \texttt{use\_MLE = FALSE}).
-In this new option, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
-\begin{equation}\label{eq:new_model}
-\beta_j = S_j \gamma_j \sim \left\{
-\begin{array}{ll}
-\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
-0 & \mbox{otherwise.}
-\end{array}
-\right.
-\end{equation}
-Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
-This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is denoted by $S$ instead \cite[]{zeng2021widespread}. 
-In SBayesS, $\alpha$ and $\sigma_\beta^2$ are estimated by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
-In the new LDpred2-auto, to add some sampling to these two parameters, we first sample causal variants with replacement before computing the maximum likelihood estimators. 
-This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
-Note that we still estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}). 
-Note that, to get local heritability estimates (e.g.\ for a single LD block), this $h^2$ estimation ($\boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$) is simply restricted to the variants from this LD block.
-Finally, in both models and sampling schemes now implemented in LDpred2-auto, we now detect strong divergence when $\boldsymbol{\beta}^T \boldsymbol{\beta} > 2 \boldsymbol{\hat{\beta}}^T \boldsymbol{\hat{\beta}}$, where $\boldsymbol{\beta}$ is the vector of scaled effect sizes from one iteration of the Gibbs sampler and $\boldsymbol{\hat{\beta}}$ is the marginal scaled effect sizes; corresponding chains are stopped and missing values are returned for effect sizes and estimates of missing iterations.
-
-\subsection*{Inference of predictive performance $r^2$}
-
-To infer the out-of-sample predictive performance $r^2$ (and its CI) of the resulting PGS from LDpred2-auto, we use the distribution of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$, where $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are two sampled vectors of causal effect sizes (after scaling) from two different chains of the Gibbs sampler.
-Intuitively, if prediction is perfect then $\boldsymbol{\beta_1} = \boldsymbol{\beta_2}$ and $r^2 = h^2$; when power is very low, $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ are almost uncorrelated and $r^2 \approx 0$.
-Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
-These previous works have shown that $r^2$ can be approximated by $\big(\boldsymbol{\bar{\beta}}^T \boldsymbol{\hat{\beta}}_\text{test}\big)^2 / \big(\boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}\big)$ where $\boldsymbol{\bar{\beta}}$ and $\boldsymbol{\hat{\beta}}_\text{test}$ are respectively the predictive effects from the training set and the marginal effects from the test set (after scaling).
-Note that $E[\boldsymbol{\beta_1}] = \boldsymbol{\bar{\beta}}$, $E[\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}] = \boldsymbol{\bar{\beta}}^T \boldsymbol{R} \boldsymbol{\bar{\beta}}$, and, when the test sample has the same genetic ancestry as the GWAS used for training (to get the summary statistics), $E[\boldsymbol{R} \boldsymbol{\beta_2}] \approx \boldsymbol{\hat{\beta}}_\text{test}$.
-Therefore, we propose to use $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ as a sample of $r^2$.
-This can also be computed for a specific chain by taking two samples $\boldsymbol{\beta_1}$ and $\boldsymbol{\beta_2}$ that are far enough on the same chain to remove the possible autocorrelation. 
-This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
-Note that \cite{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
-Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.
-
-In this paper, we check this new $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
-These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
-The partial correlation is computed with function \texttt{pcor} from R package bigstatsr, adjusting for the same eight covariates as in the GWAS, then squared (while keeping the sign).
-Corresponding 95\% confidence intervals are estimated through bootstrapping individual indices.
-
-\subsection*{Post-filtering of chains in LDpred2-auto}
-
-Because a Gibbs sampler can be unstable, with so many variants and with possible mismatches between e.g.\ the GWAS summary statistics and the LD reference used, we have always recommended to run multiple chains in LDpred2-auto, and post-filter some of them as quality control \cite[]{prive2020ldpred2}.
-We originally proposed to filter chains by keeping the ones providing PGS with the largest variances. 
-Then we tested an almost equivalent and simpler alternative \cite[]{prive2021identifying}, which is to keep only chains that provide top imputed marginal scaled effect sizes $\check{\beta} = R \bar{\beta}$, where $R$ is the correlation matrix and $\bar{\beta}$ are the PGS scaled effect sizes.
-This is the default post-filtering of LDpred2-auto chains that we use here, for both prediction and inference.
-
-Moreover, here we test two other filtering criteria in the first simulations based on continuous outcomes (and call this ``LDpred2-auto\_newfilter''). 
-First, for prediction, we test filtering on the average of $\boldsymbol{\beta_1}^T \boldsymbol{R} \boldsymbol{\beta_2}$ within each chain, which is an estimate of $r^2$ (cf.\ previous Methods section).
-Second, for inference, we filter on some convergence criterion. The split-Rhat statistic is a popular metric to test for good mixing and convergence of Markov chains \cite[]{vehtari2021rank}. 
-However, we have found this statistic to perform badly when e.g.\ one parameter gets stuck and is constant; in this case, the chain does not mix well, but a perfect Rhat of 1.0 is obtained. 
-Instead, we have found that a two-sample Cramer-von Mises statistic \cite[]{anderson1962distribution}, by similarly using both parts of the chain after burn-in, is often highly correlated with the split-Rhat statistic while not suffering from the previous issue.
-We therefore chose to use this statistic, and average the three statistics computed for $h^2$, $p$, and $\alpha$ for each chain.
-We use a threshold of 4 above which we filter out the chain, because we have found that a value of 4 for this statistic often corresponds to a value close to 1.05 for the split-Rhat.
 

### 2023-10-02 paper/paper-infer.tex
rm links to suppmat

@@
 
-\usepackage{xr}
-\externaldocument{paper-infer-supp}
+%\usepackage{xr}
+%\externaldocument{paper-infer-supp}
 
@@
 
-First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2_main}--\ref{fig:simu_alpha_main} and \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
-Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2_main}, \ref{fig:simu_h2} and \ref{fig:simu_h2_large}).
-When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of $h^2$ (i.e.\ CIs are small, and do not contain the true parameter value, Figure \ref{fig:simu10_h2}), and LDpred2-auto\_jump overestimates the heritability (Figure \ref{fig:simu_h2}).
+First, LDpred2-auto generally reliably infers the three parameters from its model, i.e.\ the SNP heritability $h^2$, polygenicity $p$, and $\alpha$ (Figures \ref{fig:simu_h2_main}--\ref{fig:simu_alpha_main} and S1--S6).
+Compared to LD Score regression, heritability estimates are as precise when power is low, and much more precise when power is large, especially for small polygenicity values (Figures \ref{fig:simu_h2_main}, S1 and S2).
+When power is low (e.g.\ $N=20000$ and $h^2=0.01)$, LDpred2-auto\_noMLE (with only two model parameters, as in previous versions of LDpred2-auto) and SBayesS are both over-confident in their estimate of $h^2$ (i.e.\ CIs are small, and do not contain the true parameter value, Figure S7), and LDpred2-auto\_jump overestimates the heritability (Figure S1).
 LDpred2-auto\_nojump looks very reliable for estimating $h^2$.
-When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure \ref{fig:simu_p}).
+When power is low, LDpred2-auto can overestimate the polygenicity when the true value is very small (e.g.\ $p=0.0005$), and underestimate it when the polygenicity is large (e.g.\ $p=0.1$, Figure S3).
 SBayesS, which uses a similar model with the same three parameters, often overestimates the polygenicity, especially when $p\le0.02$.
 The $\alpha$ estimate of LDpred2-auto (with the extended 3-parameter model) can become very unprecise when power is too low, which can be detected by a small number of chains kept from LDpred2-auto.
-Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures \ref{fig:simu10_p} and \ref{fig:simu10_alpha}).
-Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures \ref{fig:simu10_h2}--\ref{fig:simu10_alpha_large}), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
-Finally, we also investigate alternative ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_altfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures \ref{fig:simu_h2}--\ref{fig:simu_alpha_large}).
+Estimates of both $p$ and $\alpha$ from SBayesS are often over-confident with small CIs that do not include the true simulated values (Figures S9 and S11).
+Generally, 95\% CIs for all three parameters ($h^2$, $p$ and $\alpha$) obtained from LDpred2-auto\_nojump contain the true simulated value (Figures S7--S12), therefore confirming the validity of these CIs; this is the preferred method we recommend using.
+Finally, we also investigate alternative ways of post-filtering chains in LDpred2-auto, for both prediction and inference (LDpred2-auto\_altfilter compared to LDpred2-auto\_nojump, Methods); results remain practically unchanged (Figures S1--S6).
 
@@
 
-Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures \ref{fig:simu_postp_calib} and \ref{fig:simu_h2_calib}).
+Then, LDpred2-auto can also infer per-variant probabilities of being causal and local per-block heritability estimates, which are well calibrated (Figures S13 and S14).
 We recall that calibrated per-variant probabilities of being causal (also known as posterior inclusion probabilities, PIPs) can be used for fine-mapping purposes  \cite[]{wang2020simple}.
-LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure \ref{fig:susie}).
-Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures \ref{fig:simu_r2} and \ref{fig:simu_r2_large}).
-CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when both $h^2$ and $N$ are small (Figures \ref{fig:simu10_r2} and \ref{fig:simu10_r2_large}).
+LDpred2-auto provides PIPs that are more calibrated than with e.g.\ SuSiE-RSS \cite[]{zou2022fine}, which we run assuming there are 10 causal variants per LD block by default (Figure S15).
+Finally, LDpred2-auto can also be used to reliably infer the predictive performance $r^2$ of its resulting polygenic score, directly from within the Gibbs sampler (Methods), even when power is low, and we show it works with results from SBayesS's Gibbs sampler as well (Figures S16 and S17).
+CIs for this $r^2$ estimate very often encompass the true simulated value, expect for LDpred2-auto\_jump when both $h^2$ and $N$ are small (Figures S18 and S19).
 
 We then run simulations with binary outcomes where the simulated continuous liabilities are transformed to binary outcomes using a threshold corresponding to the prevalence.
-Results are very similar as with the continuous phenotypes above (Figures \ref{fig:simu_h2_bin}--\ref{fig:simu_r2_bin}), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
+Results are very similar as with the continuous phenotypes above (Figures S20--S23), and are similar whether we use either a linear regression GWAS and the total sample size $N$, or a logistic regression GWAS and the effective sample size (i.e.\ $N_\text{eff} = 4 / (1 / N_\text{case} + 1 / N_\text{control})$).
 The main difference is that the $h^2$ and $r^2$ estimates must be transformed to the liability scale \cite[]{lee2011estimating}, where $K_\text{GWAS}=0.5$ should be used for transforming estimates of $h^2$ and $r^2$ when using $N_\text{eff}$ in inference methods \cite[]{grotzinger2021pervasive}.
@@
 
-Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure \ref{fig:ukbb_h2}).
+Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure S24).
 Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
-Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure \ref{fig:ukbb_h2_p}).
-Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure \ref{fig:ukbb_h2_alpha}).
-As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and \ref{fig:ukbb_r2_noMLE}).
+Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure S25).
+Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure S26).
+As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and S27).
 Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').
@@
 To investigate whether estimates from LDpred2-auto are robust to some misspecifications, we test using two alternative LD references (Methods).
-Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$) with LDpred2-auto, while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure \ref{fig:small_LD}).
-When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure \ref{fig:alt_LD}).
-Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can always get some chains not to diverge (and therefore get non-zero predictive performance) for the seven (three and four) phenotypes mentioned before when using the previous alternative LD references (Figure \ref{fig:compare_MLE}).
+Using a smaller number of individuals for computing the LD matrix results in a slightly overestimated $p$ and $h^2$ (and $r^2$) with LDpred2-auto, while the $\alpha$ estimate remains consistent, and the predictive performance in the test set remains mostly similar, except for three phenotypes for which none of the LDpred2-auto chains is usable (Figure S28).
+When using an LD reference from an alternative population (South Europe instead of North-West Europe), $p$, $h^2$, and $r^2$ are slightly overestimated as well, and a few phenotypes have lower predictive performance while there are four phenotypes for which none of the LDpred2-auto chains is usable (distinct from the previous three, Figure S29).
+Interestingly, using the previous approach (\texttt{use\_MLE = FALSE}) seems to provide more robust results, where we can always get some chains not to diverge (and therefore get non-zero predictive performance) for the seven (three and four) phenotypes mentioned before when using the previous alternative LD references (Figure S30).
 
 Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
-Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure \ref{fig:jumpsign}).
+Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31).
 $h^2$ estimates can also be slightly lower, while $\alpha$ estimates are broadly consistent.
 As for predictive performance $r^2$ (on the test set), they are similar, suggesting there is no problem of robustness here (when using the N.W.
-European LD reference) and a standard sampling can be used in this case (Figure \ref{fig:jumpsign}).
+European LD reference) and a standard sampling can be used in this case (Figure S31).
 
@@
 However, we do investigate alternative transformations here to decide which one should be preferred and to check how this impacts the inference and prediction from LDpred2-auto.
-We first compare to using raw (untransformed) phenotypes in Figure \ref{fig:notransfo}; estimates of $p$ and $\alpha$ are highly consistent.
+We first compare to using raw (untransformed) phenotypes in Figure S32; estimates of $p$ and $\alpha$ are highly consistent.
 However, $h^2$ estimates and predictive performance $r^2$ (in the test set) are generally larger with the log-transformation; it probably makes sense to transform these phenotypes.
-We then compare to using the rank-based inverse normal (RIN) transformation in Figure \ref{fig:RINT}; estimates for $p$ and $\alpha$ are also highly consistent.
+We then compare to using the rank-based inverse normal (RIN) transformation in Figure S33; estimates for $p$ and $\alpha$ are also highly consistent.
 Except for bilirubin and lipoprotein(a) concentration, higher $h^2$ estimates and predictive performance $r^2$ are generally obtained with the RIN-transformation than the log-transformation.
@@
 In particular, a much larger $h^2$ estimate is obtained for lipoprotein(a) concentration (0.508 [0.502, 0.514] instead of 0.324 [0.320, 0.329]), which is also reflected in a larger predictive performance ($r^2$ in the test set of 0.516 [0.508, 0.524] instead of 0.344 [0.335, 0.353]).
-Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure \ref{fig:ukbb_chains}). 
-However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure \ref{fig:ukbb_runtimes}), instead of 4--12 hours before.
+Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure S34). 
+However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.
 
@@
 In this section, we use the extended set of variants constructed here, HapMap3+, for which we define 431 independent LD blocks (Methods).
-We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure \ref{fig:ukbb_local_h2}.
+We compute local per-block $h^2$ estimates, and report the UKBB phenotypes for which one block contributes to at least 10\% of the total heritability of all blocks in Figure S36.
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
-We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure \ref{fig:protein_local_h2}).
+We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure S37).
 
-Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
+Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure S38).
 The outlier block explaining a much larger heritability contains the HLA region.
-Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
+Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure S39.
 Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
@@
 We identify the most significant HapMap3+ variant for height, rs2871960 on chromosome 3, and read all the variants within a 500kb distance that have a MAF larger than 0.005 and INFO score larger than 0.5 in the UK Biobank; there are 3881 such variants.
-Then, we perform a GWAS for these variants using the same 305,338 training individuals as used before (Figure \ref{manhattan_height}).
+Then, we perform a GWAS for these variants using the same 305,338 training individuals as used before (Figure S40).
 Finally, using the resulting GWAS summary statistics and an LD reference from the same set of individuals, we run both SuSiE-RSS and LDpred2-auto (\_nojump).
 We test two different values for L, the maximum number of causal variants in SuSiE-RSS (10 and 100), and three different values for the maximum value of the estimated $p$ (1, 0.01, and 0.001).
-Per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities, PIPs) are very similar with both methods, especially when restricting $p$ in LDpred2-auto, which is similar to what SuSiE-RSS assumes (more conservative).
+Per-variant posterior probabilities of being causal (also known as posterior inclusion probabilities, PIPs) are very similar with both methods, especially when restricting $p$ in LDpred2-auto, which is similar to what SuSiE-RSS assumes (more conservative, Figure S41).
 
@@
 These results are similar irrespective of whether \texttt{allow\_jump\_sign} is used or not, which is surprising to us.
-We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure \ref{fig:postp_height}).
+We also identify 1753 SNPs with a probability of being causal larger than 95\% (fine-mapping), which are spread over the entire genome (Figure S42).
 As for $\alpha$ estimates from LDpred2-auto, they remain consistent, with -0.71 [-0.75, -0.67], -0.74 [-0.76, -0.72], and -0.78 [-0.82, -0.76], respectively.
@@
 We perform this analysis using 50 non-flanking binary annotations from the baselineLD v2.2 model \cite[]{finucane2018heritability}.
-Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K, and of slightly larger magnitude with N=1.6M (Figure \ref{fig:enrichment}).
+Heritability enrichments that we obtain from LDpred2-auto results are rather modest, ranging from 0.7 to 2.5 with a GWAS sample size of N=305K, and of slightly smaller magnitude with N=100K, and of slightly larger magnitude with N=1.6M (Figure S43).
 
@@
 Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
-Results are presented in Figure \ref{fig:sumstats}.
+Results are presented in Figure S44.
 In terms of predictive performance, using the HapMap3+ variants provides equal or better predictive performance compared to using the HapMap3 variants, except for type 1 diabetes (T1D); therefore it seems more useful to use this new set of variants for larger sample sizes.
 Using the extended model and sampling scheme (with option \texttt{use\_MLE}) provides equal or better predictive performance except for vitaminD, but proves to be less robust, especially when using \texttt{coef\_shrink} close to 1 (low or no regularization of the LD matrix).
-The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure \ref{fig:ukbb_r2_vs}).
+The 3-parameter model also generally provides better predictive performance in the UK Biobank analyses, where power and robustness is often not an issue (Figure S45).
 Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.

### 2023-10-03 code/simu.R
rename to altfilter

@@ -162,7 +162,8 @@ grid %>%
   relocate(n_keep_infer, n_keep_pred, .after = use_mle) %>%
   ungroup() %>%
   mutate(Method = factor(Method, levels = Method[1:6],
-                         labels = sub("LDpred2_", "LDpred2-auto_", Method[1:6]))) %>%
+                         labels = sub("LDpred2_", "LDpred2-auto_",
+                                      sub("newfilter", "altfilter", Method[1:6])))) %>%
   select(-starts_with("all"), -postp) %>%
   tidyr::unnest_wider(where(is.list), names_sep = "_") %>%
   mutate(across(r2_1:alpha_est_3, ~ signif(., digits = 4))) %>%
@@ -183,7 +184,8 @@ grid2 <- grid %>%
          n_keep_infer = purrr::map_dbl(all_h2, ~ `if`(is.null(.), 0, NCOL(.))),
          n_keep_pred  = purrr::map_dbl(all_r2, ~ `if`(is.null(.), 0, NCOL(.) / 500 * 20)),
          Method = factor(Method, levels = Method[1:6],
-                         labels = sub("LDpred2_", "LDpred2-auto_", Method[1:6])),
+                         labels = sub("LDpred2_", "LDpred2-auto_",
+                                      sub("newfilter", "altfilter", Method[1:6]))),
          p = as.factor(p)) %>%
   relocate(n_keep_infer, n_keep_pred, .after = Method) %>%
   print(n = 20)

### 2023-10-03 paper/paper-infer.tex
AJHG citation style

@@
 \onehalfspacing
-\usepackage[authoryear]{natbib}
-%\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
-These methods (non-exhaustively) include GCTA ($h^2$, \cite{yang2011gcta}), BOLT-REML ($h^2$ and $p$, \cite{loh2015contrasting}), LD Score regression ($h^2$, \cite{bulik2015ld}), FINEMAP (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping, \cite{benner2016finemap}), HESS (local $h^2$, \cite{shi2016contrasting}), LDAK-SumHer ($h^2$, \cite{speed2019sumher}), S-LD4M ($p$, \cite{oconnor2019extreme}), GRM-MAF-LD ($\alpha$, \cite{schoech2019quantification}), SuSiE (PIPs, \cite{wang2020simple}), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$, \cite{zeng2021widespread}), and BEAVR (local $p$, \cite{johnson2021estimation}).
+These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
 
-As previously shown by \cite{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
+As previously shown by \citet{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
@@
 	\caption{Overview of what LDpred2-auto can now provide.
-For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large,ding2022polygenic}.
+For individual CIs of polygenic scores, please refer to the work of \citet{ding2022large,ding2022polygenic}.
 HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
@@
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
-These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in section S3 of \citet{bycroft2018uk}).
+These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in \citet{bycroft2018uk}).
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
@@
 
-We construct and use the same phenotypes as in \cite{prive2021high}.
+We construct and use the same phenotypes as in \citet{prive2021high}.
 About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
@@
 
-We use the LD matrix with independent LD blocks computed in \cite{prive2021identifying}.
-We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \cite{prive2021high} (``hm3\_altpop'').
-We apply the optimal algorithm developed in \cite{prive2021optimal} to obtain independent LD blocks, as recommended in \cite{prive2021identifying}.
+We use the LD matrix with independent LD blocks computed in \citet{prive2021identifying}.
+We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \citet{prive2021high} (``hm3\_altpop'').
+We apply the optimal algorithm developed in \citet{prive2021optimal} to obtain independent LD blocks, as recommended in \citet{prive2021identifying}.
 We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').
@@
 Note that we only allow non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \cite{prive2021optimal} to obtain 431 independent LD blocks.
+This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citet{prive2021optimal} to obtain 431 independent LD blocks.
 Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
@@
 This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
-Note that \cite{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
+Note that \citet{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
 Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.
@@
 Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
-Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \cite{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods).
+Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \citet{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods).
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
@@
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}).
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}).
 
@@
 
-Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \cite{prive2021identifying}.
+Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}.
 Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31).
@@
 Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure S34). 
-However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \cite{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.
+However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \citet{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.
 
@@
 Here we run three LDpred2-auto models for height, one from the same 305K training UKBB individuals used before (with available height, out of 306K), one based on 100K UKBB individuals (as a random subset of the previous 305K), and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries \cite[]{yengo2022saturated}.
-We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \cite{prive2021using}, and find that 81.9\% are from N.W.
+We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \citet{prive2021using}, and find that 81.9\% are from N.W.
 Europe, 9.5\% from E.
@@
 A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these do not include UKBB individuals.
-Quality control of these summary statistics is performed as described in \cite{prive2021identifying}.
+Quality control of these summary statistics is performed as described in \citet{prive2021identifying}.
 We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}).
 Because of the increased mismatch between external GWAS summary statistics and the LD reference we use here (compared to in simulations and UKBB analyses), we also explore multiple values for parameter \texttt{coef\_shrink}, which is a multiplicative coefficient for shrinking/regularizing off-diagonal elements of the LD matrix in LDpred2-auto.
-Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite{fry2017comparison,van2022reweighting}.
+Note that, to transform $r^2$ and $h^2$ estimates to the liability scale (except for vitamin D, which is a continuous trait), we use the prevalence in the UK Biobank as the population prevalence, which may be slightly biased \cite[]{fry2017comparison,van2022reweighting}.
 Results are presented in Figure S44.
@@
 Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
-However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \cite{wen2010using}), which is surprising to us.
+However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \citet{wen2010using}), which is surprising to us.
 As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
@@
 Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
-Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$, \cite{prive2020ldpred2}).
+Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
 We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
-However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC, \cite{finucane2018heritability}).
-For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
+However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC)\cite{finucane2018heritability}.
+For example, \citet{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
 This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
@@
 Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% in UKBB analyses here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
-However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
+However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \citet{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
 This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more localized set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
@@
 
-\bibliographystyle{natbib}
+\bibliographystyle{ajhg}
 \bibliography{refs}

### 2023-10-03 paper/paper-infer.tex
some formating

@@
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
-These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
+These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
 
@@
 \section*{Materials and Methods}
+
+We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
 
@@
 Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
-Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (resp.\ 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).
+Indeed, from the 11.5M variants we aimed at tagging, 82.1\% (respectively 69.1\%), 80.0\% (resp.\ 66.7\%), 79.3\% (resp.\ 66.1\%), 75.4\% (resp.\ 65.3\%), and 66.6\% (resp.\ 48.6\%) are tagged at $r^2 > 50\%$ (resp.\ $r^2 > 80\%$) by at least one HapMap3+ variant, respectively in North-West Europeans, Middle Easterns, South Asians, East Asians, and West Africans (results extrapolated from variants in chromosome 22).
 
@@
 	\centerline{\includegraphics[width=0.8\textwidth]{ukbb_r2_MLE}}
-	\caption{Inferred predictive performance $r^2$ from the Gibbs sampler of LDpred2-auto versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
+	\caption{Inferred predictive performance $r^2$ in the UK Biobank.
+This compares the $r^2$ inferred from the Gibbs sampler of LDpred2-auto (Methods) versus the ones obtained in the test set, for all 248 phenotypes defined from the UK Biobank.
 These are stratified by the polygenicity estimated from LDpred2-auto.
@@
 For lipoprotein(a) concentration, ``red hair'' and ``disorders of iron metabolism'' (phecode 275.1), almost all heritability comes from one LD block only.
-We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (resp.\ 8) of them have at least 50\% (resp.\ 80\%) of their heritability explained by a single block (Figure S37).
+We also perform the same analysis with external GWAS summary statistics for 90 cardiovascular proteins \cite[]{folkersen2020genomic}; 22 (respectively 8) of them have at least 50\% (respectively 80\%) of their heritability explained by a single block (Figure S37).
 
@@
 
-\section*{Funding}
-
 F.P., C.A.\ and B.J.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to Prof.\ John McGrath), the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R102-A9118, R155-2014-1724, R248-2017-2003), and a Lundbeck Foundation Fellowship (R335-2019-2339 to B.J.V.).
@@
 The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
-Descriptions of UK Biobank phenotypes used here are available at \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}.
-Simulation and UKBB results are available at \url{https://github.com/privefl/paper-infer/tree/main/results}.
-LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper are available at \url{https://doi.org/10.6084/m9.figshare.21305061}.
+UK Biobank received ethical approval from the NHS National Research Ethics Service North West (11/NW/0382). The present analyses were conducted under UK Biobank data application number 58024.
 All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/main/code}.
-We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
-The latest version of R package bigsnpr can be installed from GitHub, and a recent enough version can be installed from CRAN.
+
+\section*{Web Resources}
+
+\noindent Descriptions of UK Biobank phenotypes used here, \url{https://github.com/privefl/paper-infer/blob/main/phenotype-description.tsv}
+
+\noindent Simulation and UKBB results from this study,  \url{https://github.com/privefl/paper-infer/tree/main/results}
+
+\noindent LD matrices for HapMap3+ variants computed from the N.W. European UKBB data used in this paper, \url{https://doi.org/10.6084/m9.figshare.21305061}
+
+\noindent A tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}
 

### 2023-10-03 paper/paper-infer.tex
add reponses to the text

@@
 Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$.
-These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
+These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher,speed2020evaluating} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
+Not all methods have the same modeling assumptions; for example, LDAK-SumHer assumes a different prior than SBayesS and LDpred2-auto, and does not estimate all the same parameters.
+It can estimate the SNP heritability and $\alpha$.
+However, it cannot estimate the polygenicity nor the per-variant probabilities of being causal (since it inherently assumes an infinitesimal model, i.e.\ $p = 1$).
+Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the new formula we propose in this paper.
 
@@
 Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure S25).
-Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure S26).
+Most phenotypes have an estimated $\alpha$ between -1.0 and -0.3 with a mode at -0.65 (Figure S26).
 As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and S27).
@@
 Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
-We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
+We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesR where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
+In that paper, they found consistent results with BayesR, although the original BayesR publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}.
+In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).
+
 However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC)\cite{finucane2018heritability}.
@@
 LDpred2-auto can also miss some causal variants when they are poorly tagged by the set of variants used.
-Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
+Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation.
+LDpred2-auto runtime currently depends on the number of causal variants (or, equivalently, the polygenicity).
+When the polygenicity is close to 0, it runs linearly with the number of variants.
+When the polygenicity is close to 1, it runs quadratically.
+However, the polygenicity is almost always lower than 0.1 for human traits and diseases (cf.\ UKBB results here), which makes running LDpred2-auto efficient (we report runtimes always under one hour, even with HapMap3+ variants).
+Currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
 For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').
@@
 
-\noindent A tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}
+\noindent The official tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}
 

### 2023-10-03 paper/paper-infer.tex
typo

@@
 Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
-We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesR where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
-In that paper, they found consistent results with BayesR, although the original BayesR publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}.
+We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesS where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
+In that paper, they found consistent results with BayesS, although the original BayesS publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}.
 In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).

### 2023-10-09 paper/paper-infer.tex
rm 'new' for AJHG

@@
 \onehalfspacing
-\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage[numbers,super]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 We also show that LDpred2-auto provides per-variant probabilities of being causal that are well calibrated, and can therefore be used for fine-mapping purposes.
-We also derive a new formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto.
+We also introduce a formula to infer the out-of-sample predictive performance $r^2$ of the resulting PGS directly from the Gibbs sampler of LDpred2-auto.
 Finally, we extend the set of HapMap3 variants recommended to use with LDpred2 with 37\% more variants to improve the coverage of this set, and show that this new set of variants captures 12\% more heritability and provides 6\% more predictive performance, on average, in UK Biobank analyses.
@@
 However, it cannot estimate the polygenicity nor the per-variant probabilities of being causal (since it inherently assumes an infinitesimal model, i.e.\ $p = 1$).
-Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the new formula we propose in this paper.
+Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the formula we propose in this paper.
 
-As previously shown by \citet{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
+As previously shown by \citep{daetwyler2008accuracy}, $h^2$ and $p$ can also be used to determine how well we can predict a phenotype from using genetic variants alone, with $r^2_\text{max} = h^2 \cdot r^2_{g\hat{g}} = \dfrac{h^2}{1 + (1 - r^2_\text{max}) \frac{M p}{N h^2}}$, where $r^2_\text{max}$ and $r^2_{g\hat{g}}$ are the maximum achievable squared correlations between the genetic predictor and respectively the phenotype and its genetic component, $N$ is the sample size, $M$ is the number of variants, so $M p$ is the number of causal variants.
 Such genetic predictors are called polygenic scores (PGS), and are getting closer to being included as part of existing clinical risk models for diseases \cite[]{torkamani2018personal,lambert2019towards,kumuthini2022clinical}.
@@
 	\caption{Overview of what LDpred2-auto can now provide.
-For individual CIs of polygenic scores, please refer to the work of \citet{ding2022large,ding2022polygenic}.
+For individual CIs of polygenic scores, please refer to the work of \citep{ding2022large,ding2022polygenic}.
 HapMap3+ is the new extended set of 1,444,196 SNPs we introduce here (Methods), and now recommend using for LDpred2 when power is sufficient (i.e.\ with a large sample size, large $h^2$, and/or small $p$).
@@
 We restrict individuals to the ones used for computing the principal components (PCs) in the UK Biobank (field 22020).
-These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in \citet{bycroft2018uk}).
+These individuals are unrelated and have passed some quality control such as removing samples with a missing rate on autosomes larger than 0.02, having a mismatch between inferred sex and self-reported sex, and outliers based on heterozygosity (more details can be found in \citep{bycroft2018uk}).
 To get a set of genetically homogeneous individuals, we compute a robust Mahalanobis distance based on the first 16 PCs (field 22009) and further restrict individuals to those within a log-distance of 4.5 \cite[]{prive2020efficient}.
@@
 
-We construct and use the same phenotypes as in \citet{prive2021high}.
+We construct and use the same phenotypes as in \citep{prive2021high}.
 About half of these consists of phecodes mapped from ICD10 and ICD9 codes using R package PheWAS \cite[]{carroll2014r,wu2019mapping}.
@@
 
-We use the LD matrix with independent LD blocks computed in \citet{prive2021identifying}.
-We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \citet{prive2021high} (``hm3\_altpop'').
-We apply the optimal algorithm developed in \citet{prive2021optimal} to obtain independent LD blocks, as recommended in \citet{prive2021identifying}.
+We use the LD matrix with independent LD blocks computed in \citep{prive2021identifying}.
+We design two other LD matrices: one using a smaller random subset of 2000 individuals from the previously selected ones (which we call ``hm3\_small''), and one based on 10,000 individuals from around South Europe by using the ``Italy'' center defined in \citep{prive2021high} (``hm3\_altpop'').
+We apply the optimal algorithm developed in \citep{prive2021optimal} to obtain independent LD blocks, as recommended in \citep{prive2021identifying}.
 We finally define a fourth LD reference by extending the set of HapMap3 variants (see next Methods section) and using 20,000 individuals from the previously selected ones (``hm3\_plus'').
@@
 Note that we only allow non-ambiguous SNPs to be included.
-This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citet{prive2021optimal} to obtain 431 independent LD blocks.
+This results in an extended set of 1,444,196 SNPs, of which we compute the LD matrix (within a 3 cM window) and apply the optimal algorithm developed in \citep{prive2021optimal} to obtain 431 independent LD blocks.
 Since we use individuals of diverse ancestries for computing the pairwise variant correlations used for constructing this extended set of variants, we expect this new set of variants to be beneficial across diverse ancestries.
@@
 This is what we use for SBayesS here, and also as an alternative means for post-filtering chains for prediction (cf.\ next Methods section).
-Note that \citet{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
+Note that \citep{ding2022large} investigated autocorrelation in LDpred2(-grid) and showed that it decays very quickly.
 Therefore picking two LDpred2-auto samples that are 100 iterations apart should be more than enough to ensure quasi-independence of these samples.
 
-In this paper, we check this new $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
+In this paper, we check this $r^2$ approximation using extensive simulations (across many genetic architectures) and real data analyses (across 248 different phenotypes).
 These are compared to the partial-$r^2$ (on individual-level data from a separate test set).
@@
 Then, we run a GWAS to obtain summary statistics using $N$ individuals (either the 200,000 dedicated to this, or a random subset of 20,000), using fast linear regressions implemented in \texttt{big\_univLinReg} from R package bigstatsr \cite[]{prive2017efficient}.
-Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \citet{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods).
+Finally, we run LDpred2-auto with and without the option \texttt{allow\_jump\_sign}, which was proposed in \citep{prive2021identifying} for robustness (when disabled, it prevents effect sizes from changing sign without going through 0 first), and with and without the extended model including a third parameter $\alpha$ (using option \texttt{use\_MLE}, Methods).
 LDpred2-auto is run with 50 Gibbs sampler chains with different starting values for $p$ (from 0.0001 to 0.2, equally spaced on a log scale).
@@
 We first use the set of 1,054,330 HapMap3 variants recommended to use for LDpred2 \cite[]{prive2020ldpred2}.
-Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}).
+Here, if not otherwise specified, we use options \texttt{use\_MLE = TRUE} (i.e.\ the extended 3-parameter model and sampling scheme) and \texttt{allow\_jump\_sign = FALSE} (when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citep{prive2021identifying}).
 
@@
 
-Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citet{prive2021identifying}.
+Then, we investigate the effect of disabling the LDpred2-auto parameter \texttt{allow\_jump\_sign} on the estimates from LDpred2-auto; when disabled, this prevents effect sizes from changing sign without going through 0 first and has been proposed for extra robustness in \citep{prive2021identifying}.
 Consistent with simulations, $p$ estimates from LDpred2-auto are conservatively lower than when allowing effects to ``jump'' sign (i.e.\ using a standard sampling, Figure S31).
@@
 Interestingly, when using this extended set of HapMap3 variants, more chains are kept on average, which is a sign of better convergence of the models (Figure S34). 
-However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \citet{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.
+However, running LDpred2 with this extended set of variants takes around 50\% more time; yet, we remind the reader that LDpred2 has been made much faster in \citep{prive2021identifying}, and now runs in less than one hour for 50 chains parallelized over 13 cores (Figure S35), instead of 4--12 hours before.
 
@@
 Here we run three LDpred2-auto models for height, one from the same 305K training UKBB individuals used before (with available height, out of 306K), one based on 100K UKBB individuals (as a random subset of the previous 305K), and one from a large GWAS meta-analysis of 1.6M individuals of European genetic ancestries \cite[]{yengo2022saturated}.
-We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \citet{prive2021using}, and find that 81.9\% are from N.W.
+We first infer the genetic ancestry proportions of individuals included in the meta-analysis using the method proposed in \citep{prive2021using}, and find that 81.9\% are from N.W.
 Europe, 9.5\% from E.
@@
 A description of the eight external GWAS summary statistics used is provided in Table \ref{tab:sumstats}; these do not include UKBB individuals.
-Quality control of these summary statistics is performed as described in \citet{prive2021identifying}.
+Quality control of these summary statistics is performed as described in \citep{prive2021identifying}.
 We run LDpred2-auto using either the HapMap3 or HapMap3+ variants, with either the extended or previous model and sampling (via parameter \texttt{use\_MLE}).
@@
 Maximum $r^2$ are achieved at different LD regularization coefficients \texttt{coef\_shrink} across phenotypes, reflecting possible substantial mismatches between the GWAS summary statistics and LD reference used.
-However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \citet{wen2010using}), which is surprising to us.
+However, results are virtually unchanged when regularizing the LD matrix (``hm3\_plus\_regul'') by multiplying correlations between variants $i$ and $j$ by $\exp(-0.5 \cdot d_{i,j})$, where $d_{i,j}$ is the distance in cM between the two variants (similarly to as proposed in \citep{wen2010using}), which is surprising to us.
 As for estimates of $h^2$ and $r^2$ (inferred from the Gibbs sampler), they tend to increase with smaller values of \texttt{coef\_shrink}. 
@@
 		\hline
-		Asthma & \citet{demenais2018multiancestry} & ~~67,341.2 & ~~2,001,280 & ~~~946,092 \\
-		Breast cancer (BRCA) & \citet{michailidou2017association} & 254,862.6 & 11,792,542 & 1,411,710 \\
-		Coronary artery disease (CAD) & \citet{nikpay2015comprehensive} & 129,014.3 & ~~9,455,778 & 1,325,052 \\
-		Depression (MDD) [without UKBB] & \citet{wray2018genome} & 168,040.2 & ~~9,874,289 & 1,314,499 \\
-		Prostate cancer (PRCA) & \citet{schumacher2018association} & 135,316.1 & 20,370,946 & 1,411,710  \\
-		Type 1 diabetes (T1D) & \citet{censin2017childhood} & ~~13,497.6  & ~~8,996,866 & 1,127,489  \\
-		Type 2 diabetes (T2D) & \citet{scott2017expanded} & ~~72,143.0  & 12,056,346 & 1,408,283  \\
-		Vitamin D & \citet{jiang2018genome} & ~~79,366~~~ & ~~2,579,296 & 1,028,171 \\
+		Asthma & \citep{demenais2018multiancestry} & ~~67,341.2 & ~~2,001,280 & ~~~946,092 \\
+		Breast cancer (BRCA) & \citep{michailidou2017association} & 254,862.6 & 11,792,542 & 1,411,710 \\
+		Coronary artery disease (CAD) & \citep{nikpay2015comprehensive} & 129,014.3 & ~~9,455,778 & 1,325,052 \\
+		Depression (MDD) [without UKBB] & \citep{wray2018genome} & 168,040.2 & ~~9,874,289 & 1,314,499 \\
+		Prostate cancer (PRCA) & \citep{schumacher2018association} & 135,316.1 & 20,370,946 & 1,411,710  \\
+		Type 1 diabetes (T1D) & \citep{censin2017childhood} & ~~13,497.6  & ~~8,996,866 & 1,127,489  \\
+		Type 2 diabetes (T2D) & \citep{scott2017expanded} & ~~72,143.0  & 12,056,346 & 1,408,283  \\
+		Vitamin D & \citep{jiang2018genome} & ~~79,366~~~ & ~~2,579,296 & 1,028,171 \\
 		\hline
@@
 We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
-We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
+We also introduce a way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
 Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
@@
 In that paper, they found consistent results with BayesS, although the original BayesS publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}.
-In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).
+In \citep{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).
 
 However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC)\cite{finucane2018heritability}.
-For example, \citet{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
+For example, \citep{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x.
 This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
@@
 Using the new HapMap3+ set also improves predictive performance by an average of 6.1\% in UKBB analyses here, and particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
-However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \citet{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
+However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \citep{prive2019efficient} on the UKBB individual-level data while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
 This means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more localized set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.
@@
 Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
-Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
+Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue (e.g.\ without substantial misspecifications such as an ancestry mismatch between the GWAS and LD panels).
 We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates.
@@
 
-\vspace*{3em}
-%\clearpage
+%\vspace*{3em}
+\clearpage
 

### 2023-10-09 paper/paper-infer-supp.tex
AJHG citation style
some formating
rm 'new' for AJHG
AJHG style

@@
 \onehalfspacing
+\pagenumbering{gobble}
 \usepackage[numbers,super]{natbib}
@@
 		The 95\% confidence intervals for the LDpred2-auto and SBayesS estimates are obtained from the 2.5\% and 97.5\% quantiles of all the $h^2$ estimates from the iterations (after burn-in) of the chains kept (note that only one chain is used and kept in SBayesS).
-The 95\% confidence interval for the LD Score (LDSc) regression estimate is obtained as $\pm$1.96 of its standard error.
+The 95\% confidence interval for the LD Score regression (LDSC) estimate is obtained as $\pm$1.96 of its standard error.
 Colors for LDpred2-auto models represent the number of chains kept (out of 50).
@@
 	\caption{Per-variant probabilities of being causal for height.
-These are provided by LDpred2-auto when using the external GWAS summary statistics of 1.6M European individuals from \citet{yengo2022saturated}.
+These are provided by LDpred2-auto when using the external GWAS summary statistics of 1.6M European individuals from \citep{yengo2022saturated}.
 \label{fig:postp_height}}

### 2023-10-09 paper/paper-infer.tex
AJHG style

@@
 \onehalfspacing
-\usepackage[numbers,super]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
+\usepackage[numbers,super]{natbib} %\bibpunct{(}{)}{;}{author-year}{}{,}
 
@@
 We have extensively used R packages bigstatsr and bigsnpr \cite[]{prive2017efficient} for analyzing large genetic data, packages from the future framework \cite[]{bengtsson2020unifying} for easy scheduling and parallelization of analyses on the HPC cluster, and packages from the tidyverse suite \cite[]{wickham2019welcome} for shaping and visualizing results.
+We have extensively used the UK Biobank data\cite[]{bycroft2018uk}, which is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
+The UK Biobank received ethical approval from the NHS National Research Ethics Service North West (11/NW/0382). The present analyses were conducted under UK Biobank data application number 58024.
 
@@
 When using this 3-parameter model in LDpred2-auto, in order to add some sampling to these two parameters, we first sample causal variants with replacement before computing the maximum likelihood estimators.
-This maximum likelihood estimation (MLE) is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
+This maximum likelihood estimation (MLE) is implemented using R package roptim (Web Resources), and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration of the Gibbs sampler.
 Note that we still estimate $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, but that $h^2$ is not used in the variance of sampled effect sizes anymore (Equation \eqref{eq:new_model}). 
@@
 
-%\clearpage
-\vspace*{3em}
+%\vspace*{3em}
+\clearpage
 
@@
 
-\section*{Code and data availability}
+\section*{Code availability}
 
-The UK Biobank data is available through a procedure described at \url{https://www.ukbiobank.ac.uk/using-the-resource/}. 
-UK Biobank received ethical approval from the NHS National Research Ethics Service North West (11/NW/0382). The present analyses were conducted under UK Biobank data application number 58024.
 All code used for this paper is available at \url{https://github.com/privefl/paper-infer/tree/main/code}.
@@
 
+\noindent roptim: an R package for general purpose optimization with C++, \url{https://cran.r-project.org/package=roptim}
+
 %\noindent[TODO: TABLE EXPORTING RESULTS (ESTIMATES)]
@@
 
-%\vspace*{3em}
-\clearpage
+\vspace*{3em}
+%\clearpage
 

### 2023-10-15 paper/paper-infer.tex
minor changes

@@
 Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; interestingly, these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
-To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a small negative correlation of -5.5\% with the probabilities of being causal.
-To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a small correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.
+To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have only a small correlation of 5.9\% with the probabilities of being causal.
+To verify that this does not correspond to regions of low LD, we compute LD scores; the median probabilities of being causal have a correlation of 32.0\% with the LD scores and of 26.4\% with the log of LD scores.
+Therefore the posterior probabilities of being causal obtained from LDpred2-auto tend to slightly increase with LD scores.
 

### 2023-10-15 code/run-ldpred2-cardio-proteins.R

@@ -1,7 +1,7 @@
 library(dplyr)
 urls <- readLines("https://zenodo.org/record/2615265#.YvuAdBxBy01") %>%
-  grep(value = TRUE, pattern = "type=\"application/gzip\" href=\"https://zenodo.org/record/2615265/files/.+\\.txt\\.gz\">") %>%
-  sub(pattern = ".+href=\"(https://zenodo.org/record/2615265/files/.+\\.txt\\.gz)\">", replacement = "\\1")
+  grep(value = TRUE, pattern = "type=\"application/gzip\" href=\"https://zenodo.org/records/2615265/files/.+\\.txt\\.gz\">") %>%
+  sub(pattern = ".+href=\"(https://zenodo.org/records/2615265/files/.+\\.txt\\.gz)\">", replacement = "\\1")
 
 bigassertr::assert_dir("cardio_proteins")
 
@@ -146,11 +146,14 @@ with(res_local_h2, sum((max_local_h2 / h2) > 0.5))  # 22
 with(res_local_h2, sum((max_local_h2 / h2) > 0.8))  # 8
 
 library(ggplot2)
+library(latex2exp)
 ggplot(tidyr::pivot_longer(res_local_h2, c(h2, max_local_h2)),
        aes(protein, value, fill = name, color = name)) +
   theme_bw(13) +
-  scale_color_manual(values = c("#0072B2FF", "#00000033")) +  # "#E69F0000"
-  scale_fill_manual(values = c("#0072B200", "#E69F0099")) +
+  scale_color_manual(values = c("#0072B2FF", "#00000033"),
+                     labels = c(TeX("$h^2$ (sum of all 431 blocks)"), TeX("maximum block-$h^2$"))) +
+  scale_fill_manual(values = c("#0072B200", "#E69F0099"),
+                    labels = c(TeX("$h^2$ (sum of all 431 blocks)"), TeX("maximum block-$h^2$"))) +
   geom_col(position = "identity") +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
         legend.position = "top") +

### 2023-10-15 code/run-ldpred2-sumstats.R

@@ -157,9 +157,12 @@ grid$res <- furrr::future_pmap(grid[1:6], function(pheno, pheno_file, set, mle,
 library(dplyr)
 library(ggplot2)
 theme_set(theme_bw(14))
+library(latex2exp)
 
 grid2 <- grid %>%
-  tidyr::unnest_wider("res") %>%
+  select(-res_file) %>%
+  tidyr::unnest_wider("res", names_sep = "_") %>%
+  rename_with(~ sub("res_", "", .)) %>%
   rowwise() %>%
   mutate(n_keep = `if`(is.null(all_h2), 0, ncol(all_h2)))
 
@@ -195,16 +198,21 @@ for (PHENO in unique(grid3$pheno)) {
 
   grid3 %>%
     filter(pheno == PHENO) %>%
-    mutate(mle = ifelse(mle, "Yes", "No")) %>%
+    mutate(mle = ifelse(mle, "Yes", "No"),
+           name = factor(name, levels = c(
+             "alpha_est", "h2_est", "h2_ldsc_est", "p_est", "r2", "r2_est"),
+             labels = sapply(c("$alpha$ (LDpred2-auto)", "$h^2$ (LDpred2-auto)",
+                               "$h^2$ (LDSC)", "$p$ (LDpred2-auto)",
+                               "$r^2$ (in test set)", "$r^2$ (LDpred2-auto)"), TeX))) %>%
     ggplot(aes(coef_shrink, value_1, color = set, shape = mle, linetype = mle)) +
     bigstatsr::theme_bigstatsr(0.65) +
-    facet_wrap(~ pheno + name, scales = "free_y") +
+    facet_wrap(~ pheno + name, scales = "free_y", labeller = label_parsed) +
     geom_point(size = 2) +
     geom_line() +
     theme(legend.position = c(0.85, 0.2), legend.key.width = unit(2, "line")) +
     scale_shape_manual(values = c(3, 16, 17)) +
     scale_color_manual(values = c('#E69F00', '#56B4E9', '#CC79A7')) +
-    labs(y = "Estimate  (+ 95% CI of the estimate)", x = "LD shrinkage factor",
+    labs(y = "Estimate", x = "LD shrinkage factor",
          color = "Set", shape = "Use MLE?", linetype = "Use MLE?")
 
   ggsave(paste0("figures/ldpred2_", PHENO, ".pdf"), width = 9, height = 6)

### 2023-10-15 code/run-ldpred2-ukbb.R

@@ -196,6 +196,7 @@ c(nrow(grid), nrow(res_grid))
 library(dplyr)
 library(ggplot2)
 theme_set(theme_bw(14))
+library(latex2exp)
 
 grid2 <- bind_cols(grid, res_grid) %>%
   filter(!grepl("rint_", pheno_file), !grepl("raw_", pheno_file), !jump_sign)
@@ -228,7 +229,7 @@ grid2 %>%
   select(-c(pheno_file, postp, res_file, time, local_h2)) %>%
   relocate(r2, .after = r2_est) %>%
   tidyr::unnest_wider(h2_ldsc_est:r2, names_sep = "_") %>%
-  mutate(across(h2_ldsc_est_1:r2_3, signif)) %>%
+  mutate(across(h2_ldsc_est_1:r2_3, ~ signif(., digits = 4))) %>%
   print() %>%
   bigreadr::fwrite2("results_infer_ukbb.csv")
 
@@ -263,7 +264,8 @@ grid3_hm3 %>%
   facet_wrap(~ use_mle + h2, scales = "free", labeller = label_both, nrow = 2) +
   geom_point(size = 1.5) +
   theme(legend.key.height = unit(50, "pt")) +
-  labs(y = "Inferred r2", x = "r2 in test set", color = "# chains")
+  labs(y = TeX("Inferred $r^2$ from LDpred2-auto"),
+       x = TeX("$r^2$ in test set"), color = "# chains")
 # ggsave("figures/ukbb_r2_MLE.pdf", width = 8, height = 7)
 # ggsave("figures/ukbb_r2_noMLE.pdf", width = 8, height = 7)
 
@@ -280,7 +282,8 @@ grid2 %>%
   geom_errorbar(aes(xmin = TRUE_2, xmax = TRUE_3), color = "chartreuse3",,
                 position = position_dodge(width = 0.9), width = 0) +
   geom_point(size = 1.5) +
-  labs(x = "r2 with 3-param LDpred2-auto", y = "r2 with 2-param LDpred2-auto")
+  labs(x = TeX("$r^2$ (in test set) with 3-parameter LDpred2-auto"),
+       y = TeX("$r^2$ (in test set) with 2-parameter LDpred2-auto"))
 # ggsave("figures/ukbb_r2_vs.pdf", width = 6, height = 6)
 
 
@@ -315,8 +318,8 @@ grid3_hm3 %>%
   geom_point(size = 1.5) +
   theme(legend.key.height = unit(25, "pt")) +
[…]

### 2023-10-15 code/simu-binary.R

@@ -52,16 +52,16 @@ grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num) {
   # Run methods
   source('code/simu-methods.R', local = TRUE)
   res_lin <- list(
-    LDpred2_noMLE = runonce::save_run(
+    `LDpred2-auto_noMLE` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = FALSE, coef_shrink = 0.95),
       file = paste0("results_simu_binary/LDpred2_noMLE_lin_", params, ".rds")),
-    LDpred2_MLE = runonce::save_run(
+    `LDpred2-auto_MLE` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = TRUE,  coef_shrink = 0.95),
       file = paste0("results_simu_binary/LDpred2_nojump_lin_", params, ".rds")),
     SBayesS = runonce::save_run(
       run_sbayess(),
       file = paste0("results_simu_binary/SBayesS_lin_", params, ".rds")),
-    LDSc = list(all_h2 = ldsc)
+    LDSC = list(all_h2 = ldsc)
   )
 
 
@@ -80,16 +80,16 @@ grid$res <- furrr::future_pmap(grid, function(p, h2, alpha, K, num) {
 
   # Run methods
   res_log <- list(
-    LDpred2_noMLE = runonce::save_run(
+    `LDpred2-auto_noMLE` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = FALSE, coef_shrink = 0.95),
       file = paste0("results_simu_binary/LDpred2_noMLE_log_", params, ".rds")),
-    LDpred2_MLE = runonce::save_run(
+    `LDpred2-auto_MLE` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = TRUE,  coef_shrink = 0.95),
       file = paste0("results_simu_binary/LDpred2_nojump_log_", params, ".rds")),
     SBayesS = runonce::save_run(
       run_sbayess(),
       file = paste0("results_simu_binary/SBayesS_log_", params, ".rds")),
-    LDSc = list(all_h2 = ldsc)
+    LDSC = list(all_h2 = ldsc)
   )
 
 
@@ -111,21 +111,20 @@ grid %>%
   mutate(
     r2 = list(unname(r2)),
[…]

### 2023-10-15 code/simu.R

@@ -115,22 +115,22 @@ grid$res <- furrr::future_pmap(grid[1:5], function(p, h2, alpha, N, num) {
   # Run methods
   source('code/simu-methods.R', local = TRUE)
   list(
-    LDpred2_newfilter = runonce::save_run(
+    `LDpred2-auto_altfilter` = runonce::save_run(
       run_ldpred2_new(jump_sign = FALSE, use_mle = TRUE, coef_shrink = 0.95),
       file = paste0("results_simu/LDpred2_newfilter_", params, ".rds")),
-    LDpred2_noMLE = runonce::save_run(
+    `LDpred2-auto_noMLE` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = FALSE, coef_shrink = 0.95),
       file = paste0("results_simu/LDpred2_noMLE_", params, ".rds")),
-    LDpred2_nojump = runonce::save_run(
+    `LDpred2-auto_nojump` = runonce::save_run(
       run_ldpred2(jump_sign = FALSE, use_mle = TRUE,  coef_shrink = 0.95),
       file = paste0("results_simu/LDpred2_nojump_", params, ".rds")),
-    LDpred2_jump = runonce::save_run(
+    `LDpred2-auto_jump` = runonce::save_run(
       run_ldpred2(jump_sign = TRUE,  use_mle = TRUE,  coef_shrink = 0.95),
       file = paste0("results_simu/LDpred2_", params, ".rds")),
     SBayesS = runonce::save_run(
       run_sbayess(),
       file = paste0("results_simu/SBayesS_", params, ".rds")),
-    LDSc = list(all_h2 = ldsc)
+    LDSC = list(all_h2 = ldsc)
   )
 })
 
@@ -148,22 +148,19 @@ grid %>%
     r2 = list(unname(r2)),
     n_keep_infer = `if`(grepl("LDpred2", Method), `if`(is.null(all_h2), 0, NCOL(all_h2)), NA),
     n_keep_pred  = `if`(grepl("LDpred2", Method), `if`(is.null(all_r2), 0, NCOL(all_r2) / 500 * 20), NA),
-    r2_est = list(`if`(Method == "LDSc", NULL,
+    r2_est = list(`if`(Method == "LDSC", NULL,
                        unname(quantile(all_r2, probs = c(0.5, 0.025, 0.975))))),
[…]
